% !TEX root = ../main.tex
\chapter{Linear, Convolution, Pooling, Normalization}
\label{ch:linear_conv_norm}

This chapter treats the four most fundamental layer families in
deep learning: the linear (fully-connected) layer, the
convolutional layers, the pooling layers, and the normalisation
layers. They are the building blocks from which essentially every
modern architecture is composed: a CNN is convolution + pooling +
norm + activation; a Transformer is linear + attention + norm; a
ResNet is convolution + norm + linear + skip connection. \Lucid's
implementations follow \refframework's API closely.

We cover, for each family: the mathematical operation, the
weight initialisation convention, the forward/backward dispatch
in \Lucid, and the common variants exposed at \code{lucid.nn}.

\section{Linear (Fully-Connected) Layer}
\label{sec:linear_conv_norm:linear}

The linear layer applies an affine transformation:
\[
y \;=\; x W^\top + b,
\]
where $x \in \mathbb{R}^{B \times d_{\text{in}}}$, $W \in
\mathbb{R}^{d_{\text{out}} \times d_{\text{in}}}$, $b \in
\mathbb{R}^{d_{\text{out}}}$, $y \in \mathbb{R}^{B \times
d_{\text{out}}}$.

\subsection{The Lucid API}
\label{ssec:linear_conv_norm:linear_api}

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid.nn as nn

fc = nn.Linear(in_features=768, out_features=3072, bias=True)
y = fc(x)   # x shape: (..., 768); y shape: (..., 3072)
\end{lstlisting}

The leading dimensions of \code{x} are preserved; the trailing
dimension is transformed.

\subsection{Initialisation}
\label{ssec:linear_conv_norm:linear_init}

The default initialisation is Kaiming uniform with $a = \sqrt{5}$:
\[
W_{ij} \;\sim\; \mathcal{U}\!\left(-\sqrt{\frac{6}{d_{\text{in}}}},\, \sqrt{\frac{6}{d_{\text{in}}}}\right).
\]
The bias (if present) is initialised from
$\mathcal{U}(-1/\sqrt{d_{\text{in}}},\, 1/\sqrt{d_{\text{in}}})$.

These defaults match \refframework's. \chref{ch:initialization}
discusses initialisation theory in depth.

\subsection{Implementation}
\label{ssec:linear_conv_norm:linear_impl}

The forward is a matrix multiplication (plus optional bias),
dispatched through the backend's \code{matmul} (\code{cblas\_sgemm}
on CPU via the AMX coprocessor; \code{mlx::core::matmul} on GPU).
The backward propagates:
\[
\bar{W} \;=\; \bar{y}^\top x,
\qquad
\bar{x} \;=\; \bar{y}\, W,
\qquad
\bar{b} \;=\; \sum_{\text{batch}} \bar{y}.
\]

The bias gradient sums across the batch axes. For a 4-D input
(batch + sequence + ... + features), the sum is over all but the
last axis.

\section{Convolution}
\label{sec:linear_conv_norm:conv}

The convolutional layer applies a sliding cross-correlation
(usually called ``convolution'' in ML even though strict
convolution would flip the kernel):
\[
y_{c, h, w} \;=\; b_c \;+\; \sum_{c'=0}^{C_{\text{in}}-1}
\sum_{i=0}^{K_h-1} \sum_{j=0}^{K_w-1}
W_{c, c', i, j} \cdot x_{c', h \cdot s + i - p,\, w \cdot s + j - p}
\]
for a 2-D convolution with stride $s$ and padding $p$. The
mathematical derivation of the output shape is in
\cite{dumoulin_guide_2016}.

\subsection{The API}
\label{ssec:linear_conv_norm:conv_api}

\begin{lstlisting}[style=lucid,language=LucidPython]
conv = nn.Conv2d(
    in_channels=3,
    out_channels=64,
    kernel_size=3,
    stride=1,
    padding=1,
    bias=True,
)
y = conv(x)  # x shape: (B, 3, H, W); y shape: (B, 64, H, W)
\end{lstlisting}

The output spatial size for a 2-D convolution is
\[
H_{\text{out}} \;=\; \left\lfloor \frac{H_{\text{in}} + 2p - K_h}{s} \right\rfloor + 1.
\]
With padding $p = (K-1)/2$ and stride 1, the spatial size is
preserved (for odd $K$).

\subsection{Variants}
\label{ssec:linear_conv_norm:conv_variants}

\code{lucid.nn} provides:
\begin{itemize}
  \item \code{Conv1d}, \code{Conv2d}, \code{Conv3d}: standard
    convolution at each rank.
  \item \code{ConvTranspose1d}, \code{ConvTranspose2d},
    \code{ConvTranspose3d}: transposed (deconvolution) for
    upsampling.
  \item \code{Conv2d(groups=g)}: grouped convolution (input
    channels split into $g$ groups, each convolved independently).
    With $g = C_{\text{in}} = C_{\text{out}}$, this becomes depthwise
    convolution (MobileNet \cite{howard_mobilenet_2017}).
  \item \code{Conv2d(dilation=d)}: dilated convolution (each kernel
    cell is spaced by $d$ in the input).
\end{itemize}

\subsection{Output-shape arithmetic}
\label{ssec:linear_conv_norm:conv_arithmetic}

The general 2-D convolution output shape with stride $s$,
padding $p$, dilation $d$ and kernel size $K$:
\[
  H_{\text{out}} \;=\;
  \left\lfloor
  \frac{H_{\text{in}} + 2p - d(K - 1) - 1}{s}
  \right\rfloor + 1.
\]
Special cases:

\begin{itemize}
  \item \textbf{Same padding} ($p = (K-1)/2$, $s = 1$, $d = 1$):
    $H_{\text{out}} = H_{\text{in}}$.
  \item \textbf{Stride-2 downsample} ($s = 2$): halves the
    spatial size (with appropriate padding).
  \item \textbf{Dilated $K = 3$ with $d = 2$}: effective
    receptive field is $5 \times 5$ but parameter count is
    that of $3 \times 3$.
\end{itemize}

For transposed convolution (output upsampling):
\[
  H_{\text{out}} \;=\; (H_{\text{in}} - 1) s - 2 p + d(K-1) + p_{\text{out}} + 1,
\]
where $p_{\text{out}}$ is the optional output-padding parameter
that disambiguates between multiple valid output sizes for a
given stride.

\subsection{Depthwise and pointwise convolution}
\label{ssec:linear_conv_norm:depthwise_pointwise}

Depthwise convolution: one filter per input channel, no
cross-channel mixing. Parameter count: $C_{\text{in}} K^2$ (vs
$C_{\text{in}} C_{\text{out}} K^2$ for standard conv).

Pointwise convolution: $1 \times 1$ kernel, mixes channels.
Parameter count: $C_{\text{in}} C_{\text{out}}$.

The MobileNet \cite{howard_mobilenet_2017} \emph{depthwise-
separable convolution} factors a standard conv into depthwise
then pointwise:
\[
  \text{cost ratio} \;=\;
  \frac{C_{\text{in}} K^2 + C_{\text{in}} C_{\text{out}}}{C_{\text{in}} C_{\text{out}} K^2}
  \;=\; \frac{1}{C_{\text{out}}} + \frac{1}{K^2}.
\]
For $K = 3$ and $C_{\text{out}} \geq 64$, this is roughly
$1/9 + 1/64 \approx 0.13$, an 8$\times$ reduction in compute
and parameters for almost-equivalent expressive power.

\subsection{Grouped convolution}
\label{ssec:linear_conv_norm:grouped_conv}

\code{Conv2d(groups=g)} splits the channel axis into $g$ groups
and convolves within each:
\[
  y_{c, h, w} \;=\;
  \sum_{c' = c_{\text{group}} \cdot C_{\text{in}}/g}^{(c_{\text{group}}+1) \cdot C_{\text{in}}/g - 1}
  \sum_{i, j} W_{c, c' \bmod C_{\text{in}}/g, i, j} \cdot x_{c', \ldots},
\]
where $c_{\text{group}} = \lfloor c / (C_{\text{out}}/g) \rfloor$.
With $g > 1$, the parameter count drops by a factor of $g$;
ResNeXt is the canonical use.

\subsection{Dilated (atrous) convolution}
\label{ssec:linear_conv_norm:dilated_conv}

Dilation inserts zeros between kernel cells:
\[
  y(i) \;=\; \sum_k x(i + d \cdot k) \cdot w(k).
\]
The effective receptive field grows as $K_{\text{eff}} = d(K-1) + 1$
while keeping the parameter count at $K^2$. Used heavily in
DeepLab \cite{chen_deeplab_2017} for semantic segmentation
(large receptive field without spatial downsampling).

\subsection{Initialisation}
\label{ssec:linear_conv_norm:conv_init}

Default is Kaiming uniform, with fan-in computed from
$C_{\text{in}} \cdot K_h \cdot K_w$ (the receptive-field size
times the input channels).

\subsection{Dispatch}
\label{ssec:linear_conv_norm:conv_dispatch}

The forward dispatches through the backend's \code{conv2d}: BNNS
on CPU (using AMX coprocessor for the GEMM-equivalent step), MLX
on GPU. For small inputs, BNNS is competitive with MLX; for
large batches and high channel counts, MLX wins.

The backward decomposes into: gradient with respect to the input
(another convolution with a flipped kernel), gradient with
respect to the weight (a convolution with the input as kernel),
gradient with respect to the bias (a sum over batch/spatial).

\section{Pooling}
\label{sec:linear_conv_norm:pooling}

Pooling layers reduce spatial resolution by aggregating each local
patch into a single value.

\subsection{Max pooling}
\label{ssec:linear_conv_norm:max_pool}

\[
y_{c, h, w} \;=\; \max_{i, j \in \text{window}}
x_{c, h \cdot s + i,\, w \cdot s + j}.
\]
Backward: gradient routes to the maximum-valued input position.
Where multiple positions tie, the convention is to route to the
first (lowest index).

\subsection{Average pooling}
\label{ssec:linear_conv_norm:avg_pool}

\[
y_{c, h, w} \;=\; \frac{1}{K_h K_w} \sum_{i, j} x_{c, h s + i, w s + j}.
\]
Backward: gradient broadcasts uniformly across the input window.

\subsection{Adaptive pooling}
\label{ssec:linear_conv_norm:adaptive_pool}

\code{AdaptiveMaxPool2d(output\_size)} and
\code{AdaptiveAvgPool2d(output\_size)} pool to a fixed output
spatial shape, computing the appropriate window and stride
automatically. Used at the end of CNN backbones (ResNet, ConvNeXt)
to produce a fixed-size feature regardless of input size.

\subsection{Global pooling}
\label{ssec:linear_conv_norm:global_pool}

\code{AdaptiveAvgPool2d(1)} is the global average pooling
operation (one value per channel). It is the standard transition
from a $(B, C, H, W)$ feature map to a $(B, C)$ classifier input.

\section{Normalisation}
\label{sec:linear_conv_norm:norm}

Normalisation layers standardise their input's distribution to
stabilise training. The standard family has four members:
batch norm, layer norm, instance norm, group norm.

\subsection{BatchNorm}
\label{ssec:linear_conv_norm:batchnorm}

\code{BatchNorm2d(num\_features)} \cite{ioffe_batchnorm_2015}
standardises each channel across the batch and spatial dimensions:
\[
\hat x_{b, c, h, w} \;=\; \frac{x_{b, c, h, w} - \mu_c}{\sqrt{\sigma_c^2 + \epsilon}},
\quad y \;=\; \gamma_c \hat x + \beta_c,
\]
where $\mu_c$ and $\sigma_c^2$ are computed over $(b, h, w)$ in
training and read from running statistics in evaluation. $\gamma$
and $\beta$ are learnable per-channel affine parameters.

The running statistics are buffers; they update via exponential
moving average in training mode:
\[
\mu_{\text{run}} \;\leftarrow\; (1 - \alpha)\,\mu_{\text{run}} + \alpha\,\mu_{\text{batch}},
\]
with momentum $\alpha = 0.1$ by default.

\subsubsection*{Backward derivation}

Let $N$ be the number of elements pooled into each channel's
statistics ($B \cdot H \cdot W$ for BatchNorm2d). The
backward through the standardisation is non-trivial because
$\mu$ and $\sigma^2$ depend on every input element:

\begin{align}
  \frac{\partial \mathcal{L}}{\partial x_i}
  &=\;
  \frac{\partial \mathcal{L}}{\partial \hat x_i} \cdot
  \frac{1}{\sqrt{\sigma^2 + \epsilon}} \notag \\
  &+\; \frac{\partial \mathcal{L}}{\partial \sigma^2}
  \cdot \frac{2(x_i - \mu)}{N}
  + \frac{\partial \mathcal{L}}{\partial \mu} \cdot \frac{1}{N},
\end{align}
where
\[
  \frac{\partial \mathcal{L}}{\partial \sigma^2}
  = -\tfrac{1}{2} (\sigma^2 + \epsilon)^{-3/2}
  \sum_i \frac{\partial \mathcal{L}}{\partial \hat x_i}(x_i - \mu),
\]
\[
  \frac{\partial \mathcal{L}}{\partial \mu}
  = -\frac{1}{\sqrt{\sigma^2 + \epsilon}}
  \sum_i \frac{\partial \mathcal{L}}{\partial \hat x_i}
  - \frac{2}{N} \frac{\partial \mathcal{L}}{\partial \sigma^2}
  \sum_i (x_i - \mu).
\]
The same fused-kernel formula applies to LayerNorm,
InstanceNorm, GroupNorm with different choices of $N$ and the
axes being summed over.

\subsubsection*{Training-vs-eval semantics}

In training mode, the batch statistics are used for
normalisation; running statistics are updated with the
exponential moving average. In eval mode (\code{model.eval()}),
the running statistics are used directly, and no update
occurs. This decoupling matters at inference time: the
running statistics absorb the cumulative training-data
distribution, giving a deterministic forward pass.

A common pitfall: forgetting \code{model.eval()} during
validation. The validation batches' statistics replace the
running ones, perturbing the eval forward and degrading
reported accuracy. \code{model.eval()} must be called before
any validation forward pass.

\subsubsection*{Batch-size dependence}

BatchNorm's batch-statistics estimate has variance $O(1/N)$
where $N$ is the effective batch size. With small batches
(say $N = 2$, common in detection workloads), the noise
dominates and training becomes unstable. The fixes:

\begin{itemize}
  \item GroupNorm ($N$ is the per-sample group size; no batch
    dependence).
  \item SyncBatchNorm (gather statistics across distributed
    workers; effective $N$ multiplies by world size).
  \item Frozen-BN (freeze running stats; only $\gamma, \beta$
    trainable).
\end{itemize}

\subsection{LayerNorm}
\label{ssec:linear_conv_norm:layernorm}

\code{LayerNorm(normalized\_shape)} \cite{ba_layernorm_2016}
standardises each sample independently across the named axes:
\[
\mu \;=\; \frac{1}{N} \sum_i x_i, \qquad
\sigma^2 \;=\; \frac{1}{N} \sum_i (x_i - \mu)^2,
\qquad
\hat x \;=\; \frac{x - \mu}{\sqrt{\sigma^2 + \epsilon}}.
\]

LayerNorm is the standard for Transformers: no running statistics,
no batch-size dependence, same behaviour in training and
evaluation. \Lucid's implementation dispatches through the
\MPSGraph\ per-op carve-out
(\chref{ch:mpsgraph_dispatch}) for substantial speedup.

\subsection{InstanceNorm}
\label{ssec:linear_conv_norm:instancenorm}

\code{InstanceNorm2d(num\_features)} \cite{ulyanov_instancenorm_2016}
standardises each channel of each sample independently across the
spatial dimensions. Used in style-transfer networks and some GAN
architectures.

\subsection{GroupNorm}
\label{ssec:linear_conv_norm:groupnorm}

\code{GroupNorm(num\_groups, num\_channels)} \cite{wu_groupnorm_2018}
splits the channel dimension into groups and normalises within each
group:
\[
\hat x_{b, g, c', h, w} \;=\; \frac{x_{b, g, c', h, w} - \mu_{b, g}}{\sqrt{\sigma_{b, g}^2 + \epsilon}}.
\]
Bridges between BatchNorm ($G = 1$, all channels in one group) and
InstanceNorm ($G = C$, each channel its own group). Robust to
small batch sizes.

\subsection{RMSNorm}
\label{ssec:linear_conv_norm:rmsnorm}

A simplification of LayerNorm \cite{zhang_rmsnorm_2019} that
omits the mean-centring:
\[
\hat x \;=\; \frac{x}{\sqrt{\frac{1}{N}\sum_i x_i^2 + \epsilon}}.
\]
Used in modern LLM architectures (LLaMA-family) for slight speed
improvement and competitive accuracy.

\subsection{Comparison: which to choose}
\label{ssec:linear_conv_norm:norm_compare}

\begin{table}[t]
\centering
\small
\begin{tabular}{llll}
\toprule
Layer & Stats over & Buffers & Typical use \\
\midrule
BatchNorm  & batch + spatial & $\mu, \sigma^2$ EMA & CNN classification \\
LayerNorm  & feature dim     & none                & Transformer \\
InstanceNorm & spatial       & optional EMA        & Style transfer, GAN \\
GroupNorm  & per-group       & none                & Detection (small batch) \\
RMSNorm    & feature dim, no centring & none      & LLaMA-style LLMs \\
\bottomrule
\end{tabular}
\caption[Normalisation layer comparison.]{The five
normalisation layers differ in which axes they pool over and
whether they maintain running statistics. The choice is driven
by batch size, task, and tradition; the math is otherwise
nearly identical.}
\label{tab:linear_conv_norm:norm_compare}
\end{table}

\subsection{Numerical stability}
\label{ssec:linear_conv_norm:norm_stability}

All normalisation layers share two numerical pitfalls:

\begin{itemize}
  \item \textbf{The $\epsilon$ in the denominator}: prevents
    division by zero when the variance is exactly zero
    (uniform input). Default is $10^{-5}$; some training
    recipes use $10^{-6}$ for tighter normalisation.
  \item \textbf{Variance computation}: the naive
    $\sigma^2 = \mathbb{E}[x^2] - \mathbb{E}[x]^2$ formula
    catastrophically cancels for large means. \Lucid\ uses
    the two-pass algorithm (compute $\mu$ then
    $\sigma^2 = \mathbb{E}[(x - \mu)^2]$) for stability.
\end{itemize}

Under AMP (\chref{ch:amp}), normalisation layers are blacklisted
(forced to FP32) precisely because of these numerics; the
cost is negligible because the layer is memory-bandwidth-bound
rather than compute-bound.

\section{Dropout}
\label{sec:linear_conv_norm:dropout}

\code{Dropout(p=0.5)} zeros out a fraction $p$ of activations in
training mode (and scales the rest by $1/(1-p)$ to preserve
expectations). In evaluation mode, it's the identity.

\subsection{Variants}
\label{ssec:linear_conv_norm:dropout_variants}

\begin{itemize}
  \item \code{Dropout}: per-element.
  \item \code{Dropout2d}: per-channel for 2-D feature maps (a
    whole feature map is either kept or dropped).
  \item \code{Dropout3d}: per-channel for 3-D.
  \item \code{AlphaDropout}: SELU-compatible variant that
    preserves the input mean and variance.
\end{itemize}

\subsection{Inverted scaling}
\label{ssec:linear_conv_norm:inverted_scaling}

The \emph{inverted} dropout convention scales kept values by
$1/(1-p)$ at training time (rather than scaling outputs by
$(1-p)$ at eval time). This means the eval-mode forward is the
identity, which matches \refframework's convention.

\section{Embedding}
\label{sec:linear_conv_norm:embedding}

\code{Embedding(num\_embeddings, embedding\_dim)} is a lookup
table: maps integer indices to dense vectors. Used as the first
layer in NLP and recommendation models.

The forward is a gather operation on the weight matrix:
\[
y_{b, t, :} \;=\; W_{\text{idx}_{b, t}, :}.
\]

\subsection{Backward through embedding}
\label{ssec:linear_conv_norm:embedding_backward}

The backward is a scatter-add (\chref{ch:indexing}). For the
common case where the same vocabulary token appears many times in
a batch, multiple gradient contributions accumulate at the same
row of \code{grad\_W}.

The 3.4 per-op MPSGraph carve-out
(\chref{ch:mpsgraph_dispatch}) makes embedding backward an order
of magnitude faster than the naive scatter-add path.

\section{Padding}
\label{sec:linear_conv_norm:padding}

\code{lucid.nn} provides explicit padding modules:
\begin{itemize}
  \item \code{ZeroPad2d}, \code{ConstantPad2d}: pad with constant.
  \item \code{ReflectionPad2d}: pad with mirror reflection.
  \item \code{ReplicationPad2d}: pad with edge replication.
  \item \code{CircularPad2d}: pad wraparound.
\end{itemize}

Most users do not need these because the convolution layers
accept a \code{padding} argument that handles zero-padding
inline. The explicit modules are useful when the padding mode
must be non-zero.

\section{Flatten and Reshape Helpers}
\label{sec:linear_conv_norm:flatten}

\code{Flatten} and \code{Unflatten} are module-equivalents of
\code{reshape}, useful inside \code{Sequential}:

\begin{lstlisting}[style=lucid,language=LucidPython]
classifier = nn.Sequential(
    nn.Conv2d(3, 64, 3, padding=1),
    nn.BatchNorm2d(64),
    nn.ReLU(),
    nn.AdaptiveAvgPool2d(1),
    nn.Flatten(),
    nn.Linear(64, 10),
)
\end{lstlisting}

\section{Functional vs Module Forms}
\label{sec:linear_conv_norm:functional_vs_module}

Every standard layer has both a module form (stateful) and a
functional form (stateless). The functional form lives in
\code{lucid.nn.functional}:

\begin{itemize}
  \item \code{nn.Linear(...)} (module) vs \code{F.linear(x, w, b)}
    (functional).
  \item \code{nn.Conv2d(...)} (module) vs \code{F.conv2d(x, w, b, ...)}.
  \item \code{nn.ReLU()} (module) vs \code{F.relu(x)}.
  \item \code{nn.BatchNorm2d(...)} (module) vs
    \code{F.batch\_norm(x, running\_mean, running\_var, w, b, ...)}.
\end{itemize}

The functional form takes weights and statistics as arguments;
the module form owns them. Both compute the same result; the
choice is whether the user wants to manage the state explicitly
(functional) or have a stateful container (module).

\section{Implementation Mapping}
\label{sec:linear_conv_norm:impl_mapping}

\tabref{tab:linear_conv_norm:impl} maps the public modules to
their backend dispatch paths.

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Module & CPU path & GPU path \\
\midrule
\code{Linear}        & \code{cblas\_sgemm} (AMX) & \code{mlx::matmul} \\
\code{Conv2d}        & BNNS direct conv         & \code{mlx::conv2d} \\
\code{BatchNorm2d}   & AccelerateBatchNorm.cpp & MLX primitives \\
\code{LayerNorm}     & AccelerateLayerNorm.cpp & \emph{MPSGraph carve-out} \\
\code{InstanceNorm}  & AccelerateInstanceNorm.cpp & MLX primitives \\
\code{GroupNorm}     & AccelerateGroupNorm.cpp & MLX primitives \\
\code{MaxPool2d}     & vDSP loop                & MLX primitive \\
\code{AvgPool2d}     & vDSP reduction           & MLX primitive \\
\code{Dropout}       & Bernoulli mask + scale   & Bernoulli mask + scale \\
\code{Embedding}     & gather                   & gather \\
\bottomrule
\end{tabular}
\caption[Standard layer backends.]{Backend dispatch path for the
standard layer family. The \emph{MPSGraph carve-out} entry marks
the per-op fast path
(\chref{ch:mpsgraph_dispatch}) that gives an order-of-magnitude
speedup over the eager MLX path for LayerNorm.}
\label{tab:linear_conv_norm:impl}
\end{table}

\section{Summary and Forward References}
\label{sec:linear_conv_norm:summary}

The standard layer library covers the four fundamental neural-
network primitives: linear (matrix multiplication + bias),
convolution (sliding-kernel cross-correlation), pooling (spatial
aggregation), and normalisation (per-axis standardisation).
\Lucid's implementations follow \refframework's API and dispatch
through the backend's optimised paths (BNNS / vDSP / MLX /
MPSGraph carve-outs where applicable).

The next chapter, \chref{ch:rnn_family}, treats recurrent
networks. \chref{ch:attention} treats attention.
\chref{ch:losses} treats loss functions.

\begin{lucidnote}
For the user: the standard layer modules are drop-in compatible
with \refframework. Migration is import-line substitution
(\code{import torch.nn as nn} $\to$ \code{import lucid.nn as nn}).
The state-dict format is also compatible, so checkpoints can be
loaded across frameworks with minor key adjustments.
\end{lucidnote}
