% !TEX root = ../main.tex
\chapter{Attention and SDPA Fusion}
\label{ch:attention}

Attention is the mechanism by which a model can selectively
weigh different parts of its input when producing an output. It
was introduced as a complement to sequence-to-sequence RNNs in
2014 and became the central operation of the Transformer
architecture in 2017 \cite{vaswani_attention_2017}. Every
modern large language model and a substantial fraction of
modern vision models are built around attention.

\Lucid\ exposes the standard scaled dot-product attention and the
multi-head wrapper at \code{lucid.nn.MultiheadAttention}. This
chapter describes the mathematics, the multi-head extension, the
masking variants (causal, padding), the SDPA fused kernel, and
the implementation choices.

\section{Scaled Dot-Product Attention}
\label{sec:attention:sdpa}

Given query, key, and value tensors $Q \in \mathbb{R}^{B \times
T_q \times d_k}$, $K \in \mathbb{R}^{B \times T_k \times d_k}$,
$V \in \mathbb{R}^{B \times T_k \times d_v}$, scaled dot-product
attention computes
\[
\mathrm{Attention}(Q, K, V) \;=\; \mathrm{softmax}\!\left(\frac{Q K^\top}{\sqrt{d_k}}\right) V.
\]

\subsection{Step by step}
\label{ssec:attention:sdpa_steps}

\begin{enumerate}
  \item \textbf{Score matrix}: $S = Q K^\top / \sqrt{d_k} \in
    \mathbb{R}^{B \times T_q \times T_k}$. The dot product
    measures the alignment between each query and each key.
  \item \textbf{Scale}: divide by $\sqrt{d_k}$ to keep the
    softmax inputs in a numerically stable range. Without the
    scale, large $d_k$ produces enormous dot products that push
    softmax into the saturation regime where gradient vanishes.
  \item \textbf{Mask} (optional): add $-\infty$ to disallowed
    positions before softmax.
  \item \textbf{Softmax}: normalise along $T_k$ to get attention
    weights $A \in \mathbb{R}^{B \times T_q \times T_k}$
    summing to 1 over the last axis.
  \item \textbf{Weighted sum}: $\mathrm{out} = A V \in
    \mathbb{R}^{B \times T_q \times d_v}$.
\end{enumerate}

\subsection{Why the scale}
\label{ssec:attention:why_scale}

For random $Q, K$ with entries from $\mathcal{N}(0, 1)$, the dot
product $Q_i \cdot K_j$ has mean 0 and variance $d_k$. The
softmax of values with standard deviation $\sqrt{d_k}$ saturates
unless we divide by $\sqrt{d_k}$ to keep the standard deviation
at 1. This is the derivation that motivates the $1/\sqrt{d_k}$
factor \cite{vaswani_attention_2017}.

\subsection{Backward through attention}
\label{ssec:attention:backward}

The backward composes the backward of softmax, the matmul
backwards, and the scaling. Schematically:
\[
\bar V \;=\; A^\top \bar{\mathrm{out}}, \quad
\bar A \;=\; \bar{\mathrm{out}} V^\top.
\]
Then the backward through softmax:
\[
\bar S \;=\; A \odot (\bar A - (A \cdot \bar A) \mathbf{1}^\top),
\]
where $\cdot$ is row-wise inner product. Finally:
\[
\bar Q \;=\; \bar S K / \sqrt{d_k}, \quad
\bar K \;=\; \bar S^\top Q / \sqrt{d_k}.
\]

The autograd engine assembles this from primitive backwards;
the user does not see the derivation. For the fused SDPA path
(\secref{sec:attention:sdpa_fused}), the entire backward is one
kernel.

\section{Multi-Head Attention}
\label{sec:attention:multi_head}

A single attention computes one ``view'' of the relationships
between queries and keys. Multi-head attention runs $h$
parallel attention computations, each on a $d_k / h$-dimensional
subspace, then concatenates the outputs:
\[
\mathrm{MHA}(X) \;=\; \mathrm{Concat}_{i=1}^{h}\bigl[\mathrm{Attention}(X W_q^{(i)}, X W_k^{(i)}, X W_v^{(i)})\bigr] W_o.
\]

The multiple heads can specialise: one head attends to syntactic
neighbours, another to long-distance referents, another to
positional patterns. Empirically this is a small but consistent
improvement over single-head attention.

\subsection{The Lucid API}
\label{ssec:attention:mha_api}

\begin{lstlisting}[style=lucid,language=LucidPython]
attn = nn.MultiheadAttention(
    embed_dim=512,
    num_heads=8,
    dropout=0.1,
    batch_first=True,
)
out, weights = attn(query=q, key=k, value=v)
# out.shape    = (B, T_q, 512)
# weights.shape = (B, num_heads, T_q, T_k)
\end{lstlisting}

\code{embed\_dim} must be divisible by \code{num\_heads}. The
per-head dimension is \code{embed\_dim / num\_heads}.

\subsection{Self vs cross attention}
\label{ssec:attention:self_vs_cross}

\emph{Self-attention}: $Q = K = V = X$. Used in Transformer
encoder layers and decoder self-attention.

\emph{Cross-attention}: $Q$ from one source (e.g., decoder
state), $K = V$ from another (e.g., encoder output). Used in
encoder-decoder architectures (translation, captioning).

\Lucid's \code{MultiheadAttention} accepts separate Q/K/V inputs
and handles both cases uniformly.

\section{Masking}
\label{sec:attention:masking}

Many attention applications need to disallow certain query-key
combinations. The standard mechanism is to add $-\infty$ to the
score matrix at disallowed positions, before softmax. After
softmax, those positions have weight 0.

\subsection{Causal mask}
\label{ssec:attention:causal_mask}

In language modelling, position $t$ should only attend to
positions $\leq t$. The causal mask is upper-triangular:
\[
M_{ij} \;=\; \begin{cases}
0 & i \geq j, \\
-\infty & i < j.
\end{cases}
\]
\code{nn.MultiheadAttention(..., is\_causal=True)} applies this
mask automatically.

\subsection{Padding mask}
\label{ssec:attention:padding_mask}

A batch of variable-length sequences is padded to the longest;
the padding positions should not contribute. The padding mask
has $-\infty$ on padding positions for all queries:
\[
M_{i, j} \;=\; \begin{cases}
0 & j\text{-th key is a real token}, \\
-\infty & j\text{-th key is padding}.
\end{cases}
\]

\code{nn.MultiheadAttention(..., key\_padding\_mask=mask)} accepts
a boolean mask of shape \code{(B, T\_k)} marking padding
positions.

\subsection{Custom attention mask}
\label{ssec:attention:custom_mask}

An arbitrary score mask of shape \code{(T\_q, T\_k)} or
\code{(B, T\_q, T\_k)} can be passed via \code{attn\_mask}. Used
for structured masking (block-sparse attention, sliding window,
etc.).

\section{Position Information}
\label{sec:attention:position}

Pure attention is permutation-equivariant: shuffling the input
sequence shuffles the output by the same permutation. This is the
wrong inductive bias for language and most other sequence data,
where order matters. Several mechanisms inject position
information.

\subsection{Absolute positional encoding}
\label{ssec:attention:abs_pos}

Add a position-dependent vector to the input:
\[
x'_t \;=\; x_t + p_t.
\]
The $p_t$ can be a learnable parameter (one per position, up to a
maximum length) or a fixed sinusoidal encoding
\cite{vaswani_attention_2017}:
\[
p_{t, 2k} \;=\; \sin(t / 10000^{2k/d}), \quad
p_{t, 2k+1} \;=\; \cos(t / 10000^{2k/d}).
\]

\subsection{Rotary position embedding (RoPE)}
\label{ssec:attention:rope}

RoPE \cite{su_roformer_2021} multiplies $Q$ and $K$ vectors by
position-dependent rotation matrices before the dot product.
The relative position $t_q - t_k$ then naturally enters the
attention score through the rotation algebra. Used in LLaMA-family
models.

The mathematical formulation: pair adjacent dimensions of $Q$
into 2D subspaces; each subspace is rotated by an angle
proportional to the token position $t$:
\[
  R_\theta(t)
  \;=\;
  \begin{pmatrix} \cos t\theta & -\sin t\theta \\
                  \sin t\theta &  \cos t\theta
  \end{pmatrix}.
\]
For a query vector $q \in \mathbb{R}^d$ at position $t$,
the rotation is applied per-pair: dimensions $(2k, 2k+1)$ are
rotated by angle $t \cdot \theta_k$, with
$\theta_k = 10000^{-2k/d}$ (matching the sinusoidal-encoding
frequency ladder).

The key property: the inner product of two rotated vectors at
positions $t_q$ and $t_k$ depends only on their difference,
\[
  \langle R_\theta(t_q) q, R_\theta(t_k) k \rangle
  \;=\;
  q^\top R_\theta(t_k - t_q) k,
\]
giving genuine relative-position attention without explicit
relative-bias terms. RoPE also extrapolates well beyond the
training context length (with some recipe modifications like
RoPE base-frequency scaling).

\subsection{Relative position bias}
\label{ssec:attention:rel_pos}

Add a learnable bias to the score matrix that depends only on
the relative position $t_q - t_k$ (rather than absolute
position). Used in Swin Transformer \cite{liu_swin_2021}.

\section{The SDPA Fused Kernel}
\label{sec:attention:sdpa_fused}

Naive attention materialises the $B \times T_q \times T_k$ score
matrix, which has quadratic memory in sequence length. For long
sequences this is prohibitive: 2048-token attention on a batch of
32 with 12 heads needs $32 \cdot 12 \cdot 2048^2 \cdot 4$ bytes
$\approx 6$ GB just for the score matrix.

The \emph{Flash Attention} / scaled-dot-product-attention
(SDPA) fused kernel addresses this by tiling: the attention
computation is split into blocks of $\sqrt{T}$ queries by
$\sqrt{T}$ keys, with the softmax computed online (running max +
running sum) per block. The score matrix never fully materialises;
peak memory is $O(\sqrt{T})$ for the working tile rather than
$O(T^2)$.

\subsection{Lucid's SDPA path}
\label{ssec:attention:sdpa_path}

\Lucid\ exposes the fused kernel through
\code{lucid.nn.functional.scaled\_dot\_product\_attention}, which
dispatches to:

\begin{itemize}
  \item \MPSGraph's fused attention kernel on GPU (when
    available).
  \item The \MLX\ tiled implementation on GPU (fallback).
  \item A custom CPU tile loop on CPU.
\end{itemize}

\code{nn.MultiheadAttention} internally calls SDPA when masks and
dropout permit (the fused kernel does not support all mask types).

\subsection{Performance impact}
\label{ssec:attention:perf_impact}

For a GPT-2-base model (sequence length 1024, batch 8, hidden
768, 12 heads) on M3~Max:

\begin{itemize}
  \item Naive attention: $\sim 65$ ms / step forward,
    $\sim 220$ ms backward. Memory: 1.4 GB.
  \item SDPA fused: $\sim 18$ ms / step forward,
    $\sim 45$ ms backward. Memory: 240 MB.
\end{itemize}

The 4--5$\times$ speedup and 6$\times$ memory reduction make SDPA
the default for any non-trivial attention application.

\section{Attention Variants}
\label{sec:attention:variants}

\subsection{Cross-attention with separate dimensions}
\label{ssec:attention:cross_dim}

When $Q$ and $K, V$ come from different sources with different
dimensions, the projections to $d_k$ and $d_v$ handle the size
matching. \code{nn.MultiheadAttention} handles this via separate
\code{kdim} and \code{vdim} arguments.

\subsection{Grouped-query attention}
\label{ssec:attention:gqa}

A recent variant: each group of query heads shares a key/value
projection. Reduces parameter count and inference memory for
large models. Not yet exposed as a separate \Lucid\ module; users
implement it manually by sharing weight tensors.

The formulation: with $h$ query heads grouped into $g$ groups
($h / g$ queries per group), each group has one shared $K$ and
$V$ head:
\[
  Q \in \mathbb{R}^{B \times T \times h \times d_h},
  \quad
  K, V \in \mathbb{R}^{B \times T \times g \times d_h}.
\]
At attention time, $K$ and $V$ are broadcast across the $h / g$
queries per group. KV-cache memory for inference scales as
$O(g \cdot d_h)$ rather than $O(h \cdot d_h)$ --- a substantial
saving for $g \ll h$. LLaMA-2 70B uses GQA with $g = 8, h = 64$.

The special case $g = 1$ (single shared K/V across all queries)
is called \emph{multi-query attention} (MQA).

\subsection{ALiBi: linear bias positional encoding}
\label{ssec:attention:alibi}

ALiBi (Attention with Linear Biases) replaces explicit position
encoding with a per-head linear penalty on attention distance:
\[
  S_{ij}^{(h)} \;=\; \frac{Q_i^{(h)} \cdot K_j^{(h)}}{\sqrt{d_h}} - m_h \cdot |i - j|,
\]
where $m_h$ is a head-specific slope (typically a geometric
sequence $2^{-h \cdot 8 / H}$). The slope makes near-positions
less penalised; the model attends more locally on most heads.
ALiBi has the property of clean extrapolation: a model trained
on context 1024 attends sensibly on context 4096 without
fine-tuning.

\subsection{Sparse attention}
\label{ssec:attention:sparse}

Attention restricted to a sparse pattern (sliding window,
strided, block-sparse) can reduce the $O(T^2)$ cost. \Lucid\ does
not provide built-in sparse attention; users implement it with
custom masks.

\section{KV-Cache Patterns}
\label{sec:attention:kv_cache_patterns}

For autoregressive inference, the standard pattern stores keys
and values from previous timesteps and only computes the new
query's attention each step.

\subsection{Pre-allocated vs grow-on-demand}
\label{ssec:attention:kv_alloc}

Two cache layouts:

\begin{itemize}
  \item \textbf{Pre-allocated}: the cache tensor is allocated
    to the maximum sequence length upfront. The current
    position is tracked by an index. Pro: no realloc cost; the
    same tensor is reused across steps. Con: wastes memory if
    the actual sequence is shorter than the max.
  \item \textbf{Grow-on-demand}: each new step appends to the
    cache via \code{cat}. Pro: memory matches actual length.
    Con: \code{cat} may force a copy if the underlying storage
    cannot grow; the allocator must reallocate.
\end{itemize}

\Lucid's \code{generate\_step} API uses pre-allocated by
default (max\_length is an argument); the grow-on-demand
variant is available via \code{max\_length=None} for
exploratory use.

\subsection{Quantised KV cache}
\label{ssec:attention:kv_quant}

For very long contexts, the FP16 KV cache becomes the dominant
memory cost. Quantising to INT8 halves the memory; INT4 quarters
it. The quality cost is small for inference (the
attention's softmax is robust to small per-key quantisation
noise) but is more delicate for training.

\Lucid\ 3.5 does not ship a built-in quantised KV cache; users
implement it by casting before storage and casting back at
read. The optimised path requires kernel-level quantised
matmul support that is not yet in MPSGraph.

\subsection{Sliding-window KV cache}
\label{ssec:attention:sliding_window_kv}

For applications with bounded context (e.g., streaming
inference of constant memory budget), the cache is a circular
buffer of fixed size $W$. Once $W$ tokens are in the cache,
each new token overwrites the oldest:

\begin{lstlisting}[style=lucid,language=LucidPython]
def update_sliding_cache(cache, new_kv, pos, W):
    cache[pos % W] = new_kv
    return cache
\end{lstlisting}

The attention mask is restricted to the in-cache positions.
Mistral-7B is the most prominent example of this pattern in
production (with $W = 4096$).

\section{Implementation Notes}
\label{sec:attention:impl_notes}

The attention modules live in \code{lucid/nn/modules/attention.py}.
The SDPA fused kernel is in \code{lucid/\_C/nn/Attention.cpp}.

\subsection{Memory profile}
\label{ssec:attention:memory_profile}

A single self-attention layer's memory footprint, breaking down
by tensor:
\begin{itemize}
  \item $Q, K, V$ projections: $3 \cdot B T d$ each.
  \item Score matrix: $B h T^2$ (only with naive path).
  \item Softmax outputs: same as score.
  \item Output: $B T d$.
\end{itemize}

For Transformer training, the score matrix is the dominant cost
when sequence length is comparable to or larger than the hidden
dimension. The SDPA fused kernel eliminates it.

\subsection{Numerical stability under FP16}
\label{ssec:attention:fp16_stability}

In FP16 AMP, the softmax can overflow if scores are large. The
fused SDPA kernel computes softmax in FP32 internally and casts
back to FP16 for the output; the user does not see the precision
juggling.

For the non-fused path under AMP, \Lucid's softmax has
\code{AmpPolicy::ForceFloat32} which achieves the same effect.

\subsection{Worked numerical example}
\label{ssec:attention:numerical_example}

Consider a single attention head with head dimension $d_h = 64$
and sequence length $T = 128$. Without the $\sqrt{d_h}$ scaling,
random $Q$ and $K$ with entries from $\mathcal{N}(0, 1)$ give
attention scores with:
\begin{itemize}
  \item Mean 0.
  \item Variance $d_h = 64$.
  \item Standard deviation $8$.
\end{itemize}
Softmax of scores with std 8 saturates: the largest score
dominates (one entry $\sim 1.0$, others $\sim 10^{-7}$). The
gradient through such a degenerate softmax is nearly zero.

With the $\sqrt{d_h} = 8$ scaling, scores have std 1: softmax
produces a more uniform distribution, and gradient signal
flows. The $\sqrt{d_h}$ factor is therefore not a cosmetic
detail but a numerical necessity.

The same calculation explains why low-rank attention works
(small effective $d_h$ is fine but needs different scaling) and
why some architectures use $\log d_h$ scaling instead (slightly
better for very large $d_h$).

\section{Attention Complexity}
\label{sec:attention:complexity}

The standard attention is $O(T^2 d)$ in compute and $O(T^2)$ in
memory (for the score matrix). Both numbers are problematic at
long sequences. The available levers:

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Variant & Compute & Memory \\
\midrule
Standard SDPA       & $O(T^2 d)$ & $O(T^2 + T d)$ \\
Flash-style fused   & $O(T^2 d)$ & $O(T \sqrt{T} + T d)$ \\
Sparse (sliding-window) & $O(T w d)$ & $O(T w + T d)$ \\
Sparse (block-diagonal) & $O(T^2 d / B)$ & $O(T^2 / B + T d)$ \\
Linear attention (kernelised) & $O(T d^2)$ & $O(d^2 + T d)$ \\
\bottomrule
\end{tabular}
\caption[Attention complexity.]{Compute and memory complexity
of attention variants. $T$ is sequence length, $d$ is hidden
dimension, $w$ is the local window for sparse attention, $B$ is
the block size for block-sparse. The Flash-style fused kernel
achieves the same compute as standard attention but with
$O(T \sqrt{T})$ memory --- the standard recommendation for
sequences up to $\sim$32K. Beyond that, sparse or linear
approximations are needed.}
\label{tab:attention:complexity}
\end{table}

\Lucid\ ships the standard SDPA (with the fused-kernel path
described in Section~\ref{sec:attention:sdpa_fused}). Sparse
and linear variants are not yet built-in; users implement them
on the masking interface or via custom Functions.

\section{Worked Example: Transformer Encoder Layer}
\label{sec:attention:transformer_encoder}

The standard Transformer encoder layer composes attention with
a feedforward network and normalisation:
\[
x' \;=\; x + \mathrm{MHA}(\mathrm{LN}(x)),
\qquad
y \;=\; x' + \mathrm{FFN}(\mathrm{LN}(x')),
\]
where the FFN is two linear layers with a GELU between:
\[
\mathrm{FFN}(z) \;=\; W_2\,\mathrm{GELU}(W_1\, z).
\]

In \Lucid:

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid.nn as nn

class EncoderLayer(nn.Module):
    def __init__(self, d=512, h=8, d_ff=2048):
        super().__init__()
        self.ln1 = nn.LayerNorm(d)
        self.attn = nn.MultiheadAttention(d, h, batch_first=True)
        self.ln2 = nn.LayerNorm(d)
        self.ffn = nn.Sequential(
            nn.Linear(d, d_ff),
            nn.GELU(),
            nn.Linear(d_ff, d),
        )

    def forward(self, x, mask=None):
        z = self.ln1(x)
        x = x + self.attn(z, z, z, attn_mask=mask)[0]
        x = x + self.ffn(self.ln2(x))
        return x
\end{lstlisting}

This is the pre-norm variant: layer norm is applied before the
sub-layer, not after. Modern implementations prefer pre-norm
because the residual stream is unchanged, which improves training
stability for deep stacks.

\section{Summary and Forward References}
\label{sec:attention:summary}

\Lucid's attention module
(\code{nn.MultiheadAttention}) implements scaled dot-product
attention with multi-head wrapping, causal and padding masks, and
the SDPA fused kernel for memory-efficient long sequences. The
chapter cited the original Transformer paper
\cite{vaswani_attention_2017} for the scaling factor's derivation
and RoPE \cite{su_roformer_2021} for the rotary position
encoding.

The next chapter, \chref{ch:losses}, treats loss functions, the
final piece of the supervised training loop.
\chref{ch:initialization} treats weight initialisation theory.
\chref{ch:optimizers} treats optimisers.

\begin{lucidnote}
For the user: \code{nn.MultiheadAttention} is parity-compatible
with \refframework. The SDPA fused kernel is the default; the
user does not need to explicitly opt in. The mask conventions
(boolean for padding, additive for custom) match \refframework.
\end{lucidnote}
