% !TEX root = ../main.tex
\chapter{Weight Initialization}
\label{ch:initialization}

The initial values of a network's parameters strongly affect
whether training converges. Initialising too small causes the
forward signal to vanish layer by layer (activations $\to 0$);
too large causes the signal to explode (activations $\to
\infty$); the gradient through the network suffers the same fate
in reverse. The standard initialisation schemes
(Xavier/Glorot, Kaiming/He, orthogonal) target an explicit
variance-preserving property that keeps signals and gradients in
a useful range across depth.

\Lucid\ exposes the standard initialisation surface at
\code{lucid.nn.init}, mirroring \refframework's
\code{torch.nn.init}. This chapter describes each scheme, the
mathematical motivation, and the user-side patterns for applying
them.

\section{Why Initialisation Matters}
\label{sec:initialization:why}

Consider a deep linear network $y = W_L W_{L-1} \cdots W_1 x$.
If each $W_\ell$ has spectral norm $\rho$, then $\|y\| \sim
\rho^L \|x\|$. For $L = 100$ and $\rho = 0.9$, $\|y\| \sim
2.6 \times 10^{-5} \|x\|$ --- vanishing. For $\rho = 1.1$,
$\|y\| \sim 1.4 \times 10^4 \|x\|$ --- exploding. Only $\rho
\approx 1$ keeps the signal in a useful range.

The same analysis applies to the backward pass: gradients scale
by the transposed matrix norms.

\subsection{The variance-preserving condition}
\label{ssec:initialization:var_preserving}

For random $W$ with i.i.d.\ entries from $\mathcal{N}(0,
\sigma^2)$ and random $x$ with $\mathrm{Var}(x_i) = 1$, the
forward signal at the next layer satisfies
\[
\mathrm{Var}(y_j) \;=\; n \sigma^2,
\]
where $n$ is the input dimension (the fan-in). To preserve
unit variance, we want $\sigma^2 = 1/n$, i.e.\ $\sigma =
1/\sqrt{n}$.

This is the original Xavier / Glorot insight: scale the
initialisation by the inverse square root of the fan-in.

\subsection{Derivation in detail}
\label{ssec:initialization:variance_derivation}

For a linear layer $y_j = \sum_i W_{ji} x_i$ with independent
$W_{ji}$ (mean 0, variance $\sigma^2$) and independent $x_i$
(mean 0, variance 1):
\begin{align*}
  \mathbb{E}[y_j] &= \sum_i \mathbb{E}[W_{ji}] \mathbb{E}[x_i] = 0, \\
  \mathrm{Var}(y_j) &= \sum_i \mathrm{Var}(W_{ji} x_i) \\
  &= \sum_i \mathbb{E}[W_{ji}^2] \mathbb{E}[x_i^2] - 0 \\
  &= n \sigma^2.
\end{align*}
Recursing across $L$ layers, $\mathrm{Var}(y^{(L)}) = (n \sigma^2)^L$.
For the signal to neither vanish nor explode, $n \sigma^2 = 1$,
giving $\sigma = 1/\sqrt{n}$.

The backward pass is symmetric: the gradient $\bar{x}_i = \sum_j W_{ji} \bar{y}_j$
has $\mathrm{Var}(\bar{x}_i) = n_{\text{out}} \sigma^2$, so for
backward variance preservation we want $\sigma = 1/\sqrt{n_{\text{out}}}$.
Forward and backward give different recommendations; Xavier
compromises by using the harmonic mean.

\section{Xavier (Glorot) Initialisation}
\label{sec:initialization:xavier}

For a linear layer with fan-in $n_{\text{in}}$ and fan-out
$n_{\text{out}}$, Xavier initialisation uses
\[
W_{ij} \;\sim\; \mathcal{N}\!\left(0,\, \frac{2}{n_{\text{in}} + n_{\text{out}}}\right)
\quad \text{or} \quad
W_{ij} \;\sim\; \mathcal{U}\!\left(-\sqrt{\frac{6}{n_{\text{in}} + n_{\text{out}}}},\, \sqrt{\frac{6}{n_{\text{in}} + n_{\text{out}}}}\right).
\]

The use of the average $(n_{\text{in}} + n_{\text{out}}) / 2$
rather than just $n_{\text{in}}$ balances forward and backward
variance: forward wants $1/n_{\text{in}}$, backward wants
$1/n_{\text{out}}$, the average is the compromise.

\subsection{The API}
\label{ssec:initialization:xavier_api}

\code{lucid.nn.init.xavier\_normal\_(tensor, gain=1.0)} and
\code{lucid.nn.init.xavier\_uniform\_(tensor, gain=1.0)}. The
\code{gain} multiplies the variance; defaults to 1 (linear or
identity), should be set higher for non-linear activations.

\subsection{Gain conventions}
\label{ssec:initialization:gain}

\code{nn.init.calculate\_gain(nonlinearity)} returns the
appropriate gain:
\begin{itemize}
  \item \code{linear}, \code{identity}: 1.
  \item \code{relu}: $\sqrt{2}$.
  \item \code{leaky\_relu(a)}: $\sqrt{2 / (1 + a^2)}$.
  \item \code{sigmoid}: 1.
  \item \code{tanh}: $5/3$.
  \item \code{selu}: $3/4$.
\end{itemize}

The gain compensates for the variance change introduced by the
non-linearity.

\section{Kaiming (He) Initialisation}
\label{sec:initialization:kaiming}

Xavier assumes a symmetric non-linearity that preserves
zero-centred Gaussian variance. ReLU is asymmetric: it zeros out
half the input, halving the variance. Kaiming initialisation
accounts for this:
\[
W_{ij} \;\sim\; \mathcal{N}\!\left(0,\, \frac{2}{n_{\text{in}}}\right).
\]
The factor of 2 (compared to $1/n_{\text{in}}$ for the
no-non-linearity case) compensates for ReLU's variance halving.

\subsection{The API}
\label{ssec:initialization:kaiming_api}

\code{nn.init.kaiming\_normal\_(tensor, a=0, mode='fan\_in', nonlinearity='leaky\_relu')}
and \code{kaiming\_uniform\_} of the same signature. The \code{mode}
chooses between fan-in (forward variance preservation) and
fan-out (backward variance preservation).

\subsection{The standard default}
\label{ssec:initialization:kaiming_default}

\Lucid's standard layer modules (Linear, Conv2d) use
\code{kaiming\_uniform\_(weight, a=sqrt(5))} as the default. This
matches \refframework's default and is empirically the choice
that works for most architectures.

\section{Orthogonal Initialisation}
\label{sec:initialization:orthogonal}

\code{nn.init.orthogonal\_(tensor, gain=1.0)} initialises a 2-D
tensor (or the first two axes of higher-rank) as an orthogonal
matrix scaled by \code{gain}. For square matrices, this means
$W^\top W = \mathrm{gain}^2 I$.

Orthogonal initialisation is especially useful for RNNs, where
the recurrent matrix $W_{hh}$ benefits from spectral radius
exactly 1 (orthogonal matrices have all singular values equal to
\code{gain}).

\subsection{Construction}
\label{ssec:initialization:orthogonal_construction}

The standard construction: sample a random Gaussian matrix, take
its QR decomposition, return $Q$. For non-square inputs, pad or
truncate appropriately.

\subsection{Why orthogonal for RNN}
\label{ssec:initialization:why_orth_rnn}

The vanilla RNN recurrence $h_t = \tanh(W_{hh} h_{t-1} + \ldots)$
involves $T$ products of $W_{hh}$. The norm of the gradient
through time is bounded by $\sigma(W_{hh})^T$ where
$\sigma$ is the largest singular value. For orthogonal $W_{hh}$,
all singular values equal 1, so the long-term gradient is
preserved (modulo the tanh saturation).

This is the original argument for using orthogonal
initialisation in RNNs, dating from the LSTM era. In practice,
LSTM and GRU also benefit, though to a lesser degree because
their gating mitigates the gradient issue independently.

\subsection{Sparse initialisation}
\label{ssec:initialization:sparse_init}

For wide layers, fully dense initialisation can over-couple all
neurons. \code{nn.init.sparse\_(tensor, sparsity, std=0.01)}
zeros out a fraction \code{sparsity} of each column and
initialises the rest from $\mathcal{N}(0, \mathrm{std}^2)$.
Less commonly used in modern architectures but historically
relevant for some early CNN papers.

\section{Simple Initialisations}
\label{sec:initialization:simple}

\Lucid\ provides the standard simple schemes:

\begin{itemize}
  \item \code{zeros\_(tensor)}: fill with 0.
  \item \code{ones\_(tensor)}: fill with 1. (Used for LayerNorm
    weight, BatchNorm weight.)
  \item \code{constant\_(tensor, val)}: fill with a constant.
  \item \code{eye\_(tensor)}: identity matrix (for 2-D tensors).
  \item \code{normal\_(tensor, mean=0, std=1)}: per-element
    Gaussian.
  \item \code{uniform\_(tensor, a=0, b=1)}: per-element uniform.
  \item \code{trunc\_normal\_(tensor, mean, std, a, b)}: truncated
    normal, redrawing samples outside $[a, b]$.
\end{itemize}

\subsection{trunc\_normal in transformers}
\label{ssec:initialization:trunc_normal}

The Transformer original paper \cite{vaswani_attention_2017}
and subsequent models initialise weights as
$\mathcal{N}(0, 0.02^2)$, truncated to $[-2 \sigma, 2 \sigma]$.
The truncation removes the heavy-tailed extreme samples that can
destabilise the start of training.

\code{nn.init.trunc\_normal\_(tensor, mean=0, std=0.02, a=-0.04, b=0.04)}
is the standard call.

\section{Bias Initialisation}
\label{sec:initialization:bias}

Biases are usually initialised to zero, except in three cases:

\begin{itemize}
  \item \textbf{LSTM forget gate bias = 1}: encourages the model
    to remember at the start of training, mitigating the cold-
    start vanishing-gradient problem.
  \item \textbf{Sigmoid output bias = \code{logit}(prior class
    probability)}: matches the dataset's class prior at
    initialisation.
  \item \textbf{Convolution bias = small constant (e.g., 0.01)}:
    occasionally used to avoid dead ReLUs.
\end{itemize}

The default \code{nn.Linear} / \code{nn.Conv2d} bias initialisation
is uniform in $[-1/\sqrt{n_{\text{in}}}, 1/\sqrt{n_{\text{in}}}]$,
which is small enough to be near-zero but provides some
asymmetry.

\section{Worked Example: VGG-Style Init}
\label{sec:initialization:vgg_example}

A typical CNN initialisation pattern:

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid.nn as nn
import lucid.nn.init as init

def init_weights(m):
    if isinstance(m, nn.Conv2d):
        init.kaiming_normal_(m.weight, nonlinearity='relu')
        if m.bias is not None:
            init.zeros_(m.bias)
    elif isinstance(m, nn.BatchNorm2d):
        init.ones_(m.weight)
        init.zeros_(m.bias)
    elif isinstance(m, nn.Linear):
        init.kaiming_normal_(m.weight, nonlinearity='relu')
        init.zeros_(m.bias)

model.apply(init_weights)
\end{lstlisting}

The pattern: Kaiming for the conv/linear weights (Re LU-aware),
ones/zeros for BatchNorm's affine parameters.

\section{Initialisation Recipes by Architecture}
\label{sec:initialization:recipes_by_arch}

\subsection{Standard CNN (ResNet, VGG, ConvNeXt)}
\label{ssec:initialization:cnn_recipe}

Standard recipe:
\begin{itemize}
  \item Convolutional weights: Kaiming normal, fan-out mode,
    nonlinearity=relu.
  \item Linear weights (classifier head): Kaiming normal or
    truncated normal.
  \item BatchNorm weight: 1; BatchNorm bias: 0.
  \item Convolutional / linear biases: 0.
\end{itemize}

ResNet \cite{he_resnet_2016} additionally initialises the last
BatchNorm in each residual block's weight to 0, so the block's
output is initially zero and the residual stream is unchanged.
This is the ``zero-init residual'' trick that lets very deep
ResNets train stably from scratch.

\subsection{Transformers (BERT, GPT-2, ViT)}
\label{ssec:initialization:transformer_recipe}

Standard recipe:
\begin{itemize}
  \item Linear weights: truncated normal $\sigma = 0.02$,
    bias 0.
  \item Embedding weights: same as linear.
  \item LayerNorm weight 1, bias 0.
\end{itemize}

GPT-2 \cite{radford_gpt2_2019} additionally scales the
initialisation of the residual projections (the second
linear in each FFN and the output of attention) by
$1/\sqrt{2 L}$ where $L$ is the depth. The motivation: with $L$
residual additions, the variance accumulates as $L$; the
$1/\sqrt{2L}$ scaling keeps the residual-stream variance
bounded across depth.

\subsection{LLaMA-family (RMSNorm, SwiGLU)}
\label{ssec:initialization:llama_recipe}

LLaMA-style \cite{touvron_llama_2023} initialisation:
\begin{itemize}
  \item Linear weights: truncated normal $\sigma = 0.02$,
    no bias (LLaMA omits biases).
  \item RMSNorm weight: 1.
  \item RoPE: no parameters (fixed sin/cos lookup).
\end{itemize}

The depth-scaling correction of GPT-2 is not applied directly;
LLaMA uses a slightly different residual scaling via the FFN
hidden dimension choice.

\subsection{Diffusion U-Net}
\label{ssec:initialization:diffusion_recipe}

Diffusion model U-Net \cite{ho_ddpm_2020}:
\begin{itemize}
  \item Standard CNN init for the conv layers.
  \item The final output convolution is initialised to zero
    weights and zero bias. The denoiser then starts as the
    identity (predicts zero noise), which is a sensible
    starting point.
  \item Time-embedding linear: standard init.
\end{itemize}

\section{Worked Example: Transformer Init}
\label{sec:initialization:transformer_example}

A typical Transformer initialisation:

\begin{lstlisting}[style=lucid,language=LucidPython]
def init_transformer(module):
    if isinstance(module, nn.Linear):
        init.trunc_normal_(module.weight, mean=0, std=0.02,
                           a=-0.04, b=0.04)
        if module.bias is not None:
            init.zeros_(module.bias)
    elif isinstance(module, nn.Embedding):
        init.trunc_normal_(module.weight, mean=0, std=0.02,
                           a=-0.04, b=0.04)
    elif isinstance(module, nn.LayerNorm):
        init.ones_(module.weight)
        init.zeros_(module.bias)

transformer.apply(init_transformer)
\end{lstlisting}

Truncated normal at the GPT-2 standard ($\sigma = 0.02$) for
both Linear and Embedding; LayerNorm to identity.

\section{Embedding Initialisation}
\label{sec:initialization:embedding}

Word and token embeddings have their own conventions:

\begin{itemize}
  \item Default: $\mathcal{N}(0, 1)$ or $\mathcal{U}(-0.1, 0.1)$,
    depending on the architecture.
  \item Transformer-style: truncated normal $\sigma = 0.02$.
  \item Tied to output projection: initialise once (Section
    \ref{ssec:module_system:shared_params}); both consumers
    use the shared tensor.
\end{itemize}

For positional embeddings, the convention is similar to weight
embeddings: small random initialisation, with the gradient
flowing through during training. Some recent recipes start
positional embeddings at exactly zero (no positional bias at
init) and let the model learn the appropriate values.

\section{Initialisation for AMP}
\label{sec:initialization:amp_init}

Mixed-precision training changes the initialisation
considerations slightly:

\begin{itemize}
  \item Default Xavier / Kaiming initialisation can produce
    weights at the very low end of FP16's representable range.
    Truncated normal $\sigma = 0.02$ is FP16-safe.
  \item LayerNorm weight = 1 is exactly representable.
  \item Initialising parameters in FP32 (then casting via
    autocast at forward) is the standard pattern; the master
    copy stays FP32 for optimiser precision.
\end{itemize}

In practice, the initialisations above all play nicely with
AMP without modification. The exception is the
\code{nn.init.constant\_(weight, very\_small)} pattern (e.g.,
$10^{-5}$); under FP16, this may round to a denormal or zero.
Use $\geq 10^{-4}$ as the smallest meaningful constant.

\section{Initialisation Theory: A Deeper Look}
\label{sec:initialization:theory}

The variance-preservation arguments above assume:
\begin{itemize}
  \item Inputs to each layer are zero-mean and unit-variance.
  \item Weights are i.i.d.
  \item The activation function is well-behaved.
\end{itemize}

In practice these assumptions are approximate. BatchNorm /
LayerNorm explicitly enforce the input assumption; residual
connections preserve the input distribution through depth (which
is why they let very deep networks train at all). The mathematics
of why deep ResNets can be trained at depth 100+ without careful
initialisation is the subject of dynamical-isometry analyses; we
do not pursue it here.

\subsection{Dynamical isometry}
\label{ssec:initialization:dynamical_isometry}

A network is said to have \emph{dynamical isometry} if its
forward and backward Jacobians have all singular values
approximately 1. This is the strongest variance-preservation
condition and the one that orthogonal initialisation explicitly
targets. For practical depths (10--100 layers), the standard
Kaiming + BatchNorm/LayerNorm combination achieves enough
approximate isometry to train; for very deep networks (1000+),
explicit orthogonality matters more.

\subsection{The signal/gradient covariance}
\label{ssec:initialization:signal_grad_cov}

A more refined analysis tracks both signal and gradient variance
through the layers. The forward variance at layer $\ell$ is
\[
\mathrm{Var}(x^{(\ell)}) \;\propto\; \prod_{k=1}^{\ell} (n_k \sigma_k^2 \cdot g_k),
\]
where $g_k$ is the activation's gain. The backward gradient
variance has the analogous form with $n_k$ replaced by
$n_{k+1}$. Xavier balances both; Kaiming targets just the
forward (and is fine because the backward inherits).

\section{Initialisation in lucid.nn}
\label{sec:initialization:nn_defaults}

The default initialisations in \Lucid's standard layers:

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Module & Weight init & Bias init \\
\midrule
\code{Linear}      & \code{kaiming\_uniform\_(a=sqrt(5))} & \code{uniform\_(-1/sqrt(n),...)} \\
\code{Conv1d/2d/3d} & \code{kaiming\_uniform\_(a=sqrt(5))} & same as Linear \\
\code{LayerNorm}   & \code{ones\_}                         & \code{zeros\_} \\
\code{BatchNorm}   & \code{ones\_}                         & \code{zeros\_} \\
\code{Embedding}   & \code{normal\_(0, 1)}                 & --- \\
\code{LSTM, GRU}   & \code{uniform\_(-1/sqrt(H), 1/sqrt(H))} & same \\
\code{MultiheadAttention} & \code{kaiming\_uniform\_(a=sqrt(5))} & \code{zeros\_} \\
\bottomrule
\end{tabular}
\caption[Default initialisations.]{Standard-layer default
weight and bias initialisations in \Lucid, matching
\refframework. Most architectures override at least the weight
init (e.g., truncated-normal for Transformers); the defaults are
sensible starting points but rarely optimal.}
\label{tab:initialization:defaults}
\end{table}

\section{Implementation Notes}
\label{sec:initialization:impl_notes}

The \code{lucid.nn.init} module lives in \code{lucid/nn/init.py}.
Each initialisation is an in-place operation on the tensor
(the trailing underscore in the name marks in-place by
convention). The operations are differentiable (an initialisation
that produces a tensor with \code{requires\_grad=True} retains
that flag), though gradient through initialisation is rarely
meaningful.

\subsection{Composability with autograd}
\label{ssec:initialization:autograd}

Initialisations are typically called once at model construction,
before training begins, and do not interact with autograd. If
called during training (e.g., to reset a layer), the in-place
operation increments the version counter
(\chref{ch:tensor_model}~Section~\ref{ssec:tensor_model:version_counter})
and may invalidate saved tensors for backward.

\section{Summary and Forward References}
\label{sec:initialization:summary}

Weight initialisation is the choice that lets gradient descent
converge from random starting points. The standard schemes
(Xavier, Kaiming, orthogonal, truncated normal) target variance
preservation across the forward and backward passes. \Lucid\
exposes them at \code{lucid.nn.init} with the same API as
\refframework. The standard layer modules in \code{lucid.nn} have
sensible default initialisations but can be overridden for
architecture-specific patterns (Transformer's truncated normal,
LSTM's forget-gate bias).

The next chapter, \chref{ch:optimizers}, treats the optimiser
family. \chref{ch:lr_schedulers} treats learning-rate schedules.
\chref{ch:amp} treats mixed precision.

\begin{lucidnote}
For the user: the standard layers have working defaults. For
production training, override with the architecture-appropriate
pattern (Kaiming for CNNs, truncated normal for Transformers,
orthogonal for RNNs). Apply via \code{model.apply(init\_fn)}.
\end{lucidnote}
