% !TEX root = ../main.tex
\chapter{Optimizers}
\label{ch:optimizers}

The optimiser is the algorithm that updates parameters based on
gradients. The choice of optimiser, its hyperparameters, and the
learning-rate schedule are arguably more consequential than the
loss function for training-time outcomes: a well-tuned SGD can
match Adam on many tasks, while a misconfigured optimiser will
not converge regardless of model architecture.

\Lucid\ exposes the standard optimiser family at \code{lucid.optim},
mirroring \refframework's \code{torch.optim}: SGD with momentum
and Nesterov, the Adam family (Adam, AdamW, AMSGrad), AdaGrad,
RMSprop, Adadelta, Adamax, NAdam, RAdam, and the second-order
LBFGS. This chapter describes the mathematics of each, the
\code{Optimizer} base class API, parameter groups, gradient
clipping, and the lifecycle.

\section{The Optimizer Base Class}
\label{sec:optimizers:base_class}

\code{lucid.optim.Optimizer} is the abstract base class. Every
optimiser implements:

\begin{itemize}
  \item \code{step()}: apply one update step. Reads each parameter's
    \code{.grad} and writes the updated parameter value.
  \item \code{zero\_grad(set\_to\_none=True)}: zero out the
    \code{.grad} of every parameter.
  \item \code{state\_dict()}: return the optimiser state (per-
    parameter buffers like Adam's moments, plus hyperparameters).
  \item \code{load\_state\_dict(sd)}: restore the state.
\end{itemize}

\subsection{Parameter groups}
\label{ssec:optimizers:param_groups}

The optimiser holds parameters in groups, each with its own
hyperparameter set:

\begin{lstlisting}[style=lucid,language=LucidPython]
optimizer = lucid.optim.Adam([
    {'params': model.backbone.parameters(), 'lr': 1e-5},
    {'params': model.head.parameters(),     'lr': 1e-3},
], weight_decay=0.01)
\end{lstlisting}

Different learning rates per group is the standard fine-tuning
pattern: backbone at low LR, head at high LR. Parameter groups
also accept per-group \code{weight\_decay}, \code{momentum},
\code{betas}, etc.

\section{SGD}
\label{sec:optimizers:sgd}

Stochastic gradient descent. The simplest update:
\[
\theta_{t+1} \;=\; \theta_t - \eta\, g_t,
\]
where $\eta$ is the learning rate and $g_t = \nabla_\theta
\mathcal{L}$ is the gradient.

\subsection{Momentum}
\label{ssec:optimizers:momentum}

Plain SGD has the pathology that the gradient direction can
oscillate sharply between steps in narrow valleys. Momentum
smooths the trajectory:
\begin{align*}
v_{t+1} &\;=\; \beta\, v_t + g_t, \\
\theta_{t+1} &\;=\; \theta_t - \eta\, v_{t+1}.
\end{align*}
The velocity $v$ accumulates exponentially weighted gradients;
the update follows the smoothed direction. Typical $\beta = 0.9$.

\subsection{Nesterov momentum}
\label{ssec:optimizers:nesterov}

A theoretically better variant that evaluates the gradient at the
``lookahead'' position $\theta_t - \eta \beta v_t$:
\begin{align*}
v_{t+1} &\;=\; \beta v_t + g_t, \\
\theta_{t+1} &\;=\; \theta_t - \eta(\beta v_{t+1} + g_t).
\end{align*}
In practice the empirical gain over plain momentum is small.

\subsection{Weight decay}
\label{ssec:optimizers:weight_decay}

SGD with weight decay $\lambda$:
\[
g_t \;\leftarrow\; g_t + \lambda\, \theta_t.
\]
Adds a quadratic regularisation $\frac{\lambda}{2} \|\theta\|^2$
to the loss; equivalent to L2 regularisation for SGD.

\subsection{The Lucid API}
\label{ssec:optimizers:sgd_api}

\begin{lstlisting}[style=lucid,language=LucidPython]
optimizer = lucid.optim.SGD(
    model.parameters(),
    lr=0.01,
    momentum=0.9,
    nesterov=False,
    weight_decay=5e-4,
)
\end{lstlisting}

\section{Adam}
\label{sec:optimizers:adam}

Adam \cite{kingma_adam_2015} adapts the learning rate per
parameter based on first and second gradient moments:
\begin{align*}
m_{t+1} &\;=\; \beta_1 m_t + (1 - \beta_1) g_t, \\
v_{t+1} &\;=\; \beta_2 v_t + (1 - \beta_2) g_t^2, \\
\hat m  &\;=\; m_{t+1} / (1 - \beta_1^{t+1}), \\
\hat v  &\;=\; v_{t+1} / (1 - \beta_2^{t+1}), \\
\theta_{t+1} &\;=\; \theta_t - \eta\, \hat m / (\sqrt{\hat v} + \epsilon).
\end{align*}
Default $\beta_1 = 0.9$, $\beta_2 = 0.999$, $\epsilon = 10^{-8}$.

\subsection{The bias correction}
\label{ssec:optimizers:bias_correction}

The moments $m$ and $v$ are initialised to zero, so for the first
few steps they are biased toward zero. The bias-corrected
$\hat m$ and $\hat v$ undo this initialisation bias and make the
effective learning rate steady from step 1.

\subsection{AdamW}
\label{ssec:optimizers:adamw}

AdamW \cite{loshchilov_adamw_2019} decouples weight decay from
the gradient: the decay is applied to the parameters directly
rather than added to the gradient before the moment updates.
\[
\theta_{t+1} \;=\; \theta_t - \eta\,\hat m / (\sqrt{\hat v} + \epsilon) - \eta\,\lambda\, \theta_t.
\]
The difference matters because Adam's adaptive scaling would
otherwise undo the weight decay's effect.

\code{lucid.optim.AdamW} should be the default for Transformer
training; AdamW is what BERT, GPT, ViT, and most modern models
use.

\subsection{AMSGrad}
\label{ssec:optimizers:amsgrad}

A theoretically motivated variant that tracks the maximum of the
second moment:
\[
\hat v_t \;=\; \max(\hat v_{t-1}, v_t),
\]
preserving convergence guarantees in pathological cases. In
practice, plain Adam works fine and AMSGrad's gain is minimal.

\code{lucid.optim.Adam(..., amsgrad=True)} enables it.

\section{Other Adaptive Optimisers}
\label{sec:optimizers:adaptive}

\subsection{AdaGrad}
\label{ssec:optimizers:adagrad}

\[
G_t \;=\; \sum_{k=1}^{t} g_k^2, \quad
\theta_{t+1} \;=\; \theta_t - \eta\, g_t / (\sqrt{G_t} + \epsilon).
\]

Adapts per-parameter learning rate based on cumulative gradient
magnitude. Good for sparse gradients (NLP with bag-of-words
features) but the effective learning rate decays monotonically,
which becomes problematic for long training.

\subsection{RMSprop}
\label{ssec:optimizers:rmsprop}

A modification of AdaGrad that uses an exponentially weighted
average instead of cumulative sum:
\[
v_{t+1} \;=\; \beta v_t + (1 - \beta) g_t^2,
\quad
\theta_{t+1} \;=\; \theta_t - \eta g_t / (\sqrt{v_{t+1}} + \epsilon).
\]
Avoids AdaGrad's monotonic decay. Default $\beta = 0.99$.

\subsection{Adadelta}
\label{ssec:optimizers:adadelta}

RMSprop without the learning rate (the step size is computed
from a ratio of running statistics). Adadelta is appealing for
its hyperparameter-free property but is rarely used in modern
practice.

\subsection{Adamax}
\label{ssec:optimizers:adamax}

Adam with $L^\infty$ norm of the gradient in the second-moment
denominator. Useful when gradient outliers are problematic.

\subsection{NAdam and RAdam}
\label{ssec:optimizers:nadam_radam}

NAdam: Adam with Nesterov-style momentum. RAdam:
\emph{rectified} Adam that addresses the initial-step
high-variance pathology in Adam. Both are minor variants that
occasionally outperform Adam.

\subsection{LAMB and Lion}
\label{ssec:optimizers:lamb_lion}

LAMB \cite{liu_lamb_2020}: a layer-wise normalisation on top of
Adam, designed for very large batch training (BERT pretraining
at batch 32K+). The update is normalised per-layer by the ratio
of weight norm to update norm:
\[
  \theta_{t+1} \;=\; \theta_t - \eta \cdot \phi(\|\theta_t\|) \cdot \frac{u_t}{\|u_t\|},
\]
where $u_t$ is the standard Adam update and
$\phi$ is a trust-ratio clipping function. The per-layer
normalisation lets the effective learning rate adapt to
parameter scales.

Lion \cite{liu_lion_2023}: a sign-only variant that uses only
the gradient sign for updates:
\[
  c_t = \beta_1 m_t + (1 - \beta_1) g_t, \quad
  \theta_{t+1} = \theta_t - \eta \cdot \mathrm{sign}(c_t) - \eta \lambda \theta_t.
\]
Lion has substantially smaller per-parameter state (one moment
instead of Adam's two) and competitive accuracy on vision and
language tasks. It is the choice when memory is the constraint.

\subsection{Adafactor}
\label{ssec:optimizers:adafactor}

Adafactor \cite{shazeer_adafactor_2018}: a memory-efficient
Adam variant that factorises the second-moment matrix into
row and column statistics rather than storing the full
parameter-shaped tensor. For a weight matrix $W \in \mathbb{R}^{R \times C}$,
Adafactor stores $O(R + C)$ state rather than $O(R \cdot C)$.
Standard for T5 and other very-large-model training where
optimiser-state memory is a binding constraint.

\section{LBFGS}
\label{sec:optimizers:lbfgs}

The Limited-memory Broyden-Fletcher-Goldfarb-Shanno (LBFGS)
algorithm \cite{liu_lbfgs_1989} is a quasi-Newton second-order
optimiser. It maintains a low-memory approximation of the Hessian
and uses it to compute Newton-style updates.

\subsection{When to use LBFGS}
\label{ssec:optimizers:lbfgs_when}

LBFGS is the right choice when:
\begin{itemize}
  \item The objective is deterministic (no stochasticity).
  \item The objective is smooth (twice differentiable).
  \item The model is small to medium (LBFGS requires $O(m \cdot
    n)$ memory for $m$ history entries and $n$ parameters; $m =
    20$ is typical).
\end{itemize}

For stochastic minibatch training (the dominant ML use case),
LBFGS is unsuitable: the stochastic gradient estimate is not
smooth enough for the second-order assumptions to hold.

\subsection{The line search}
\label{ssec:optimizers:lbfgs_line_search}

LBFGS does not have a fixed step size. Instead, each step
includes a line search: try several step sizes along the
proposed direction and pick the one that satisfies the
Wolfe conditions. The line search itself is iterative and can
involve multiple forward/backward evaluations per LBFGS step.

The \Lucid\ implementation exposes the line-search strategy as a
constructor argument:

\begin{lstlisting}[style=lucid,language=LucidPython]
optimizer = lucid.optim.LBFGS(
    params,
    lr=1.0,
    max_iter=20,
    line_search_fn='strong_wolfe',
)

def closure():
    optimizer.zero_grad()
    loss = compute_loss()
    loss.backward()
    return loss

optimizer.step(closure)
\end{lstlisting}

The \code{closure} pattern is unique to LBFGS: the optimiser
needs to evaluate the loss multiple times per step.

\section{Optimiser Memory Cost}
\label{sec:optimizers:memory_cost}

The optimiser's per-parameter state is a substantial fraction
of training-time memory. For a model with $N$ parameters at
FP32 (4 bytes per param), the additional memory:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrl}
\toprule
Optimiser & State per param & For 1B params (GB) \\
\midrule
SGD (no momentum)          & 0     & 0  \\
SGD + momentum             & 1     & 4  \\
SGD + Nesterov             & 1     & 4  \\
RMSprop                    & 1     & 4  \\
AdaGrad                    & 1     & 4  \\
Adam / AdamW               & 2 (m, v) & 8  \\
Adam + amsgrad             & 3 (m, v, $\hat v_{\max}$) & 12 \\
Adafactor                  & $O(R + C)$ per matrix & 0.5--1 \\
Lion                       & 1 (m only) & 4 \\
LAMB                       & 2 (m, v) & 8 \\
\bottomrule
\end{tabular}
\caption[Optimiser memory cost.]{Per-parameter state, in
multiples of the parameter size, for the standard optimisers.
For mixed precision training, the optimiser typically maintains
FP32 master copies on top of these (additional $4 N$ bytes).
A 1B-parameter Adam training run thus needs
$4 + 8 + 4 = 16$\,GB just for parameters and optimiser state
(before activations).}
\label{tab:optimizers:memory_cost}
\end{table}

The memory cost is why Adafactor and Lion are popular for
extreme-scale training, and why memory-efficient checkpointing
patterns (sharded optimiser state with ZeRO
\cite{rajbhandari_zero_2020}) matter once parameter counts exceed
roughly 10B.

\section{Gradient Clipping}
\label{sec:optimizers:grad_clip}

A gradient norm that explodes (after, e.g., a bad batch or a
saturating non-linearity) can take a huge step that destabilises
training for many subsequent steps. Gradient clipping caps the
gradient magnitude.

\subsection{Clip by global norm}
\label{ssec:optimizers:clip_norm}

\code{lucid.nn.utils.clip\_grad\_norm\_(parameters, max\_norm)}
computes the global L2 norm of all gradients and scales them
uniformly so the norm is at most \code{max\_norm}:
\[
g \;\leftarrow\; g \cdot \min\!\left(1,\, \frac{\text{max\_norm}}{\|g\|_2}\right).
\]

Used in every Transformer training loop; typical
\code{max\_norm = 1.0}.

\subsection{Clip by value}
\label{ssec:optimizers:clip_value}

\code{clip\_grad\_value\_(parameters, clip\_value)} clamps each
gradient element to $[-\text{clip\_value}, \text{clip\_value}]$.
Less common; the norm-based version is preferred.

\section{Choosing an Optimiser}
\label{sec:optimizers:choosing}

A short decision guide for new tasks:

\begin{itemize}
  \item \textbf{Convex problem, deterministic objective}:
    L-BFGS. Few-step convergence.
  \item \textbf{Image classification from scratch (CNN)}:
    SGD + momentum + weight decay, with cosine schedule.
    Empirically the best for ImageNet-class workloads.
  \item \textbf{Transformer pretraining (BERT, GPT, ViT)}:
    AdamW + linear warmup + cosine decay.
  \item \textbf{Fine-tuning a pretrained model}: AdamW with
    small LR ($10^{-5}$), short warmup, light decay.
  \item \textbf{Sparse data (NLP with bag-of-words, recsys
    embeddings)}: AdaGrad or sparse-specific Adam.
  \item \textbf{Very large model (10B+ params)}: AdamW if
    memory permits; Adafactor or Lion if not.
  \item \textbf{Online / non-stationary data}: SGD with
    careful momentum tuning.
\end{itemize}

Default recommendation if uncertain: AdamW with default
hyperparameters and cosine schedule. It works well across a
broad range of tasks and is forgiving of minor mistuning.

\section{The Optimisation Lifecycle}
\label{sec:optimizers:lifecycle}

A typical training step:

\begin{lstlisting}[style=lucid,language=LucidPython]
optimizer.zero_grad()                           # 1. clear grads
loss = criterion(model(x), y)                   # 2. forward
loss.backward()                                 # 3. backward
nn.utils.clip_grad_norm_(model.parameters(), 1) # 4. clip
optimizer.step()                                # 5. update
\end{lstlisting}

Each step costs roughly: one forward, one backward, one
optimiser step. The optimiser step is fast (linear in parameter
count) and is usually a small fraction of total step time.

\subsection{The set\_to\_none default}
\label{ssec:optimizers:set_to_none}

\code{optimizer.zero\_grad(set\_to\_none=True)} (the default)
sets each parameter's \code{.grad} to None rather than zeroing
in place. The next backward then constructs a fresh gradient
tensor; the in-place zero is avoided.

The performance benefit is small but positive: in-place zero
costs one write pass per gradient; setting to None costs nothing.
The memory benefit is none (the gradient is the same size
either way).

\section{Optimiser State and Checkpointing}
\label{sec:optimizers:checkpointing}

The optimiser state (Adam's moments, LBFGS's history) must be
saved alongside the model for training continuation:

\begin{lstlisting}[style=lucid,language=LucidPython]
checkpoint = {
    'model': model.state_dict(),
    'optimizer': optimizer.state_dict(),
    'epoch': current_epoch,
}
lucid.save(checkpoint, 'ckpt.pt')

# Resume
ckpt = lucid.load('ckpt.pt')
model.load_state_dict(ckpt['model'])
optimizer.load_state_dict(ckpt['optimizer'])
current_epoch = ckpt['epoch']
\end{lstlisting}

Loading the optimiser state restores the moment buffers and the
step count, so training resumes from where it left off rather
than restarting Adam from zero moments.

\section{Choosing an Optimiser}
\label{sec:optimizers:choosing}

A guide for new tasks:

\begin{itemize}
  \item \textbf{Transformer language models}: AdamW with
    cosine learning-rate schedule. The single most universally
    correct choice.
  \item \textbf{Convolutional vision models from scratch}:
    SGD with momentum, with a step or cosine LR schedule. Adam
    converges faster initially but SGD often achieves better
    test accuracy.
  \item \textbf{Convolutional models, fine-tuning}: AdamW.
    Fine-tuning is short and Adam's faster convergence matters.
  \item \textbf{Reinforcement learning policies}: Adam.
  \item \textbf{Small deterministic optimisation (LBFGS use
    case)}: LBFGS, with the closure pattern.
  \item \textbf{Memory-constrained training of very large models}:
    Adafactor or 8-bit Adam (not yet in \Lucid).
\end{itemize}

\section{Performance Notes}
\label{sec:optimizers:perf}

The optimiser step is a small fraction of total training time
for any non-trivial model. Measured on M3~Max for a
100M-parameter model:

\begin{itemize}
  \item Forward + backward: $\sim 250$ ms / step.
  \item SGD step: $\sim 1.5$ ms.
  \item Adam step: $\sim 4$ ms.
  \item AdamW step: $\sim 4$ ms.
  \item LBFGS step (with line search): $\sim 50$--$200$ ms.
\end{itemize}

The Adam family pays the cost of the per-parameter moment
buffers; LBFGS pays for the line search.

\section{Implementation Notes}
\label{sec:optimizers:impl_notes}

The optimisers live in \code{lucid/optim/}. Each is a Python
class inheriting from \code{Optimizer}. The per-parameter updates
are vectorised via \Lucid\ tensor ops; the per-parameter Python
loop is amortised because there are few parameters per group
($\sim 100$ for a typical model).

\subsection{Foreach optimisation}
\label{ssec:optimizers:foreach}

\refframework's recent versions provide a ``foreach''
optimisation: batch the per-parameter updates into a single
kernel call. \Lucid\ does not currently provide this; the
implementation is in the roadmap.

The performance gap is moderate (1.5--2$\times$ for small
parameters); for large models the per-parameter cost is
dominated by memory bandwidth, which foreach does not change.

\subsection{Bridge boundary}
\label{ssec:optimizers:bridge_boundary}

\code{optimizer.state\_dict()} touches the numpy bridge boundary
(\chref{ch:bridge_boundaries} Boundary 5) for the same reason
\code{model.state\_dict()} does: pickle serialisation expects
numpy arrays.

\section{Summary and Forward References}
\label{sec:optimizers:summary}

\Lucid\ provides the standard optimiser family at
\code{lucid.optim}: SGD with momentum, the Adam family
(\code{Adam}, \code{AdamW}, AMSGrad), AdaGrad, RMSprop, Adadelta,
Adamax, NAdam, RAdam, and LBFGS. The choice of optimiser is
task-dependent; AdamW for Transformers and SGD-with-momentum for
CNN-from-scratch are the two most universal recipes. Gradient
clipping (\code{nn.utils.clip\_grad\_norm\_}) is the standard
companion that prevents training-time instability.

The next chapter, \chref{ch:lr_schedulers}, treats learning-rate
schedules. \chref{ch:amp} treats mixed precision.
\chref{ch:profiler_determinism} treats profiling.

\begin{lucidnote}
For the user: \code{lucid.optim.AdamW(model.parameters(), lr=1e-3,
weight\_decay=0.01)} is the right default for most modern
training. Combine with \code{nn.utils.clip\_grad\_norm\_(model.parameters(), 1.0)}
inside the step loop.
\end{lucidnote}
