% !TEX root = ../main.tex
\chapter{Learning-Rate Schedulers}
\label{ch:lr_schedulers}

The learning rate is the single most important hyperparameter
in gradient-based training. A constant learning rate is rarely
optimal: training benefits from a large initial rate to make fast
progress and a small final rate to settle into a precise minimum.
The \emph{learning-rate schedule} prescribes how the rate
changes over the course of training.

\Lucid\ exposes the standard scheduler family at
\code{lucid.optim.lr\_scheduler}, mirroring \refframework. This
chapter describes the main scheduler types, the formulas, the
typical use cases, and the warm-up pattern that begins most
modern training runs.

\section{The Scheduler API}
\label{sec:lr_schedulers:api}

\code{lucid.optim.lr\_scheduler} schedulers wrap an existing
optimiser:

\begin{lstlisting}[style=lucid,language=LucidPython]
optimizer = lucid.optim.AdamW(model.parameters(), lr=1e-3)
scheduler = lucid.optim.lr_scheduler.CosineAnnealingLR(
    optimizer, T_max=100, eta_min=1e-6)

for epoch in range(100):
    train_one_epoch(...)
    scheduler.step()  # advance schedule
\end{lstlisting}

Each \code{step()} call advances the schedule one unit. The
update modifies the optimiser's parameter groups' \code{lr}
fields in place; subsequent \code{optimizer.step()} calls use the
updated rate.

\subsection{Per-epoch vs per-step}
\label{ssec:lr_schedulers:per_epoch_vs_step}

The convention varies: some schedulers (\code{StepLR},
\code{MultiStepLR}, \code{CosineAnnealingLR}) advance once per
epoch; others (\code{OneCycleLR}, \code{CosineAnnealingWarmRestarts})
once per training step. The user must place the \code{scheduler.step()}
call appropriately.

\section{Step-Based Schedules}
\label{sec:lr_schedulers:step_based}

\subsection{StepLR}
\label{ssec:lr_schedulers:steplr}

\code{StepLR(optimizer, step\_size, gamma=0.1)}: every
\code{step\_size} epochs, multiply the LR by \code{gamma}:
\[
\eta_t \;=\; \eta_0 \cdot \gamma^{\lfloor t / \text{step\_size} \rfloor}.
\]

Classical convention for CNN training: $\gamma = 0.1$,
\code{step\_size} = 30 (drop by 10$\times$ every 30 epochs over a
90-epoch run).

\subsection{MultiStepLR}
\label{ssec:lr_schedulers:multisteplr}

\code{MultiStepLR(optimizer, milestones=[30, 60, 90], gamma=0.1)}:
multiply by gamma at each listed epoch. More flexible than
\code{StepLR}.

\subsection{ExponentialLR}
\label{ssec:lr_schedulers:exponentiallr}

\code{ExponentialLR(optimizer, gamma)}: continuous exponential
decay,
\[
\eta_t \;=\; \eta_0 \cdot \gamma^t.
\]

Smooth alternative to the discrete steps. Typical
$\gamma = 0.95$ (about 5\% decay per epoch).

\section{Cosine Schedule}
\label{sec:lr_schedulers:cosine}

\code{CosineAnnealingLR(optimizer, T\_max, eta\_min=0)}: smoothly
decay from $\eta_0$ to $\eta_{\min}$ following a cosine:
\[
\eta_t \;=\; \eta_{\min} + \tfrac{1}{2}(\eta_0 - \eta_{\min})
\left(1 + \cos\left(\frac{\pi t}{T_{\max}}\right)\right).
\]

The cosine decay is the modern standard for Transformer training
(and many CNN training procedures). The slow decay at the start
preserves the high LR's exploration; the fast decay at the end
settles into precise updates.

\subsection{Cosine with warm restarts}
\label{ssec:lr_schedulers:cosine_warm_restart}

\code{CosineAnnealingWarmRestarts(optimizer, T\_0, T\_mult=1)}:
periodic cosine decay with restart. After each $T_0$ epochs, the
LR jumps back to $\eta_0$; the next period is $T_0 \cdot T_{\text{mult}}$
epochs long; and so on.

Useful when the loss landscape has multiple minima and the
restarts help escape local ones. Less common than plain cosine
in modern practice.

\section{Plateau-Triggered}
\label{sec:lr_schedulers:plateau}

\code{ReduceLROnPlateau(optimizer, mode='min', factor=0.1, patience=10)}:
monitor a metric (validation loss); if it has not improved for
\code{patience} epochs, multiply the LR by \code{factor}.

Useful when the right training duration is unknown: training
continues as long as progress is being made, with the LR
adapting automatically.

\subsection{The metric argument}
\label{ssec:lr_schedulers:plateau_metric}

\code{ReduceLROnPlateau.step(metric)} takes the latest metric
value as an argument (unlike other schedulers, which take no
argument). The user is responsible for passing the appropriate
metric.

\section{Warm-Up}
\label{sec:lr_schedulers:warmup}

For very large batch sizes or very large models, the first few
steps of training are unstable: the initial gradient direction
can be wildly off due to random initialisation, and a large LR
step can destabilise. The \emph{warm-up} pattern linearly
increases the LR from a small value (or zero) to the target over
the first $W$ steps:
\[
\eta_t \;=\; \begin{cases}
\eta_0 \cdot t / W & t < W, \\
\eta_{\text{schedule}}(t - W) & t \geq W.
\end{cases}
\]

\subsection{The combined warmup + cosine}
\label{ssec:lr_schedulers:warmup_cosine}

The dominant pattern in modern Transformer training: warm up for
$W$ steps (typically 1--10\% of total training), then cosine
decay over the remainder. \Lucid\ does not yet ship a built-in
combined scheduler; the user composes:

\begin{lstlisting}[style=lucid,language=LucidPython]
from lucid.optim.lr_scheduler import LambdaLR

def lr_lambda(step):
    if step < warmup_steps:
        return step / warmup_steps
    progress = (step - warmup_steps) / (total_steps - warmup_steps)
    return 0.5 * (1 + lucid.math.cos(lucid.math.pi * progress))

scheduler = LambdaLR(optimizer, lr_lambda)
\end{lstlisting}

The \code{LambdaLR} accepts a function from step to LR multiplier,
returning a value in $[0, 1]$ that scales $\eta_0$.

\section{One-Cycle}
\label{sec:lr_schedulers:one_cycle}

\code{OneCycleLR(optimizer, max\_lr, total\_steps)}: warm up to
\code{max\_lr} over half the training, then anneal to zero over
the other half. The schedule shape is symmetric.

The motivation \cite{smith_one_cycle_2018}: large LRs early in
training help find good parameters; small LRs late help settle.
The one-cycle policy is empirically effective and reduces the
number of hyperparameters to tune.

The full one-cycle has three phases:

\begin{enumerate}
  \item \textbf{Warm-up} ($t \in [0, 0.3 T]$): LR linearly
    rises from $\max\_lr / 25$ to $\max\_lr$, with momentum
    descending in parallel from 0.95 to 0.85.
  \item \textbf{Annealing} ($t \in [0.3 T, T]$): cosine decay
    from $\max\_lr$ down to $\max\_lr / 10^4$, with momentum
    rising back to 0.95.
  \item \textbf{Final descent} (optional): if specified, a
    final very small LR for the last few epochs to settle.
\end{enumerate}

The coupling of LR and momentum (high LR with low momentum;
low LR with high momentum) is part of the recipe: the high LR
benefits from rapid exploration (low momentum doesn't damp the
new directions), the low LR benefits from inertia (high momentum
smooths the small steps).

\subsection{LR finder}
\label{ssec:lr_schedulers:lr_finder}

Choosing $\max\_lr$ for OneCycle is the one tunable; the LR
finder \cite{smith_one_cycle_2018} automates it. The finder
runs a brief training (one epoch or fewer) with LR exponentially
increasing from very small to very large; the loss curve as a
function of LR shows a characteristic shape:

\begin{itemize}
  \item Flat at small LR (negligible progress).
  \item Sharply decreasing in a useful band.
  \item Sharply increasing at too-large LR (training diverges).
\end{itemize}

The recommended $\max\_lr$ is roughly 1 decade below the
divergence point. \Lucid\ does not ship a built-in LR finder
but the recipe is easy to implement with
\code{LambdaLR(lambda step: 1.1 ** step)} and loss logging.

\subsection{SGDR (warm restarts)}
\label{ssec:lr_schedulers:sgdr}

\code{CosineAnnealingWarmRestarts} (mentioned above) is the
SGDR \cite{loshchilov_sgdr_2017} schedule: periodic cosine
decay with the period doubling each restart:
\[
  \eta_t = \eta_{\min} + \tfrac{1}{2}(\eta_{\max} - \eta_{\min})
  \left(1 + \cos\left(\frac{\pi t_{\text{cur}}}{T_i}\right)\right),
\]
where $T_i = T_0 \cdot T_{\text{mult}}^i$ is the length of the
$i$-th cycle. The restarts let the optimiser escape the
attractor basin of one minimum and explore others. SGDR is
particularly effective for SGD; for AdamW the benefit is less
clear.

\section{LambdaLR}
\label{sec:lr_schedulers:lambda}

\code{LambdaLR(optimizer, lr\_lambda)}: most general. The
function \code{lr\_lambda(step)} returns a multiplier on
$\eta_0$. Useful for custom schedules that don't fit the
predefined patterns.

\section{The Schedule, in Effect}
\label{sec:lr_schedulers:visualisation}

\figref{fig:lr_schedulers:shapes} sketches the shape of the main
scheduler types.

\begin{figure}[t]
\centering
\begin{adjustbox}{max width=\textwidth, center}
\begin{tikzpicture}[
  ax/.style={->, >=stealth, thick},
  curve/.style={very thick, color=lucid.accent.dark},
  drop/.style={very thick, color=lucid.warn.fg},
  exp/.style={very thick, color=lucid.ok.fg},
  cos/.style={very thick, color=lucid.info.fg, dashed},
  lbl/.style={font=\small\sffamily, align=center},
]
  % Axes
  \draw[ax] (0, 0) -- (10.5, 0) node[right] {step};
  \draw[ax] (0, 0) -- (0, 3.2) node[above] {$\eta$};
  \draw (0, 3) -- (-0.15, 3) node[left] {$\eta_0$};

  % Step LR (red staircase)
  \draw[drop] (0, 3) -- (3, 3) -- (3, 2) -- (6, 2) -- (6, 1) -- (10, 1);

  % Exponential (green smooth)
  \draw[exp] plot[domain=0:10, samples=40] (\x, {3 * exp(-0.25 * \x)});

  % Cosine (blue dashed)
  \draw[cos] plot[domain=0:10, samples=40] (\x, {1.5 + 1.5 * cos(180 * \x / 10)});

  % Legend
  \node[lbl, anchor=west] at (5, 2.7) {{\color{lucid.warn.fg}StepLR}};
  \node[lbl, anchor=west] at (5, 0.6) {{\color{lucid.ok.fg}ExponentialLR}};
  \node[lbl, anchor=west] at (5, 1.5) {{\color{lucid.info.fg}CosineAnnealingLR}};
\end{tikzpicture}
\end{adjustbox}
\caption[Common LR schedule shapes.]{Three common learning-rate
schedules over a fixed-length training run. \code{StepLR} is the
staircase (sharp drops at fixed intervals); \code{ExponentialLR}
is the smooth exponential decay; \code{CosineAnnealingLR} is the
cosine arc that decays slowly at first and fast at the end.}
\label{fig:lr_schedulers:shapes}
\end{figure}

\section{Why Schedules Matter}
\label{sec:lr_schedulers:why_matter}

A constant learning rate has two failure modes:
\begin{itemize}
  \item Too small: training is slow; the model converges to a
    point but not as far as it could.
  \item Too large: training oscillates around the minimum without
    settling; final accuracy is lower than achievable.
\end{itemize}

Schedules combine the best of both: large rate early (fast
progress), small rate late (precise convergence). The optimal
schedule depends on the loss landscape, the batch size, and the
total training budget; the standard recipes (step, cosine,
one-cycle) are reasonable defaults that have been validated
across many tasks.

\section{Schedule Recipes by Architecture}
\label{sec:lr_schedulers:recipes}

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Architecture & Optimiser & Schedule \\
\midrule
ResNet ImageNet            & SGD + momentum  & StepLR (drop 10$\times$ at 30/60/90) \\
ResNet ImageNet (modern)   & SGD + Nesterov  & Cosine, no warmup \\
ConvNeXt ImageNet          & AdamW           & Cosine, 20-epoch warmup \\
ViT ImageNet               & AdamW           & Cosine, 10K-step warmup \\
BERT pretrain              & AdamW           & Linear warmup + linear decay \\
GPT-2 pretrain             & AdamW           & Cosine, 2K-step warmup \\
LLaMA pretrain             & AdamW           & Cosine, 2K-step warmup, $\eta_{\min} = 0.1 \eta_{\max}$ \\
Diffusion (DDPM)           & Adam            & Constant (no schedule) \\
DETR detection             & AdamW           & MultiStepLR drop at epoch 200 \\
Fine-tuning (any backbone) & AdamW           & Constant or cosine, very small LR \\
\bottomrule
\end{tabular}
\caption[LR schedule by architecture.]{The standard
LR-schedule recipes for common architectures and training
tasks. The modern default for new architectures is
``AdamW + cosine + short linear warmup''. Older recipes
(StepLR for ResNet) still work but are no longer recommended
for fresh runs.}
\label{tab:lr_schedulers:recipes}
\end{table}

The warmup duration scales with batch size: larger batches
need longer warmup. The rule of thumb: warmup over the first
1--5\% of training. For BERT-base pretraining at batch 256
for 1M steps, warmup is typically 10K steps (1\%).

\section{Linear Warmup + Linear Decay}
\label{sec:lr_schedulers:linear_warmup_decay}

A common BERT/T5 alternative to warmup+cosine: warmup linearly
then decay linearly to zero:
\[
  \eta_t = \begin{cases}
    \eta_{\max} \cdot t / W & t < W, \\
    \eta_{\max} \cdot \max(0, (T - t) / (T - W)) & t \geq W.
  \end{cases}
\]
At $t = T$, the LR is exactly zero. This produces a sharper
ending than cosine but works comparably in practice.

\code{LambdaLR} implements it directly:

\begin{lstlisting}[style=lucid,language=LucidPython]
def lr_lambda(step):
    if step < warmup_steps:
        return step / warmup_steps
    return max(0.0, (total_steps - step) / (total_steps - warmup_steps))

scheduler = LambdaLR(optimizer, lr_lambda)
\end{lstlisting}

\section{Chaining and Sequential Schedules}
\label{sec:lr_schedulers:chaining}

\code{SequentialLR(optimizer, schedulers, milestones)}: applies
\code{schedulers[0]} for the first \code{milestones[0]} steps,
then \code{schedulers[1]} for the next interval, etc.

Useful for warmup + main schedule: a linear warmup followed by
cosine decay can be expressed as two scheduler objects chained
together.

\code{ChainedScheduler(schedulers)}: applies multiple schedulers
simultaneously (each step calls each scheduler's step). Used
when separate parameter groups have different schedules.

\section{Polynomial Decay}
\label{sec:lr_schedulers:polynomial}

\code{PolynomialLR(optimizer, total\_iters, power=1.0)}:
polynomial decay to zero,
\[
  \eta_t = \eta_0 \cdot (1 - t/T)^p.
\]
With $p = 1$ this is linear; $p = 2$ is quadratic, faster
initially and slower at the end; $p = 0.5$ is the opposite.
Used in some BERT-pretraining recipes and in certain
reinforcement-learning setups.

\section{Cyclic Schedules}
\label{sec:lr_schedulers:cyclic}

\code{CyclicLR(optimizer, base\_lr, max\_lr, step\_size\_up)}:
LR oscillates triangularly between \code{base\_lr} and
\code{max\_lr} every \code{2 * step\_size\_up} iterations:
\[
  \eta_t = \eta_{\min} + (\eta_{\max} - \eta_{\min}) \cdot
  \max(0, 1 - |t \bmod (2W) - W| / W).
\]
The intuition: oscillating LR helps the optimiser escape local
minima. Variants include triangular2 (each cycle's amplitude
halves) and exp\_range (amplitude decays exponentially).

Used less in modern recipes than cosine + warmup, but the
underlying ``vary the LR aggressively'' principle is the same.

\section{Implementation Notes}
\label{sec:lr_schedulers:impl_notes}

The schedulers live in \code{lucid/optim/lr\_scheduler.py}. Each
inherits from \code{\_LRScheduler}, which provides the
\code{step()} mechanics and the \code{get\_lr()} hook that
subclasses override.

\subsection{Performance}
\label{ssec:lr_schedulers:perf}

The scheduler's per-step cost is negligible (a few microseconds
to compute the next LR and write it into the parameter groups).
Not a performance concern at any scale.

\section{Summary and Forward References}
\label{sec:lr_schedulers:summary}

\Lucid\ provides the standard learning-rate scheduler family at
\code{lucid.optim.lr\_scheduler}: step / multi-step / exponential
discrete decays, cosine annealing (with and without warm
restarts), plateau-triggered reduction, one-cycle, and the
general \code{LambdaLR} for custom schedules. The warmup + cosine
pattern is the modern default for Transformer training.

The next chapter, \chref{ch:amp}, treats mixed precision (which
interacts with the LR schedule when using gradient scaling).
\chref{ch:profiler_determinism} treats profiling and determinism.

\begin{lucidnote}
For the user: \code{CosineAnnealingLR} is the default for
Transformer training, with \code{LambdaLR} for the
warmup-plus-cosine variant. \code{StepLR} is the classical
choice for CNN training-from-scratch.
\end{lucidnote}
