% !TEX root = ../main.tex
\chapter{Mixed Precision and Gradient Scaling}
\label{ch:amp}

Modern hardware computes FP16 (or BF16) matrix products
substantially faster than FP32: the half-precision tensor cores
on NVIDIA hardware are 2--8$\times$ faster than the FP32 cores,
and Apple silicon's GPU exhibits a similar (though smaller)
asymmetry. \emph{Mixed precision training} exploits this by
running most of the forward and backward passes in half precision
while keeping a few sensitive operations (notably the parameter
update and certain reductions) in FP32. The result is a
substantial speedup with negligible accuracy loss
\cite{micikevicius_mixed_2018}.

This chapter describes \Lucid's AMP (automatic mixed precision)
system: the \code{autocast} context manager, the
\code{GradScaler} for loss scaling, the per-op AMP policy that
the schema declares, and the workflow patterns.

\section{Why Mixed Precision}
\label{sec:amp:why}

FP16 has a smaller representable range than FP32 (max $\sim 65504$
vs $\sim 3.4 \times 10^{38}$) and lower precision (10 mantissa
bits vs 23). The smaller range causes two specific failure modes
in deep learning:

\begin{itemize}
  \item \textbf{Overflow}: large activations or weights exceed the
    FP16 max. Manifests as \code{Inf} in the forward.
  \item \textbf{Underflow}: small gradients fall below the FP16
    min (smallest normal $\sim 6.1 \times 10^{-5}$). Manifests
    as gradients silently becoming zero.
\end{itemize}

Without mitigation, FP16 training is unreliable. AMP's
contribution is the engineering pattern that keeps the speedup
while addressing both failure modes.

\subsection{Floating-point format anatomy}
\label{ssec:amp:fp_anatomy}

The three relevant IEEE 754 formats:

\begin{table}[t]
\centering
\small
\begin{tabular}{lcccc}
\toprule
Format & Total bits & Exponent & Mantissa & Range $|x|$ \\
\midrule
FP32   & 32  & 8 & 23 & $1.2 \cdot 10^{-38}$ to $3.4 \cdot 10^{38}$ \\
FP16   & 16  & 5 & 10 & $6.1 \cdot 10^{-5}$  to $6.5 \cdot 10^{4}$ \\
BF16   & 16  & 8 & 7  & $1.2 \cdot 10^{-38}$ to $3.4 \cdot 10^{38}$ \\
FP8 E4M3 & 8 & 4 & 3 & $1.95 \cdot 10^{-3}$ to $448$ \\
FP8 E5M2 & 8 & 5 & 2 & $6.1 \cdot 10^{-5}$ to $5.7 \cdot 10^{4}$ \\
\bottomrule
\end{tabular}
\caption[IEEE 754 format comparison.]{Storage layout and
representable range for the floating-point formats relevant to
mixed-precision training. BF16 trades FP16's mantissa precision
for FP32-equivalent range. The two FP8 variants
\cite{micikevicius_fp8_2022} differ in their exponent/mantissa
split, optimised for different uses (E4M3 for activations,
E5M2 for gradients).}
\label{tab:amp:fp_formats}
\end{table}

The smallest normal positive of FP16, $\sim 6 \times 10^{-5}$,
is the relevant cliff for gradient underflow. Real deep
learning gradients regularly include values $10^{-6}$--$10^{-8}$;
without loss scaling, FP16 silently rounds these to zero,
killing learning signal.

\subsection{Underflow probability calculation}
\label{ssec:amp:underflow_calc}

A rough calculation: for ResNet-50 training, the gradient
distribution (across all parameters at a typical step) is
approximately log-normal with mean $-5$ and standard deviation
$2$ on the $\log_{10}$ scale. The probability of a gradient
$< 6 \times 10^{-5} = 10^{-4.2}$ is then the cumulative density
of $\log_{10}|g|$ below $-4.2$, which under our log-normal
gives $\sim 36\%$ of gradients underflowing.

A loss scale of $2^{15} = 32768$ multiplies each gradient by
$2^{15}$, shifting the distribution by $+4.5$ on the
$\log_{10}$ scale. The new underflow probability drops to
$<0.1\%$ --- a negligible loss of signal. The scale's role is
quantitative: shift the gradient distribution into the
representable range without hitting the overflow end.

\subsection{The two mitigations}
\label{ssec:amp:mitigations}

\begin{enumerate}
  \item \textbf{Per-op precision policy}: matrix multiplications
    run in FP16 (fast); reductions, normalisations, and softmax
    run in FP32 (precise). The selection is per-op, based on the
    op schema's \code{AmpPolicy}.
  \item \textbf{Loss scaling}: multiply the loss by a large
    scalar before backward, so the small gradients are shifted
    into the FP16 representable range. Divide gradients by the
    same scalar before the optimiser step.
\end{enumerate}

\section{BF16 vs FP16}
\label{sec:amp:bf16_vs_fp16}

The bfloat16 format \cite{kalamkar_bfloat16_2019} has the same
8-bit exponent as FP32 (range matches) but only 7 mantissa bits
(precision lower than FP16's 10). The trade-off:

\begin{itemize}
  \item \textbf{FP16}: better precision, narrower range. Needs
    loss scaling.
  \item \textbf{BF16}: same range as FP32, lower precision. No
    loss scaling needed.
\end{itemize}

Modern hardware (NVIDIA Ampere+, Apple M3+) supports BF16, and
the convention has been shifting toward BF16-default training
because the no-loss-scaling simplicity is worth the precision
loss for most workloads.

\Lucid\ supports both. The user picks via the \code{autocast}
dtype:

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.amp.autocast(dtype=lucid.float16):  # or bfloat16
    output = model(input)
    loss = criterion(output, target)
\end{lstlisting}

\section{autocast}
\label{sec:amp:autocast}

\code{lucid.amp.autocast(dtype=lucid.float16)} is a context
manager that switches the dispatcher into AMP mode. Inside the
context, each op is cast according to its schema's
\code{AmpPolicy} (\chref{ch:op_registry}):

\begin{itemize}
  \item \code{CastToHalf}: cast inputs to FP16 (matmul, conv,
    linear). The bulk of compute.
  \item \code{ForceFloat32}: keep inputs in FP32 (softmax,
    layernorm reductions, loss functions). Numerically sensitive.
  \item \code{PromoteToWidest}: standard promotion (most
    elementwise ops). Behaves like outside autocast.
  \item \code{PreserveInputs}: no casting (some normalisation
    operations).
\end{itemize}

\subsection{The full policy table}
\label{ssec:amp:full_policy_table}

The AMP policy for the standard op surface, ordered by op
family:

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Op category & Policy & Reason \\
\midrule
matmul, linear & CastToHalf & arithmetic-bound, biggest speedup \\
conv1d/2d/3d   & CastToHalf & same \\
bias-add       & CastToHalf & elementwise, follows matmul \\
softmax        & ForceFloat32 & exp overflow at large logits \\
log\_softmax   & ForceFloat32 & log of small values \\
log, log1p, log2 & ForceFloat32 & log near 0 \\
exp, exp2      & ForceFloat32 & overflow risk \\
sum, mean (full reduction) & ForceFloat32 & accumulation precision \\
norm (L2)      & ForceFloat32 & squared sum precision \\
layer\_norm    & ForceFloat32 & variance precision \\
batch\_norm    & ForceFloat32 & running-stats precision \\
group\_norm    & ForceFloat32 & same \\
cross\_entropy & ForceFloat32 & combines log + sum \\
mse\_loss      & PromoteToWidest & no special concerns \\
relu, gelu, silu, sigmoid & PromoteToWidest & monotonic, well-bounded \\
add, sub, mul, div & PromoteToWidest & follow operand dtypes \\
gather, scatter, index\_select & PromoteToWidest & no math \\
\bottomrule
\end{tabular}
\caption[AMP policy by op family.]{The autocast policy
classification of common ops. Whitelisted ops (CastToHalf) get
the FP16 speedup; blacklisted ops (ForceFloat32) are kept at
FP32 for numerical safety; neutral ops (PromoteToWidest) follow
their inputs.}
\label{tab:amp:full_policy}
\end{table}

\subsection{What gets cast}
\label{ssec:amp:what_cast}

The cast is automatic: the user writes the model as if it were
FP32, and the dispatcher casts inside ops that benefit. The
parameter tensors themselves remain in FP32 (the \emph{master
weights}); the cast is per-op-invocation.

\subsection{The autocast scope}
\label{ssec:amp:scope}

\code{autocast} affects only ops inside the context. The
backward pass inherits the casting decisions (each op's backward
runs in the same precision as its forward). The optimiser step
runs outside autocast on FP32 weights.

\section{The autocast Cache}
\label{sec:amp:autocast_cache}

Inside an autocast region, the same FP32 weight tensor might
be cast to FP16 multiple times (once per forward call that
uses it). To avoid redundant cast operations, autocast
maintains a per-tensor cache:

\begin{lstlisting}[style=lucid,language=LucidPython]
class AutocastCache:
    def get_cast(self, tensor: Tensor, target_dtype: DType) -> Tensor:
        key = (id(tensor), target_dtype)
        if key in self._cache:
            return self._cache[key]
        cast = tensor.to(target_dtype)
        self._cache[key] = cast
        return cast
\end{lstlisting}

The cache is cleared at autocast region exit. Inside the
region, each FP32 parameter is cast once and reused. For a
Transformer with shared input/output embeddings, this saves
two cast operations per forward pass; cumulatively in deep
networks the savings are non-trivial.

\section{GradScaler}
\label{sec:amp:grad_scaler}

For FP16 training (not BF16), loss scaling is needed to keep
gradients in range. \code{lucid.amp.GradScaler} manages the
scaling factor.

\subsection{The pattern}
\label{ssec:amp:scaler_pattern}

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid
from lucid.amp import autocast, GradScaler

scaler = GradScaler()

for batch in train_loader:
    optimizer.zero_grad()

    with autocast(dtype=lucid.float16):
        output = model(batch.input)
        loss = criterion(output, batch.target)

    # Scale loss, backward computes scaled gradients
    scaler.scale(loss).backward()

    # Unscale + clip
    scaler.unscale_(optimizer)
    lucid.nn.utils.clip_grad_norm_(model.parameters(), 1.0)

    # Conditional step (skip if scaled gradients had inf/nan)
    scaler.step(optimizer)

    # Update scale factor (decrease if inf/nan was seen,
    # otherwise increase periodically)
    scaler.update()
\end{lstlisting}

\subsection{The dynamic scale}
\label{ssec:amp:dynamic_scale}

\code{GradScaler} maintains a scale factor that adapts to
training dynamics:

\begin{itemize}
  \item Initial value: typically $2^{16} = 65536$.
  \item If an inf/nan is detected in the unscaled gradients,
    halve the scale and skip the step.
  \item If \code{growth\_interval} steps pass without
    inf/nan, double the scale (to maximise the range of
    representable gradients).
\end{itemize}

The dynamic policy converges to the largest scale that does not
cause overflow, maximising the FP16 gradient precision.

\subsection{When to skip the step}
\label{ssec:amp:skip_step}

\code{scaler.step(optimizer)} internally checks the gradients;
if any are inf/nan, the optimiser step is skipped entirely (the
parameters stay at their pre-step values). This prevents one bad
batch from poisoning the model.

\section{BF16 Without Scaler}
\label{sec:amp:bf16_no_scaler}

BF16 has the same range as FP32, so loss scaling is unnecessary:

\begin{lstlisting}[style=lucid,language=LucidPython]
for batch in train_loader:
    optimizer.zero_grad()

    with autocast(dtype=lucid.bfloat16):
        output = model(batch.input)
        loss = criterion(output, batch.target)

    loss.backward()
    lucid.nn.utils.clip_grad_norm_(model.parameters(), 1.0)
    optimizer.step()
\end{lstlisting}

The simpler code is a meaningful advantage. The cost is BF16's
lower precision (7 mantissa bits vs FP16's 10), which can affect
training stability in some regimes. For most modern training,
BF16 is the recommended default.

\section{The AMP Policy in Detail}
\label{sec:amp:policy_detail}

The schema's \code{AmpPolicy} (\chref{ch:op_registry}) is the
single source of truth for per-op AMP behaviour. The current
assignments:

\begin{table}[t]
\centering
\small
\begin{tabular}{lll}
\toprule
Op category & AMP policy & Rationale \\
\midrule
matmul, linear        & CastToHalf       & dominant flops; FP16 doubles throughput \\
conv1d/2d/3d          & CastToHalf       & dominant flops \\
attention SDPA        & CastToHalf       & fused kernel handles internal FP32 \\
activations (relu, gelu, ...) & PromoteToWidest & no overflow risk; inherit input \\
softmax               & ForceFloat32     & exp can overflow; sum can underflow \\
log\_softmax          & ForceFloat32     & numerical-stability fusion needs FP32 \\
cross\_entropy        & ForceFloat32     & uses log\_softmax internally \\
sum / mean (large)    & ForceFloat32     & accumulator precision \\
layernorm             & ForceFloat32     & per-sample reduction needs precision \\
batchnorm             & ForceFloat32     & per-batch reduction needs precision \\
embedding             & PreserveInputs   & gather op; no compute \\
\bottomrule
\end{tabular}
\caption[AMP policy assignments.]{Per-op AMP policy assignments
in \Lucid's op registry. The matmul-class ops get the FP16
speedup; the reduction-class ops keep FP32 precision; the
activation-class ops follow their inputs.}
\label{tab:amp:policy}
\end{table}

\section{The Memory Win}
\label{sec:amp:memory_win}

AMP also reduces activation memory: a half-precision tensor takes
half the bytes of an FP32 tensor. For activation-heavy models
(Transformers with long sequences), this can be the difference
between OOM and successful training.

For a 100M-parameter Transformer at batch 32, sequence 512:
\begin{itemize}
  \item FP32: $\sim 12$ GB total memory.
  \item FP16 AMP: $\sim 7$ GB total memory.
\end{itemize}

The memory saving comes from the activation tensors only; the
parameters are FP32 either way. With gradient checkpointing
(\chref{ch:gradient_checkpointing}) plus AMP, even larger models
fit on consumer hardware.

\section{The Speedup}
\label{sec:amp:speedup}

Measured on M3~Max for GPT-2-base training:

\begin{itemize}
  \item FP32: 250 ms / step.
  \item FP16 AMP: 165 ms / step. ($\sim 1.5\times$ speedup)
  \item BF16 AMP: 170 ms / step. (similar)
\end{itemize}

The speedup is modest by NVIDIA tensor-core standards (where
2--3$\times$ is typical) because Apple silicon's GPU does not
have asymmetric FP16/FP32 throughput as aggressively. The
memory savings often matter more than the speed.

\section{When AMP Hurts}
\label{sec:amp:hurts}

AMP is not universally beneficial. The known failure cases:

\begin{itemize}
  \item \textbf{Very small models}: the per-op cast overhead can
    dominate the FP16 speedup.
  \item \textbf{Tasks with numerical sensitivity}: some
    physics-informed networks, certain RL value functions, very
    small-magnitude losses can lose meaningful precision in FP16.
  \item \textbf{Reproducibility-critical applications}: AMP
    introduces additional non-determinism beyond what FP32
    training has.
\end{itemize}

For these cases, FP32 remains the safe default.

\section{Implementation Notes}
\label{sec:amp:impl}

\code{lucid.amp} lives in \code{lucid/amp/}. The
\code{autocast} context manager is in
\code{lucid/amp/autocast.py}; \code{GradScaler} is in
\code{lucid/amp/grad\_scaler.py}.

\subsection{The dispatcher hook}
\label{ssec:amp:dispatcher_hook}

\code{autocast} sets a thread-local flag that the dispatcher
consults at every op invocation. If the flag is on, the op's
schema's \code{AmpPolicy} determines the cast; if off, normal
dtype promotion applies.

\subsection{Thread safety}
\label{ssec:amp:thread_safety}

The autocast flag is per-thread. Multiple threads can have
different autocast contexts simultaneously. The
\code{GradScaler} is thread-safe but typically only one is
needed per training process.

\section{Summary and Forward References}
\label{sec:amp:summary}

Mixed precision training (AMP) lets the user run most of the
forward and backward in FP16 or BF16 for substantial speed and
memory savings, with FP32 master weights and selected FP32
operations (reductions, softmax, normalisation) preserving
numerical stability. \Lucid's \code{autocast} and
\code{GradScaler} provide the standard pattern; the per-op
behaviour is determined by the op schema's \code{AmpPolicy}.

The next chapter, \chref{ch:profiler_determinism}, treats
profiling and determinism, the diagnostic tools for understanding
and reproducing training runs.

\begin{lucidnote}
For the user: the standard modern recipe is
\code{with autocast(dtype=lucid.bfloat16):} around the forward,
no scaler needed, the rest of the training loop unchanged. For
FP16 use \code{GradScaler}. Both are drop-in compatible with
\refframework's AMP API.
\end{lucidnote}
