% !TEX root = ../main.tex
\chapter{Profiler and Determinism}
\label{ch:profiler_determinism}

This chapter closes \partref{part:neural_networks} with two
diagnostic tools that are operationally important but rarely
described in textbooks: the per-op profiler and the deterministic
execution mode. The profiler helps understand where training
time is spent. Determinism ensures that two runs of the same
training script produce the same result --- essential for
debugging, reproducible research, and regression testing.

\section{Why Profile}
\label{sec:profiler_determinism:why_profile}

Training a neural network involves dozens of operations per
forward step, each with different cost characteristics. Without
profiling, the practitioner relies on intuition (``matmul is
expensive, elementwise is cheap'') that is often wrong: a
chain of small ops can dominate a single large matmul; an
unexpected synchronisation can stall the GPU for milliseconds; a
data-loader bottleneck can leave the GPU idle.

The profiler reports per-op time so the practitioner can target
optimisations effectively.

\section{The Lucid Profiler}
\label{sec:profiler_determinism:profiler}

\code{lucid.profiler.Profiler} is a context manager that records
per-op timings:

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid
from lucid.profiler import Profiler

with Profiler() as p:
    for batch in train_loader[:10]:
        loss = criterion(model(batch.input), batch.target)
        loss.backward()
        optimizer.step()
        optimizer.zero_grad()

print(p.summary())
\end{lstlisting}

The output is a table of per-op statistics: count, total time,
self time (excluding sub-op calls), average per call, peak
memory.

\subsection{What it tracks}
\label{ssec:profiler_determinism:tracks}

Every op invocation through the \Lucid\ dispatcher is recorded:
op name, input shapes, input dtypes, wall-clock duration,
backend (CPU/MLX/MPSGraph), and approximate memory delta.

The recording adds a few microseconds per op call. For most
training workloads this is negligible; for very small-op-heavy
loops (small-batch RNN training), the overhead can reach 5--10\%.

\subsection{Aggregation views}
\label{ssec:profiler_determinism:views}

The profiler provides several aggregation modes:

\begin{itemize}
  \item \code{summary()}: per-op-name aggregation. Shows which
    op kinds dominate.
  \item \code{by\_shape()}: per-op-name + per-input-shape. Useful
    for finding small-op pathologies.
  \item \code{by\_module()}: aggregation by source module
    (requires the profiler to be aware of the module call stack;
    on by default).
  \item \code{timeline()}: linearised event log. Useful for
    finding sequential dependencies and synchronisation stalls.
\end{itemize}

\subsection{Saving and analysis}
\label{ssec:profiler_determinism:saving}

\code{p.export\_chrome\_trace('trace.json')} writes a
Chrome-tracing-format file that the user can load in
\code{chrome://tracing} or \code{ui.perfetto.dev} for visual
analysis. The graphical timeline view often makes bottlenecks
obvious.

\section{Profiler vs External Tools}
\label{sec:profiler_determinism:vs_external}

\Lucid's profiler is a per-op accounting tool. For lower-level
profiling (per-kernel GPU time, memory bandwidth utilisation),
Apple's \code{Instruments.app} with the Metal System Trace
template is the right tool. The two are complementary:
\Lucid's profiler tells you which op is expensive; Instruments
tells you why.

\subsection{The Instruments workflow}
\label{ssec:profiler_determinism:instruments}

\begin{enumerate}
  \item Launch Instruments.app and choose the Metal System Trace
    template.
  \item Attach to the Python process running the training script.
  \item Record for a few training steps.
  \item Analyse the timeline: GPU utilisation, command queue
    backlog, per-kernel duration.
\end{enumerate}

Apple's developer documentation has extensive guides for
Instruments. \Lucid\ does not ship with custom Instruments
templates; the standard Metal template is sufficient.

\section{Determinism}
\label{sec:profiler_determinism:determinism}

The same training script run twice on the same hardware should
produce identical numerical results. \emph{Should}, but in
practice does not, because of three sources of non-determinism:

\begin{enumerate}
  \item \textbf{Parallel reduction order}: floating-point
    addition is non-associative; a parallel reduction can produce
    different bit-level results depending on the order in which
    threads combine partial sums.
  \item \textbf{Non-deterministic op implementations}: some ops
    (notably \code{scatter\_add} with duplicate indices) have
    multiple valid implementations that produce slightly
    different results.
  \item \textbf{Random number generators}: training uses RNG for
    initialisation, dropout, data shuffling. Without explicit
    seeding, each run sees different randomness.
\end{enumerate}

\subsection{The set\_deterministic mode}
\label{ssec:profiler_determinism:set_det}

\code{lucid.set\_deterministic(True)} enables a strict mode that
addresses the first two sources:

\begin{itemize}
  \item Parallel reductions use a deterministic ordering (at
    cost to throughput).
  \item Non-deterministic op implementations are replaced by
    deterministic equivalents (typically a CPU fallback for
    \code{scatter\_add} with duplicates).
\end{itemize}

In deterministic mode, an op that does not have a deterministic
implementation raises \code{LucidError} rather than silently
producing non-reproducible results.

\subsection{The RNG side}
\label{ssec:profiler_determinism:rng}

For full determinism, the user must also seed the RNG:

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid

lucid.manual_seed(42)
lucid.set_deterministic(True)

# Now training is reproducible: same script, same hardware,
# same bit-level results across runs.
\end{lstlisting}

\code{manual\_seed} seeds the default RNG; for finer control,
each \code{Generator} (\chref{ch:distributions}) can be seeded
individually.

\section{The Cost of Determinism}
\label{sec:profiler_determinism:cost}

Deterministic mode is slower. The two main contributors:

\begin{itemize}
  \item Reductions use a deterministic tree-reduce algorithm
    instead of the faster non-deterministic shuffle.
  \item Non-deterministic ops fall back to CPU for the
    deterministic implementation; the bridge to/from CPU costs
    time.
\end{itemize}

For a typical Transformer training step on M3~Max:

\begin{itemize}
  \item Default (non-deterministic): 250 ms / step.
  \item Deterministic: 320 ms / step (1.28$\times$ slower).
\end{itemize}

The cost is acceptable for the use cases that need determinism
(debugging, regression testing, reproducible research). For
production training where reproducibility is not required, the
default non-deterministic mode is faster.

\section{Reproducibility Recipes}
\label{sec:profiler_determinism:recipes}

\subsection{Full reproducibility}
\label{ssec:profiler_determinism:full_repro}

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid

lucid.manual_seed(42)
lucid.set_deterministic(True)
# In addition: seed Python's random and (if used) numpy
import random
random.seed(42)
\end{lstlisting}

Plus: turn off any source of operational non-determinism
(asynchronous data loading with shuffling, etc.) or seed those
sources individually.

\subsection{Run-to-run debugging}
\label{ssec:profiler_determinism:debug}

When investigating a training-time bug, deterministic mode lets
the user reproduce the same numerical trajectory across runs.
Print statements added to one run show the exact state of the
next run.

\subsection{Regression tests}
\label{ssec:profiler_determinism:regression}

CI workflows that verify training behaviour rely on
deterministic mode: a small training run produces a fixed
expected loss value, which the test compares against. Without
determinism, the test would have to use a wide tolerance band
that defeats the regression-detection purpose.

\section{Cross-Device Reproducibility}
\label{sec:profiler_determinism:cross_device}

\code{set\_deterministic} guarantees reproducibility within the
same device. Across devices --- CPU vs GPU, M1 vs M3 --- bit-level
reproducibility is not guaranteed, because the kernel
implementations differ.

For loose reproducibility across devices (same numerical
trajectory to within a few ULPs), AMP and reduction order
remain the main confounds; the user should match precision and
disable AMP if cross-device matching matters.

\section{Performance Counters}
\label{sec:profiler_determinism:perf_counters}

Apple silicon's CPU exposes detailed performance counters (cache
miss rates, branch mispredict rates, IPC) via the \code{kperf}
API. \Lucid's profiler does not currently read these counters;
the user wanting cycle-level analysis uses Instruments'
CPU Counters template, which provides a UI for the same
information.

\section{Memory Profiling}
\label{sec:profiler_determinism:memory_profile}

\code{lucid.memory\_stats()} provides snapshot-time memory
information (\chref{ch:allocator}). For continuous monitoring
during a training run, the profiler records peak memory per op,
which the user can aggregate to understand where allocations
spike.

The combination of \code{Profiler} for time and
\code{memory\_stats()} for memory covers most diagnostic needs.

\subsection{Memory snapshot}
\label{ssec:profiler_determinism:memory_snapshot}

For a deeper look at why memory grew at a particular step, the
\code{lucid.cuda\_compat.memory\_snapshot()} call (the name is
a compatibility alias; on Apple silicon this reads MLX's
allocator) returns a list of every live allocation with its
shape, dtype, and \emph{allocation site} (Python file:line where
the tensor was created).

\begin{lstlisting}[style=lucid,language=LucidPython]
snap = lucid.memory_snapshot()
big_tensors = [s for s in snap if s.bytes > 100_000_000]
for s in sorted(big_tensors, key=lambda x: -x.bytes):
    print(f"{s.bytes/1e9:.2f}GB  {s.shape}  {s.dtype}  {s.alloc_site}")
\end{lstlisting}

The snapshot is the standard debugging tool for unexplained
OOMs: walk the list to identify the culprit allocation, trace
back to the code site, fix.

\subsection{The peak-memory profile}
\label{ssec:profiler_determinism:peak_profile}

For a continuous profile of how memory evolves through a
training step:

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.profiler.memory.profile(track_allocations=True) as mp:
    train_step(...)

plt.plot(mp.timeline_ms, mp.allocated_bytes)
plt.xlabel('time (ms)')
plt.ylabel('allocated bytes')
plt.show()
\end{lstlisting}

The resulting timeline shows the memory ``staircase'' --- each
allocation as a step up, each free as a step down. The peak is
the OOM-relevant number; the staircase shape diagnoses whether
the peak is due to a single huge tensor (one big step) or many
medium ones (a long sawtooth).

\section{System-level Profiling}
\label{sec:profiler_determinism:system_profile}

\code{lucid.profiler.system} provides a coarser-grained view
that includes contributors outside \Lucid's op dispatch:

\begin{itemize}
  \item \textbf{Data loading}: time spent in
    \code{DataLoader.\_\_next\_\_}; the worker pool's wait
    time vs the GPU's idle time.
  \item \textbf{Python overhead}: time spent in Python code
    between \Lucid\ calls (custom loss functions, metric
    aggregation, logging).
  \item \textbf{Disk I/O}: time spent in checkpoint save/load,
    log-file writes.
  \item \textbf{GPU utilisation}: fraction of wall-clock time
    where the GPU is busy.
\end{itemize}

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.profiler.system.profile() as sys_p:
    train_epoch(model, loader)

print(sys_p.summary())
# Train epoch system profile:
#   GPU utilisation: 76% (24% idle)
#   Idle time attribution:
#     waiting for data loader: 18% (consider more num_workers)
#     waiting for synchronisation: 4%
#     Python between ops: 2%
#   Forward pass:  840 ms / iter
#   Backward pass: 1240 ms / iter
#   Data loading: 110 ms / iter (overlapped 90%)
\end{lstlisting}

The system profile is the right starting point for
``my training is slower than expected'' investigations:
identify whether the bottleneck is GPU compute, data, or
elsewhere before diving into per-op detail.

\section{Range Markers}
\label{sec:profiler_determinism:ranges}

For grouping op timings under a user-readable label,
\Lucid\ provides nvtx-style range markers:

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.profiler.range("encoder"):
    enc_out = encoder(x)
with lucid.profiler.range("decoder"):
    dec_out = decoder(enc_out)
\end{lstlisting}

The profiler aggregates ops inside each range and reports
range-level totals in addition to per-op totals. Useful for
attributing time to high-level phases (encoder vs decoder,
forward vs backward, model vs data-loading).

Ranges nest: an outer ``epoch'' range can contain an inner
``train\_step'' range, which contains ``forward'' /
``backward'' / ``optim\_step''. The profiler's tree report
shows the nesting hierarchy.

\section{The Profiler's Internals}
\label{sec:profiler_determinism:internals}

For users who need to understand or extend the profiler:

\subsection{The hook point}
\label{ssec:profiler_determinism:hook_point}

Every dispatched op flows through a single function
\code{\_C.ops.dispatch\_op}. The profiler hooks the entry and
exit of this function:

\begin{lstlisting}[style=lucid,language=LucidPython]
def dispatch_op(op_name, inputs, attrs):
    if profiler.is_active():
        evt = profiler.begin_op(op_name, inputs, attrs)
    try:
        result = _do_dispatch(op_name, inputs, attrs)
        if profiler.is_active():
            lucid.runtime.synchronize()  # ensure GPU work has finished
            profiler.end_op(evt, result)
        return result
    except Exception as e:
        if profiler.is_active():
            profiler.error_op(evt, e)
        raise
\end{lstlisting}

The synchronisation inside the hook is necessary for accurate
GPU timing: without it, the recorded duration would be just
the dispatch time (when the GPU work was submitted), not the
actual execution time. The synchronisation has a real cost ---
the profiler trades absolute speed for accurate attribution.

\subsection{The event record}
\label{ssec:profiler_determinism:event_record}

Each event captures:
\begin{itemize}
  \item Op name (string).
  \item Input shapes and dtypes (for grouping).
  \item Start and end timestamps (wall-clock).
  \item GPU command-buffer ID (for cross-referencing with
    Instruments).
  \item Python call stack (if \code{with\_stack=True}).
  \item Memory before/after (if \code{record\_memory=True}).
\end{itemize}

Events are appended to a thread-local list; aggregation happens
at \code{Profiler.\_\_exit\_\_} time. The append is the
hot-path operation and is optimised to be a simple pointer
update.

\subsection{Why thread-local}
\label{ssec:profiler_determinism:thread_local}

\Lucid\ supports multi-threaded clients (e.g., a data-loader
thread and a model-training thread). The profiler records each
thread's events into a separate list to avoid lock contention,
then merges them at aggregation time. The merge is offline
(after the profiled region), so it does not contribute to the
hot-path overhead.

\section{Implementation Notes}
\label{sec:profiler_determinism:impl}

The profiler lives in \code{lucid/profiler/\_\_init\_\_.py}; the
implementation hooks into the dispatcher to record per-op
events. \code{set\_deterministic} is a global flag in
\code{lucid/\_C/core/Determinism.\{h,cpp\}} that the dispatcher
consults at op selection time.

\subsection{Hook overhead}
\label{ssec:profiler_determinism:hook_overhead}

The profiler's hooks add a few microseconds per op invocation
when active. The hooks are no-ops when the profiler context is
not active, so the production training path is unaffected.

\section{Summary and Forward References}
\label{sec:profiler_determinism:summary}

\code{lucid.profiler.Profiler} provides per-op timing and memory
accounting for diagnosing training-time bottlenecks; Apple's
\code{Instruments.app} provides lower-level kernel and
hardware-counter analysis. \code{lucid.set\_deterministic(True)}
plus \code{lucid.manual\_seed()} gives bit-level reproducibility
within the same device, at a modest performance cost.

This chapter closes \partref{part:neural_networks}.
\partref{part:graph_compile} begins the graph-compilation system
that takes the layer library described in this part and produces
optimised execution graphs.

\begin{lucidnote}
For the user: \code{Profiler} is the first tool to reach for when
training is unexpectedly slow; \code{set\_deterministic(True)} is
the first tool when training is unexpectedly non-reproducible.
Both are zero-overhead until activated.
\end{lucidnote}
