% !TEX root = ../main.tex
\chapter{Graph-Capture Design Goals}
\label{ch:compile_design}

\partref{part:graph_compile} treats \Lucid's graph-capture
compiler --- \code{lucid.compile} --- which converts an
imperative model definition into a compiled \MPSGraph\ executable
that runs significantly faster than the eager equivalent. The
compiler is the largest single performance feature in the
\Lucid\ 3.5 release and the most architecturally complex
subsystem in the framework.

This chapter opens the part by stating the design goals, the
non-goals, the locked design decisions, and the high-level
architecture. The chapters that follow elaborate each component:
the tracer (\chref{ch:tracer}), the MpsBuilder
(\chref{ch:mps_builder}), the op emitters
(\chref{ch:op_emitters}), the forward+backward unification
(\chref{ch:fwd_bwd_compile}), the user-facing API
(\chref{ch:compiled_module}), and the hardening work that made
the 3.5 release production-ready (\chref{ch:compile_hardening}).

\section{The Performance Motivation}
\label{sec:compile_design:perf_motivation}

The eager dispatch path (\chref{ch:backend_dispatch}) has
roughly $1.5\,\mu$s of framework overhead per op call. For a
typical training step with several thousand op invocations, the
total framework overhead is a few milliseconds. On models where
individual kernels are also a few milliseconds, the overhead is
a third of total step time --- substantial waste.

The eager path also re-decides per call: each op's dispatch
chooses a backend, validates shapes, builds an autograd node,
saves intermediate tensors. None of these decisions change
across training steps with the same input shape.

Graph capture amortises both costs. The first call traces the
forward pass, identifies the sequence of primitives,
compiles them into a single \MPSGraph\ executable, and caches.
Subsequent calls hit the cached executable and bypass the eager
dispatch entirely.

The measured speedup: $1.4$--$1.8\times$ on Transformer training
workloads, $1.2$--$1.5\times$ on CNN training, on M3~Max. The
exact factor depends on how much of the step time was framework
overhead vs kernel time; smaller models benefit more.

\section{Goals}
\label{sec:compile_design:goals}

The 3.5 compile path was designed with five explicit goals.

\subsection{Goal 1: Transparent acceleration}
\label{ssec:compile_design:goal_transparent}

The user should not have to rewrite their model to benefit from
compilation. The eager-mode forward and the compiled forward
should produce numerically identical results (to within
floating-point tolerance) and accept the same inputs. The
contract is a wrapper:

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid
compiled = lucid.compile(model)
y = compiled(x)  # numerically equivalent to model(x), but faster
\end{lstlisting}

This is the equivalent of \refframework's \code{torch.compile}.
The pattern of opt-in compilation around an unchanged module is
what allows the user to introduce compilation incrementally.

\subsection{Goal 2: Forward and backward together}
\label{ssec:compile_design:goal_fwd_bwd}

The compiled graph should include the backward pass, not just
the forward. The whole training step --- forward, loss, backward
--- should compile as one unit. This is the design that gives the
largest speedup, because intermediate tensors between forward
and backward can be fused away by the compiler.

The alternative --- compile only the forward, run the backward
eagerly --- gives a smaller speedup and was rejected in the early
3.4 prototype.

\subsection{Goal 3: Shape specialisation}
\label{ssec:compile_design:goal_shape}

The compiled graph specialises on the input shapes. A call with
a new shape triggers re-tracing and re-compilation; subsequent
calls with the same shapes hit the cache.

The trade-off: shape specialisation gives the compiler more
information (constant shapes enable more fusion) at the cost of
re-compilation overhead when shapes change. For most training
workloads where the batch shape is fixed, this is the right
trade.

\subsection{Goal 4: Fall-back safety}
\label{ssec:compile_design:goal_fallback}

Not every operation can be compiled (the compiler does not
support every primitive; some have unsupported dtype
combinations; some require dynamic shapes the compiler does not
handle). The compile path should gracefully fall back to eager
execution for unsupported regions, rather than failing
outright.

The default policy is permissive (fallback silently with a
warning); the strict mode (\code{lucid.compile(model, strict=True)})
raises on any unsupported op so the user can diagnose.

\subsection{Goal 5: Debugging visibility}
\label{ssec:compile_design:goal_debug}

A compiled graph is opaque: the user cannot \code{print(tensor)}
intermediate values because they were optimised away. The
compile path should provide diagnostic hooks:
\code{lucid.compile.dump\_trace} prints the trace IR;
\code{lucid.compile.dump\_executable} prints the generated
\MPSGraph\ structure. Both let the user inspect what the
compiler decided.

\section{Non-Goals}
\label{sec:compile_design:non_goals}

Equally important: the things the 3.5 compile path explicitly
does not attempt.

\subsection{Non-goal 1: Dynamic shape support}
\label{ssec:compile_design:non_goal_dyn_shape}

The compile path does not handle dynamic shapes (input shape that
varies per call). For an input shape that changes, the compiler
recompiles.

The user-facing implication: a training loop that uses
\code{drop\_last=True} on the data loader (so every batch has
the same size) compiles once and reuses; a loop that lets the
last batch be smaller recompiles for the last batch every epoch.

Dynamic shape support is on the roadmap for 3.5.1 but deferred
because the parity tests on representative models do not pass
under symbolic primary shapes.

\subsection{Non-goal 2: Whole-program compilation}
\label{ssec:compile_design:non_goal_whole_program}

The compile path compiles per-call. A training script with
multiple distinct compiled regions (an encoder, a decoder, a
loss) compiles each separately. The compiler does not attempt to
fuse across the boundaries.

The trade-off: per-call compilation is simpler and lets the user
incrementally introduce \code{compile()} where it helps.
Whole-program compilation is more complex and rarely beneficial
for typical training workflows.

\subsection{Non-goal 3: Custom op compilation}
\label{ssec:compile_design:non_goal_custom_ops}

Custom \code{Function} subclasses (\chref{ch:custom_functions})
do not compile. The compile path falls back to eager for any
subgraph that contains a custom Function.

The user-side workaround: implement performance-critical custom
ops as primitives (\chref{ch:op_registry}) rather than Functions.
This is the same advice that applies to any framework's compiler.

\subsection{Non-goal 4: Training-loop compilation}
\label{ssec:compile_design:non_goal_loop}

\code{lucid.compile} wraps a single forward (or forward + backward
+ loss) call. It does not wrap the training loop itself
(\code{for batch in loader: ...}). The loop runs in Python; only
the per-batch computation is compiled.

This is the same design choice \refframework's
\code{torch.compile} makes; the loop's overhead is dominated by
the batch compute anyway.

\section{Architecture Sketch}
\label{sec:compile_design:architecture}

The compile path has four major components:

\begin{enumerate}
  \item \textbf{Tracer}: runs the user's model forward in
    shape-only mode, recording the sequence of primitive op
    calls as a trace IR.
  \item \textbf{MpsBuilder}: translates the trace IR into an
    \MPSGraph\ graph, calling the per-op emitters
    (\chref{ch:op_emitters}) to produce the corresponding
    \code{MPSGraphTensor} operations.
  \item \textbf{Executable cache}: caches compiled
    \code{MPSGraphExecutable} objects keyed by the input shape
    tuple. Hits return the cached executable; misses trigger a
    fresh trace and compile.
  \item \textbf{CompiledModule}: the user-facing wrapper that
    holds the cache and dispatches per-call.
\end{enumerate}

\figref{fig:compile_design:flow} sketches the per-call flow.

\begin{figure}[t]
\centering
\begin{adjustbox}{max width=\textwidth, center}
\begin{tikzpicture}[
  node distance=2.5cm and 1.2cm,
  user/.style={draw, rounded corners=2pt, fill=lucid.note.bg,
               font=\small\sffamily, minimum width=3.6cm,
               minimum height=0.8cm, align=center},
  internal/.style={draw, rounded corners=2pt, fill=lucid.info.bg,
                   font=\small\sffamily, minimum width=3.6cm,
                   minimum height=0.8cm, align=center},
  decision/.style={draw, diamond, aspect=2, fill=lucid.code.bg,
                   font=\small\sffamily, align=center,
                   minimum width=3.2cm},
  result/.style={draw, rounded corners=2pt, fill=lucid.ok.bg,
                 font=\small\sffamily, minimum width=3.6cm,
                 minimum height=0.8cm, align=center},
  arrow/.style={->, >=stealth, thick, color=lucid.accent.dark,
                shorten >=2pt, shorten <=2pt},
  lbl/.style={font=\footnotesize\itshape, color=lucid.muted,
              fill=white, inner sep=2pt},
]
  \node[user] (call) at (0, 6) {compiled(x)};
  \node[decision] (cache) at (0, 4.4) {shape in cache?};
  \node[result] (hit) at (-4.5, 2.7) {run cached executable};
  \node[internal] (trace) at (4.5, 2.7) {Tracer};
  \node[internal] (build) at (4.5, 1.2) {MpsBuilder};
  \node[internal] (compile) at (4.5, -0.3) {MPSGraph compile};
  \node[result] (cached) at (4.5, -1.8) {cache + run};

  \draw[arrow] (call) -- (cache);
  \draw[arrow] (cache.west) -- node[lbl, above] {yes} (hit.north);
  \draw[arrow] (cache.east) -- node[lbl, above] {no} (trace.north);
  \draw[arrow] (trace) -- (build);
  \draw[arrow] (build) -- (compile);
  \draw[arrow] (compile) -- (cached);
\end{tikzpicture}
\end{adjustbox}
\caption[Compile path flow.]{Per-call flow of
\code{compiled(x)}. The shape-keyed cache lookup is the first
step. On hit, the cached executable runs directly. On miss, the
tracer + MpsBuilder + MPSGraph compile chain produces a new
executable, which is cached for future calls.}
\label{fig:compile_design:flow}
\end{figure}

\section{The Six Locked Decisions}
\label{sec:compile_design:locked_decisions}

The 3.5 design phase produced six decisions that are now load-
bearing: changing any of them would require redesigning the
compile path. We name them here so the rest of
\partref{part:graph_compile} can reference back.

\subsection{Decision 1: MPSGraph as the IR target}
\label{ssec:compile_design:decision_mpsgraph}

The compile path emits to \MPSGraph, not directly to \Metal.
\MPSGraph\ provides a higher-level IR that \Lucid\ does not have
to maintain.

The alternative would have been a custom IR + a direct Metal
back-end. This would have been substantially more code (a
custom compiler plus per-op Metal kernels), with no clear
performance advantage over \MPSGraph's already-optimised
implementation.

\subsection{Decision 2: Shape specialisation}
\label{ssec:compile_design:decision_shape_spec}

The cache key is the input shape tuple. Different shapes mean
different cache entries.

The alternative (one compiled graph that handles all shapes via
\MPSGraph's symbolic-shape mode) was prototyped and found to
underperform: the symbolic graph cannot do the constant-folding
and shape-specific optimisations that shape-specialised graphs
do.

\subsection{Decision 3: Forward + backward in one graph}
\label{ssec:compile_design:decision_fwd_bwd_single}

The compile path traces both forward and backward into a single
\MPSGraph. The backward is constructed via \MPSGraph's
\code{gradientForPrimaryTensor} method.

The alternative (compile forward only, run backward eagerly) was
prototyped in 3.4 and gave roughly half the speedup of the
combined approach.

\subsection{Decision 4: Fallback on unsupported ops}
\label{ssec:compile_design:decision_fallback}

The compile path silently falls back to eager mode for ops that
\MPSGraph\ does not support (notably: linalg decompositions,
data-dependent output ops, custom Functions).

The alternative (raise on any unsupported op) was rejected as
too brittle. The fallback is what makes \code{compile()} usable
on real-world models without per-op vetting.

\subsection{Decision 5: Cache per CompiledModule, not global}
\label{ssec:compile_design:decision_per_module_cache}

Each \code{CompiledModule} has its own cache. Multiple
\code{lucid.compile(model)} calls produce independent caches.

This trades cache size (each cache holds at most a few entries)
for cache-management simplicity (no global eviction policy
needed).

\subsection{Decision 6: Optional strict mode}
\label{ssec:compile_design:decision_strict}

\code{lucid.compile(model, strict=True)} disables the fallback
and raises on unsupported ops. Used during model development to
catch unsupported patterns; default-off for production training.

\section{What Compilation Achieves}
\label{sec:compile_design:achievements}

The empirical wins for the 3.5 release:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Model & Eager (ms/step) & Compiled (ms/step) \\
\midrule
GPT-2-base (12 layers, seq 1024, batch 8)  & 252 & 142 \\
GPT-2-medium (24 layers, seq 1024, batch 4) & 615 & 348 \\
ViT-B/16 (224x224, batch 32)               & 95 & 64 \\
ResNet-50 (224x224, batch 32)              & 78 & 56 \\
Transformer encoder (12 layers, seq 256, batch 16) & 38 & 22 \\
\bottomrule
\end{tabular}
\caption[Compile speedups.]{Per-step training time on M3~Max for
several models in eager mode vs compiled mode. The compile
speedup ranges from 1.3$\times$ (smaller models where framework
overhead is already low) to 1.8$\times$ (Transformer training
where the savings from inter-op fusion compound).}
\label{tab:compile_design:speedups}
\end{table}

\subsection{First-call vs steady-state}
\label{ssec:compile_design:first_call_cost}

The first call pays the compile cost (typically 0.5--3 seconds
for the models above). The steady-state cost is what's reported
in \tabref{tab:compile_design:speedups}. For training runs of
more than a few hundred steps, the amortised cost is negligible.

\section{Summary and Forward References}
\label{sec:compile_design:summary}

The 3.5 compile path is opt-in graph capture that amortises
per-op framework overhead across many training steps. It targets
\MPSGraph\ as the IR, specialises on input shapes, captures
forward and backward in a single graph, falls back gracefully on
unsupported ops, and gives 1.3--1.8$\times$ speedup on
representative training workloads.

The next chapter, \chref{ch:tracer}, treats the tracer that
converts the user's model into trace IR.
\chref{ch:mps_builder} treats the MPSGraph emission and
executable cache. \chref{ch:op_emitters} catalogues the per-op
emitter table. \chref{ch:fwd_bwd_compile} treats the
forward+backward unification. \chref{ch:compiled_module} treats
the user-facing API. \chref{ch:compile_hardening} treats the
edge cases that the 3.5 release stabilised.

\begin{lucidnote}
For the user: \code{lucid.compile(model)} is the entry point.
The first call is slow; subsequent calls are 30--40\% faster.
For most training workflows, the wrap is one line.
\end{lucidnote}
