% !TEX root = ../main.tex
\chapter{CompiledModule API}
\label{ch:compiled_module}

The user-facing entry point to \Lucid's compile path is
\code{lucid.compile(model)}, which returns a
\code{CompiledModule}. This chapter describes the API surface,
the typical usage patterns, the configuration options, and the
diagnostic interfaces.

\section{Basic Usage}
\label{sec:compiled_module:basic_usage}

\begin{lstlisting}[style=lucid,language=LucidPython]
import lucid

model = MyModel(...)
compiled = lucid.compile(model)

# First call: slow (cold compile).
y = compiled(x)

# Subsequent calls with same input shape: fast.
y = compiled(x)
\end{lstlisting}

The wrapping is one line. The compiled object behaves like the
original module: callable, has the same parameters, the same
\code{state\_dict()}, the same \code{train()} / \code{eval()}
toggles.

\subsection{Inheritance from Module}
\label{ssec:compiled_module:module_inheritance}

\code{CompiledModule} inherits from \code{nn.Module} and
delegates the standard Module interface to the wrapped model.
The user can pass a \code{CompiledModule} anywhere a module is
expected (inside a larger module, to an optimiser, etc.).

\section{Compile Options}
\label{sec:compiled_module:options}

\code{lucid.compile()} accepts several optional arguments:

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled = lucid.compile(
    model,
    mode='default',         # 'default' | 'reduce-overhead' | 'max-autotune'
    fullgraph=False,        # require single compiled graph (no fallback)
    dynamic=False,          # enable dynamic-shape support (3.5.1+)
    strict=False,           # raise on unsupported ops (no fallback)
    backend='mpsgraph',     # 'mpsgraph' (the only option currently)
    cache_size=32,          # max compiled executables to keep
)
\end{lstlisting}

\subsection{The mode argument}
\label{ssec:compiled_module:mode}

\code{mode} chooses a trade-off between compile time and
execution speed:

\begin{itemize}
  \item \code{'default'}: balanced. Compile time roughly 1
    second; runtime 1.3--1.5$\times$ over eager.
  \item \code{'reduce-overhead'}: minimum compile time, less
    aggressive optimisation. Compile time around 300 ms; runtime
    1.2$\times$ over eager. Useful when the model is called for
    only a few hundred steps.
  \item \code{'max-autotune'}: maximum compile time, most
    aggressive optimisation. Compile time up to 10 seconds;
    runtime up to 1.7$\times$ over eager. Useful for very long
    training runs where the amortised compile cost is
    negligible.
\end{itemize}

\subsection{fullgraph}
\label{ssec:compiled_module:fullgraph}

\code{fullgraph=True} disables partial fallback: if any op cannot
be compiled, the entire model fails to compile (raise
\code{LucidError}). Useful during model development to ensure
the user is aware of every compile-time issue.

\subsection{dynamic}
\label{ssec:compiled_module:dynamic}

\code{dynamic=True} enables dynamic shape support (one compiled
graph for variable shapes). Not yet stable in 3.5; default off.
The roadmap targets 3.5.1.

\subsection{strict}
\label{ssec:compiled_module:strict}

\code{strict=True} disables the silent fallback at the op level.
Any unsupported op raises with a diagnostic message naming the
op. Used during development; default off for production.

\section{The Cache}
\label{sec:compiled_module:cache}

The CompiledModule owns an ExecutableCache
(\chref{ch:mps_builder}~Section~\ref{sec:mps_builder:cache}).

\subsection{Cache statistics}
\label{ssec:compiled_module:cache_stats}

\begin{lstlisting}[style=lucid,language=LucidPython]
stats = compiled.cache_stats()
# {'hits': 1500, 'misses': 3, 'size': 3, 'max_size': 32}
\end{lstlisting}

A misses count > 1 over a long training run signals that input
shapes are varying; the user should investigate
(\code{drop\_last=True} on the data loader, etc.).

\subsection{Pre-warming the cache}
\label{ssec:compiled_module:prewarming}

For a workload where the user knows the expected shapes in
advance:

\begin{lstlisting}[style=lucid,language=LucidPython]
# Pre-compile for the two shapes we'll see
for batch_size in [32, 16]:
    dummy = lucid.zeros(batch_size, 1024, 768)
    compiled(dummy)

# Now training does not see compile latency
for batch in train_loader:
    loss = compiled(batch).sum()
    loss.backward()
\end{lstlisting}

\subsection{Cache eviction}
\label{ssec:compiled_module:eviction}

\code{compiled.empty\_cache()} clears the cache. Useful for
freeing memory or for forcing a recompile (e.g., after
parameter shape changes).

\section{Diagnostic Hooks}
\label{sec:compiled_module:diagnostics}

\subsection{dump\_trace}
\label{ssec:compiled_module:dump_trace}

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled.dump_trace(input_shape=(32, 1024))
# Prints the trace IR for the given shape, without compiling
\end{lstlisting}

\subsection{dump\_executable}
\label{ssec:compiled_module:dump_executable}

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled.dump_executable(input_shape=(32, 1024))
# Prints the MPSGraph structure post-compile
\end{lstlisting}

\subsection{explain}
\label{ssec:compiled_module:explain}

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled.explain(input_shape=(32, 1024))
# Prints a summary: number of compiled ops, number of fallback
# ops, fallback reasons, predicted vs measured speedup
\end{lstlisting}

The \code{explain} output is the recommended first diagnostic
for understanding why a given model gets less speedup than
expected.

\section{Save and Load}
\label{sec:compiled_module:save_load}

A \code{CompiledModule}'s state\_dict is the wrapped model's
state\_dict; the compiled executables are not serialised (they
are tied to the specific hardware and macOS version).

\begin{lstlisting}[style=lucid,language=LucidPython]
lucid.save(compiled.state_dict(), 'ckpt.pt')

# Later (on possibly different hardware)
model = MyModel(...)
model.load_state_dict(lucid.load('ckpt.pt'))
compiled = lucid.compile(model)  # re-compile from scratch
\end{lstlisting}

This is the right design because \MPSGraph\ executables are
hardware-tied; serialising them would lock the user to the
compile-time hardware version.

\section{Composing with Autograd and Optimisers}
\label{sec:compiled_module:composing}

The CompiledModule integrates with the rest of the framework:

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled = lucid.compile(model)
optimizer = lucid.optim.AdamW(compiled.parameters(), lr=1e-3)

for batch in train_loader:
    optimizer.zero_grad()
    output = compiled(batch.input)
    loss = criterion(output, batch.target)
    loss.backward()
    nn.utils.clip_grad_norm_(compiled.parameters(), 1.0)
    optimizer.step()
\end{lstlisting}

The optimiser sees the underlying model's parameters; the
gradient flows through the compiled graph; the optimiser step
runs eagerly outside the compile. Standard training loop.

\section{Composing with AMP}
\label{sec:compiled_module:composing_amp}

AMP (\chref{ch:amp}) composes with compile: the compile path
captures the AMP-cast ops correctly.

\begin{lstlisting}[style=lucid,language=LucidPython]
compiled = lucid.compile(model)
scaler = lucid.amp.GradScaler()

for batch in train_loader:
    optimizer.zero_grad()
    with lucid.amp.autocast(dtype=lucid.float16):
        output = compiled(batch.input)
        loss = criterion(output, batch.target)
    scaler.scale(loss).backward()
    scaler.step(optimizer)
    scaler.update()
\end{lstlisting}

The first compile happens under autocast, so the compiled graph
includes the AMP casts. Subsequent calls hit the cache.

A subtlety: switching AMP dtype (FP16 to BF16) invalidates the
cache. The user should pick one and stick with it.

\section{Composing with Gradient Checkpointing}
\label{sec:compiled_module:composing_checkpoint}

Gradient checkpointing (\chref{ch:gradient_checkpointing}) marks
some regions for recomputation. The compile path respects the
mark: a checkpointed region's forward is compiled, but the
backward triggers re-execution rather than reading saved
activations.

The combination (compile + checkpoint) is the
memory-and-speed-optimal recipe for large-model training on
constrained hardware.

\section{Errors and Diagnostics}
\label{sec:compiled_module:errors}

\subsection{Compile-time errors}
\label{ssec:compiled_module:compile_errors}

If the compile fails (unsupported op in \code{strict} mode,
internal MPSGraph error, etc.), \code{CompiledModule} raises
with a diagnostic message naming the failing op and the trace
position:

\begin{verbatim}
LucidError: compile failed at trace node #245
  op: lucid.linalg.cholesky
  reason: no emitter (MPSGraph does not provide LAPACK)
  suggestion: use strict=False to fall back to eager
\end{verbatim}

\subsection{Run-time errors}
\label{ssec:compiled_module:runtime_errors}

A compiled graph that runs into a runtime issue (NaN gradient,
OOM, kernel failure) raises through the standard
\code{LucidError} channel. The user gets a stack trace but not
the compiled-graph context; the workaround is to drop to eager
(\code{model(x)} bypasses the compile) and reproduce the error.

\section{Common Patterns}
\label{sec:compiled_module:patterns}

\subsection{Per-step compile}
\label{ssec:compiled_module:per_step}

The most common pattern: wrap the model, train normally.

\begin{lstlisting}[style=lucid,language=LucidPython]
model = MyModel(...)
compiled = lucid.compile(model)
optimizer = lucid.optim.AdamW(model.parameters(), lr=1e-3)
for batch in loader:
    optimizer.zero_grad()
    loss = loss_fn(compiled(batch.input), batch.target)
    loss.backward()
    optimizer.step()
\end{lstlisting}

\subsection{Whole-step compile}
\label{ssec:compiled_module:whole_step}

For models with non-trivial loss computation, the user can wrap
the entire step:

\begin{lstlisting}[style=lucid,language=LucidPython]
def step(input, target):
    loss = loss_fn(model(input), target)
    return loss

step_compiled = lucid.compile(step)

for batch in loader:
    optimizer.zero_grad()
    loss = step_compiled(batch.input, batch.target)
    loss.backward()
    optimizer.step()
\end{lstlisting}

\subsection{Inference-only compile}
\label{ssec:compiled_module:inference}

For deployment scenarios:

\begin{lstlisting}[style=lucid,language=LucidPython]
model.eval()
for p in model.parameters():
    p.requires_grad = False

compiled = lucid.compile(model)
with lucid.no_grad():
    for batch in eval_loader:
        prediction = compiled(batch.input)
\end{lstlisting}

The no-grad context plus the parameter freezing produces a
forward-only graph (no backward emission), which is faster than
the training-graph form.

\section{Implementation Notes}
\label{sec:compiled_module:impl}

\code{CompiledModule} lives in
\code{lucid/compile/\_compiled\_module.py}. The class wraps a
reference to the original module and an \code{ExecutableCache},
plus the \code{compile()} options.

\subsection{Memory footprint}
\label{ssec:compiled_module:mem_footprint}

A CompiledModule with a few cached executables holds: the
original module (unchanged), the cache (a few hundred KB per
entry for the metadata, plus the \MPSGraph\ executables which
are several MB each). For most training workloads (one or two
cached shapes), the overhead is a few tens of MB.

\section{Summary and Forward References}
\label{sec:compiled_module:summary}

\code{lucid.compile(model)} returns a \code{CompiledModule}
that behaves like the wrapped model but executes through a
compiled \MPSGraph. Options control the compile-time/runtime
trade-off (\code{mode}), the strict-vs-fallback behaviour
(\code{strict}, \code{fullgraph}), and the cache size. Diagnostic
hooks (\code{dump\_trace}, \code{dump\_executable},
\code{explain}) expose the compile process to the user.

The next chapter, \chref{ch:compile_hardening}, treats the
edge cases that the 3.5 release stabilised: dynamic shapes,
unsupported ops, NaN handling under compile, SDK version
sensitivity.

\begin{lucidnote}
For the user: the recipe is \code{compiled = lucid.compile(model)}
plus a normal training loop. The compile cost is paid once on
the first call; subsequent calls are 30--40\% faster. For
debugging, use \code{strict=True} to surface unsupported ops.
\end{lucidnote}
