% !TEX root = ../main.tex
\chapter{Vision: CNN Family}
\label{ch:vision_cnn}

The convolutional vision models are the oldest and largest part
of \Lucid's model zoo. They span thirty years of computer-vision
research: from LeNet (1998) through VGG, ResNet, DenseNet,
Inception, MobileNet, EfficientNet, ConvNeXt. This chapter
catalogues the families, gives the architectural sketch of each,
and notes the design choices that \Lucid's implementation makes.

\section{LeNet}
\label{sec:vision_cnn:lenet}

The original convolutional network, designed for handwritten
digit recognition. Two conv-pool blocks followed by three fully
connected layers. Total around 60K parameters.

\Lucid\ provides \code{lenet} (the original 1998 architecture)
and \code{lenet\_modern} (a sanitised variant with ReLU instead
of tanh and standard max-pooling). The historical interest is
greater than the practical use; both are in the zoo for
completeness.

\section{VGG}
\label{sec:vision_cnn:vgg}

VGG \cite{simonyan_vgg_2014} demonstrated that deep stacks of
small ($3 \times 3$) convolutions could outperform shallow stacks
of larger kernels. The architecture is a uniform tower:

\begin{itemize}
  \item Five stages, each with 2--4 convolutional layers and a
    max-pooling layer.
  \item Three fully connected layers at the top (the FC layers
    hold most of VGG's parameters).
\end{itemize}

\Lucid\ provides \code{vgg\_11}, \code{vgg\_13}, \code{vgg\_16},
\code{vgg\_19} (the four variants from the paper, named by
total layer count).

\subsection{Parameter cost}
\label{ssec:vision_cnn:vgg_params}

VGG-16 has 138M parameters, of which 124M are in the three FC
layers. The FC layers' size is the historical reason for VGG's
dominance giving way to ResNet: ResNet replaces the FC layers
with global average pooling, slashing parameter count.

\section{ResNet}
\label{sec:vision_cnn:resnet}

ResNet \cite{he_resnet_2016} introduced the residual block
\[
y \;=\; F(x) + x,
\]
where $F$ is a stack of conv-BN-ReLU layers. The skip connection
lets gradient flow directly through, addressing the degradation
problem of very deep networks.

\subsection{Variants}
\label{ssec:vision_cnn:resnet_variants}

The paper lists five variants:
\begin{itemize}
  \item \code{resnet\_18}, \code{resnet\_34}: BasicBlock (two
    conv layers per block).
  \item \code{resnet\_50}, \code{resnet\_101}, \code{resnet\_152}:
    Bottleneck (1x1, 3x3, 1x1 with channel expansion).
\end{itemize}

\Lucid\ provides all five.

\subsection{ResNeXt and SE-ResNet}
\label{ssec:vision_cnn:resnext}

\code{resnext\_50}, \code{resnext\_101}: group-convolution
variant that improves parameter efficiency.

\code{se\_resnet\_50}, \code{se\_resnet\_101}: squeeze-and-
excitation channel attention added to each residual block.

\subsection{Pre-activation ResNet}
\label{ssec:vision_cnn:preact}

\code{resnet\_50\_preact}: the v2 variant where BN+ReLU is
applied before convolution rather than after. Slightly improved
training dynamics for very deep networks.

\section{DenseNet}
\label{sec:vision_cnn:densenet}

DenseNet \cite{huang_densenet_2017} extends the skip connection
to all previous layers within each dense block:
\[
x_\ell \;=\; F_\ell([x_0, x_1, \ldots, x_{\ell-1}]),
\]
where $[\cdot]$ is concatenation along the channel axis. Each
layer receives concatenated features from all previous layers.

The trade-off: better gradient flow, fewer parameters per layer
(each layer adds only a small \emph{growth rate} number of
channels), but high activation memory due to the
concatenations.

\Lucid\ provides \code{densenet\_121}, \code{densenet\_169},
\code{densenet\_201}, \code{densenet\_161} (the four variants
from the paper).

\section{Inception and Xception}
\label{sec:vision_cnn:inception}

The Inception family uses parallel multi-scale convolution
branches within each block, concatenated along the channel
axis. The architecture has gone through several iterations:

\begin{itemize}
  \item \code{inception\_v3}: the 2015 GoogLeNet-derived version.
  \item \code{inception\_v4}: refinement.
  \item \code{xception}: depthwise-separable convolution
    replaces the standard convolution.
  \item \code{inception\_resnet\_v2}: combines Inception's
    multi-scale branches with ResNet's skip connections.
\end{itemize}

The Inception family is less commonly used in modern practice;
ResNet variants dominate for most applications.

\section{MobileNet}
\label{sec:vision_cnn:mobilenet}

MobileNet \cite{howard_mobilenet_2017} targets efficient
inference on resource-constrained devices. The key building
block is the depthwise-separable convolution: a depthwise
convolution (one filter per input channel) followed by a
pointwise convolution (1x1 conv that mixes channels).

\subsection{Variants}
\label{ssec:vision_cnn:mobilenet_variants}

\code{mobilenet\_v2}: inverted residuals with linear bottlenecks.

\code{mobilenet\_v3\_small}, \code{mobilenet\_v3\_large}:
neural-architecture-searched variants with squeeze-and-
excitation and hardswish activation.

\section{EfficientNet}
\label{sec:vision_cnn:efficientnet}

EfficientNet \cite{tan_efficientnet_2019} introduces
\emph{compound scaling}: jointly scale depth, width, and
resolution by a single coefficient $\phi$:
\[
\text{depth} = \alpha^\phi, \quad
\text{width} = \beta^\phi, \quad
\text{resolution} = \gamma^\phi,
\]
with $\alpha^2 \beta \gamma^2 \approx 2$ for a 2x flop budget.

The scaled variants \code{efficientnet\_b0}--\code{efficientnet\_b7}
cover a wide range of accuracy/flops trade-offs from
mobile-class to very large.

\subsection{Building block}
\label{ssec:vision_cnn:efficientnet_block}

The block is the \emph{MBConv} (Mobile Inverted Bottleneck):
expand the channels, apply depthwise convolution, apply
squeeze-and-excitation, project back to the bottleneck width.

\section{ConvNeXt}
\label{sec:vision_cnn:convnext}

ConvNeXt \cite{liu_convnext_2022} re-examines the ResNet
architecture with Transformer-era design choices: larger
kernels (7x7 depthwise), inverted bottlenecks, LayerNorm,
GELU. The result matches ViT-B's accuracy at similar parameter
counts while remaining purely convolutional.

\Lucid\ provides \code{convnext\_tiny}, \code{convnext\_small},
\code{convnext\_base}, \code{convnext\_large},
\code{convnext\_xlarge} (the five variants from the paper).

\subsection{The block}
\label{ssec:vision_cnn:convnext_block}

\begin{itemize}
  \item 7x7 depthwise convolution.
  \item LayerNorm.
  \item 1x1 conv expand to 4x channels.
  \item GELU.
  \item 1x1 conv project back.
\end{itemize}

The pattern mirrors the Transformer's FFN block, with the
depthwise convolution replacing the attention. The motivation:
attention is replaceable by a local mixing operation; depthwise
convolution is a good local mixer.

\section{Other CNN Families}
\label{sec:vision_cnn:others}

\Lucid\ also includes:

\begin{itemize}
  \item \code{alexnet}: the 2012 ImageNet winner (historical
    interest).
  \item \code{shufflenet\_v2}: efficient mobile architecture with
    channel shuffling.
  \item \code{regnet}: a family discovered by neural-architecture-
    search that explicitly trades off accuracy and flops.
  \item \code{nfnet}: normalizer-free networks that achieve
    ResNet-comparable accuracy without BatchNorm (useful for
    settings where BN's batch dependence is problematic).
\end{itemize}

\section{Common Implementation Patterns}
\label{sec:vision_cnn:patterns}

Across the CNN families, several patterns recur:

\subsection{The stem}
\label{ssec:vision_cnn:stem}

Every CNN starts with a \emph{stem}: typically a 7x7 conv with
stride 2 + max pooling, halving the spatial resolution twice.
The stem brings the input from $(B, 3, 224, 224)$ to
$(B, 64, 56, 56)$ in the first block.

\subsection{The pyramid}
\label{ssec:vision_cnn:pyramid}

The body of a CNN is a sequence of stages, each at progressively
lower spatial resolution and higher channel count:

\begin{itemize}
  \item Stage 1: $(56, 56)$ at 64 channels.
  \item Stage 2: $(28, 28)$ at 128 channels.
  \item Stage 3: $(14, 14)$ at 256 channels.
  \item Stage 4: $(7, 7)$ at 512 channels.
\end{itemize}

The 4-times-2-stride pyramid is the standard for ImageNet-
sized inputs.

\subsection{The head}
\label{ssec:vision_cnn:head}

The classification head is global average pooling + linear:
\begin{lstlisting}[style=lucid,language=LucidPython]
self.head = nn.Sequential(
    nn.AdaptiveAvgPool2d(1),
    nn.Flatten(),
    nn.Linear(feature_dim, num_classes),
)
\end{lstlisting}

The global average pooling makes the head input-size-
independent: the network handles any input resolution where the
final feature map has at least 1x1 spatial size.

\section{Performance on Apple Silicon}
\label{sec:vision_cnn:perf}

Measured training throughput on M3~Max for selected CNNs at
batch 32, FP32:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Model & Params (M) & Step time (ms) \\
\midrule
\code{resnet\_18}      & 11.7   & 32  \\
\code{resnet\_50}      & 25.6   & 78  \\
\code{resnet\_101}     & 44.5   & 142 \\
\code{convnext\_tiny}  & 29.0   & 95  \\
\code{convnext\_base}  & 88.6   & 215 \\
\code{mobilenet\_v3\_small} & 2.5  & 18  \\
\code{efficientnet\_b0} & 5.3   & 28  \\
\code{efficientnet\_b3} & 12.2  & 65  \\
\code{vgg\_16}          & 138.4 & 215 \\
\code{densenet\_121}    & 8.0   & 72  \\
\bottomrule
\end{tabular}
\caption[CNN training throughput.]{Per-step training time on
M3~Max for selected CNNs at batch 32, ImageNet input size
(224x224), FP32, no compile. The throughput scales roughly with
parameter count, modulo architectural-specific differences
(DenseNet's concatenations slow it more than its parameter
count suggests; MobileNet's depthwise convolutions are fast).}
\label{tab:vision_cnn:perf}
\end{table}

\section{Pretrained Weights}
\label{sec:vision_cnn:pretrained}

All CNN families have pretrained weights on ImageNet-1K.
\code{from\_pretrained} downloads them from the hub:

\begin{lstlisting}[style=lucid,language=LucidPython]
model = lucid.models.AutoModel.from_pretrained("resnet_50")
# Weights ready for inference or fine-tuning
\end{lstlisting}

The hub URLs are in \code{lucid/models/vision/<family>/\_pretrained.py}.

\section{Summary and Forward References}
\label{sec:vision_cnn:summary}

The CNN family in \Lucid's vision zoo spans the historical
arc of convolutional networks: LeNet, VGG, ResNet (and
variants), DenseNet, Inception, MobileNet, EfficientNet,
ConvNeXt, and several others. The implementations follow the
original papers; pretrained ImageNet weights are available
through the hub. The common patterns (stem, pyramid, head;
residual connections; depthwise convolutions) are shared across
families.

The next chapter, \chref{ch:vision_transformers}, treats the
vision transformer family.
\chref{ch:detection_segmentation} treats detection and
segmentation. \chref{ch:pretrained} treats the hub.

\begin{lucidnote}
For the user: \code{AutoModel.from\_pretrained("resnet\_50")} is
the typical entry point. For fine-tuning, freeze the backbone
and train a new head; for from-scratch training, the default
initialisation is appropriate.
\end{lucidnote}
