% !TEX root = ../main.tex
\chapter{Vision Transformers}
\label{ch:vision_transformers}

The vision transformer (ViT) family adapted the Transformer
architecture, originally designed for language, to image
classification. Since the 2020 ViT paper
\cite{dosovitskiy_vit_2021}, vision transformers have become
the dominant alternative to CNNs for image-recognition tasks at
ImageNet scale. \Lucid\ provides the canonical ViT and several
derivatives: Swin Transformer, MaxViT, PVT, CoAtNet, CvT.

\section{The Original ViT}
\label{sec:vision_transformers:vit}

ViT \cite{dosovitskiy_vit_2021} applies a Transformer encoder
to image patches. The processing:

\begin{enumerate}
  \item Split the image into non-overlapping patches (typically
    $16 \times 16$ or $32 \times 32$).
  \item Linearly project each patch to a $D$-dimensional
    embedding.
  \item Prepend a learnable [CLS] token.
  \item Add positional embeddings.
  \item Apply $L$ Transformer encoder blocks.
  \item Project the final [CLS] token's embedding to class
    logits.
\end{enumerate}

\subsection{The patch embedding}
\label{ssec:vision_transformers:patch_embed}

The patch extraction can be implemented as either a strided
convolution or as an einops-style rearrange + linear:

\begin{lstlisting}[style=lucid,language=LucidPython]
# As strided conv (efficient):
self.patch_embed = nn.Conv2d(3, D, kernel_size=P, stride=P)

# As rearrange + linear (equivalent):
from lucid.einops import rearrange
patches = rearrange(image, 'b c (h p1) (w p2) -> b (h w) (p1 p2 c)',
                    p1=P, p2=P)
embedding = self.linear(patches)
\end{lstlisting}

\Lucid\ uses the strided-conv form because it's more efficient
on Apple silicon (the conv kernel is one BNNS call).

\subsection{Variants}
\label{ssec:vision_transformers:vit_variants}

The paper lists three sizes, each at two patch sizes:
\begin{itemize}
  \item \code{vit\_base\_16}, \code{vit\_base\_32}
  \item \code{vit\_large\_16}, \code{vit\_large\_32}
  \item \code{vit\_huge\_14}
\end{itemize}

The dimension after the underscore is the patch size. ViT-Base
has $D = 768$, 12 layers, 12 heads; ViT-Large has $D = 1024$,
24 layers, 16 heads.

\section{Swin Transformer}
\label{sec:vision_transformers:swin}

Swin Transformer \cite{liu_swin_2021} introduces \emph{shifted
windows} to reduce attention's quadratic cost. Instead of
attending to all tokens (cost $O(T^2)$ for $T$ tokens), Swin
partitions the tokens into local windows and attends within each.
The window-partition is shifted between layers to allow
cross-window connections.

\subsection{Architecture}
\label{ssec:vision_transformers:swin_arch}

Swin is a hierarchical architecture (like CNNs):
\begin{itemize}
  \item Stage 1: small patches, high spatial resolution, small
    channels.
  \item Stage 2--4: progressive patch merging (2x downsampling),
    doubling channels.
\end{itemize}

The hierarchy makes Swin a good backbone for dense-prediction
tasks (detection, segmentation) where multi-scale features
matter.

\subsection{Variants}
\label{ssec:vision_transformers:swin_variants}

\code{swin\_tiny}, \code{swin\_small}, \code{swin\_base},
\code{swin\_large}. Each at the standard 224x224 input; some
also support 384x384.

\section{MaxViT}
\label{sec:vision_transformers:maxvit}

MaxViT combines block-local attention with grid-local attention,
together with depthwise convolution stages. The result is a
hybrid that scales to high resolution while preserving long-
range dependencies.

\code{maxvit\_tiny}, \code{maxvit\_small}, \code{maxvit\_base}.

\section{PVT (Pyramid Vision Transformer)}
\label{sec:vision_transformers:pvt}

PVT introduces a pyramid structure similar to Swin's but with
\emph{spatial reduction attention}: at lower resolutions of the
pyramid, the keys and values are reduced by a learned
projection, capping attention's cost.

\code{pvt\_tiny}, \code{pvt\_small}, \code{pvt\_medium},
\code{pvt\_large}.

\section{CoAtNet}
\label{sec:vision_transformers:coatnet}

CoAtNet hybridises convolution and attention in a multi-stage
architecture: early stages use depthwise convolution (good for
locality and translation equivariance), late stages use
attention (good for long-range modelling).

\code{coatnet\_0}, \code{coatnet\_1}, ..., \code{coatnet\_4}
(five variants at increasing capacity).

\section{CvT (Convolutional Vision Transformer)}
\label{sec:vision_transformers:cvt}

CvT augments the standard Transformer block with convolutional
projections in the query/key/value computation. The convolutions
add inductive bias for locality while preserving the
Transformer's flexibility.

\code{cvt\_13}, \code{cvt\_21} (named by total layer count).

\section{Common ViT Implementation Details}
\label{sec:vision_transformers:common_details}

\subsection{Position embeddings}
\label{ssec:vision_transformers:pos_embed}

ViT uses learnable absolute position embeddings: a $(1, T, D)$
tensor added to the patch embeddings.

Swin uses relative position bias inside attention:
\[
\mathrm{Attention}(Q, K, V) = \mathrm{softmax}(QK^\top/\sqrt{d} + B) V,
\]
where $B$ is a learnable bias indexed by relative position.

CoAtNet uses both: absolute embeddings at the patch level,
relative bias inside attention.

\subsection{The CLS token}
\label{ssec:vision_transformers:cls_token}

ViT prepends a learnable [CLS] token to the patch sequence. The
[CLS] token's final embedding is the input to the classification
head.

Swin omits the [CLS] token; instead, the classification head
uses global average pooling over all tokens. This makes Swin
more suitable for dense-prediction tasks.

\subsection{Dropout regimes}
\label{ssec:vision_transformers:dropout}

ViT-style training uses several dropout-like regularisations:
\begin{itemize}
  \item Standard dropout in the FFN.
  \item Attention dropout on the attention weights.
  \item DropPath (stochastic depth): randomly skip an entire
    residual block.
\end{itemize}

DropPath is the strongest of the three; the standard rate is
$0.1$--$0.4$ scaled linearly across layers.

\section{Performance on Apple Silicon}
\label{sec:vision_transformers:perf}

Measured training throughput on M3~Max for selected ViTs at
batch 32, FP32, 224x224 input:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Model & Params (M) & Step time (ms) \\
\midrule
\code{vit\_base\_16}    & 86.4   & 125 \\
\code{vit\_large\_16}   & 304.3  & 410 \\
\code{swin\_tiny}       & 28.3   & 88  \\
\code{swin\_base}       & 87.8   & 215 \\
\code{maxvit\_tiny}     & 30.9   & 110 \\
\code{coatnet\_0}       & 25.0   & 95  \\
\bottomrule
\end{tabular}
\caption[ViT training throughput.]{Per-step training time on
M3~Max for selected vision transformers. ViT-Base is comparable
to ResNet-50 in step time despite having 3$\times$ the
parameters; the attention dominates at higher batch sizes.}
\label{tab:vision_transformers:perf}
\end{table}

\section{Compile-Time Wins}
\label{sec:vision_transformers:compile_wins}

Vision transformers benefit substantially from
\code{lucid.compile} (\chref{ch:compile_design}), more than
CNNs do, because attention's overhead per call is high:

\begin{itemize}
  \item \code{vit\_base\_16}: 125 ms eager $\to$ 80 ms compiled
    (1.56$\times$).
  \item \code{swin\_tiny}: 88 ms $\to$ 58 ms (1.52$\times$).
  \item \code{convnext\_tiny}: 95 ms $\to$ 72 ms (1.32$\times$,
    for comparison).
\end{itemize}

The transformer's many normalisations (LayerNorm per block) plus
the SDPA fused kernel are the primary contributors.

\section{Pretrained Weights}
\label{sec:vision_transformers:pretrained}

All ViT variants have pretrained ImageNet-1K and (for the larger
sizes) ImageNet-21K weights. \code{from\_pretrained} downloads
the standard weights:

\begin{lstlisting}[style=lucid,language=LucidPython]
model = lucid.models.AutoModel.from_pretrained(
    "vit_base_16", pretrained_on="imagenet21k"
)
\end{lstlisting}

\section{Choosing Between ViT and CNN}
\label{sec:vision_transformers:vit_vs_cnn}

A guide:

\begin{itemize}
  \item \textbf{Pure classification on standard sizes}: either
    works. ViT-Base is comparable to ResNet-50.
  \item \textbf{Dense prediction (detection, segmentation)}:
    Swin or ConvNeXt as backbone. Pure ViT lacks the
    multi-scale pyramid.
  \item \textbf{Very limited compute budget}: MobileNet or
    EfficientNet (CNN-side) or MobileViT (transformer-side).
  \item \textbf{Very large compute budget (research-scale)}:
    ViT-Huge or larger; the Transformer scales better with
    data and compute than CNNs.
\end{itemize}

\section{Summary and Forward References}
\label{sec:vision_transformers:summary}

The vision transformer family in \Lucid's zoo: ViT (the
original), Swin (shifted-window hierarchy), MaxViT (block + grid
attention), PVT (spatial-reduction pyramid), CoAtNet (hybrid),
CvT (convolutional projection). Each follows the standard zoo
family contract; pretrained weights are available through the
hub.

The next chapter, \chref{ch:detection_segmentation}, treats
object detection and segmentation models, several of which use
ViT or Swin as backbone. \chref{ch:language_models} treats
language models (the original home of the Transformer).

\begin{lucidnote}
For the user: \code{vit\_base\_16} is the default ViT entry
point. For dense-prediction backbones, prefer Swin or ConvNeXt.
All ViT variants benefit substantially from compile.
\end{lucidnote}
