% !TEX root = ../main.tex
\chapter{Generative Models}
\label{ch:generative}

Generative models learn the distribution $p(x)$ from samples,
allowing the user to sample new $x$, score the likelihood of
new $x$, or condition on partial observations. \Lucid\ provides
the four major families: variational autoencoders (VAE)
\cite{kingma_vae_2014}, generative adversarial networks (GAN)
\cite{goodfellow_gan_2014, karras_stylegan_2019}, normalizing
flows \cite{rezende_nf_2015}, and diffusion models
\cite{ho_ddpm_2020, song_ncsn_2019, nichol_iddpm_2021,
song_score_sde_2021}. This chapter treats each family's
mathematical formulation, the \Lucid\ implementation, the
training stability issues, and the Apple-silicon performance
profile.

\section{Variational Autoencoder (VAE)}
\label{sec:generative:vae}

VAE \cite{kingma_vae_2014} models $p(x)$ by introducing a
latent variable $z$:
\[
  p(x) \;=\; \int p(x \mid z) p(z) \, dz.
\]
The encoder $q_\phi(z \mid x)$ approximates the posterior
$p(z \mid x)$; the decoder $p_\theta(x \mid z)$ models the
likelihood. Both $q$ and $p$ are typically Gaussian (the
encoder outputs $\mu, \sigma$; the decoder outputs $\hat{x}$
with a fixed noise scale or learned scale).

\subsection{The ELBO}
\label{ssec:generative:elbo}

Training minimises the negative \emph{Evidence Lower BOund}:
\begin{align}
  -\mathrm{ELBO} \;=\;& -\mathbb{E}_{z \sim q_\phi(\cdot \mid x)}[\log p_\theta(x \mid z)] \notag \\
  &+ D_{\mathrm{KL}}(q_\phi(z \mid x) \,\|\, p(z)).
\end{align}
The first term is the reconstruction loss; the second is the
regularisation that pushes the encoder distribution toward the
prior $p(z) = \mathcal{N}(0, I)$.

For Gaussian encoder and standard-normal prior, the KL has the
closed form:
\[
  D_{\mathrm{KL}}(\mathcal{N}(\mu, \sigma^2 I) \,\|\, \mathcal{N}(0, I))
  \;=\;
  \tfrac{1}{2} \sum_i \bigl(\mu_i^2 + \sigma_i^2 - \log \sigma_i^2 - 1\bigr).
\]

\subsection{The reparameterisation trick}
\label{ssec:generative:reparam}

Sampling $z \sim \mathcal{N}(\mu, \sigma^2)$ is not
differentiable in $\mu, \sigma$ via standard sampling. The trick:
\[
  z = \mu + \sigma \cdot \epsilon, \quad \epsilon \sim \mathcal{N}(0, I).
\]
Now $z$ is a differentiable function of $\mu, \sigma$, and
$\epsilon$ has no parameters. The gradient flows through $\mu$
and $\sigma$.

\subsection{Implementation}
\label{ssec:generative:vae_impl}

\begin{lstlisting}[style=lucid,language=LucidPython]
class VAE(nn.Module):
    def __init__(self, in_dim, hidden_dim, latent_dim):
        super().__init__()
        self.encoder = nn.Sequential(
            nn.Linear(in_dim, hidden_dim), nn.ReLU(),
        )
        self.mu_head    = nn.Linear(hidden_dim, latent_dim)
        self.logvar_head = nn.Linear(hidden_dim, latent_dim)
        self.decoder = nn.Sequential(
            nn.Linear(latent_dim, hidden_dim), nn.ReLU(),
            nn.Linear(hidden_dim, in_dim),
        )

    def forward(self, x):
        h = self.encoder(x)
        mu, logvar = self.mu_head(h), self.logvar_head(h)
        sigma = (0.5 * logvar).exp()
        z = mu + sigma * lucid.randn_like(sigma)
        return self.decoder(z), mu, logvar
\end{lstlisting}

\Lucid\ provides \code{vae} (the standard MLP VAE) and
\code{conv\_vae} (with a convolutional encoder/decoder for
images).

\subsection{VQ-VAE}
\label{ssec:generative:vqvae}

The VQ-VAE \cite{vandenoord_vqvae_2017} replaces the
continuous latent with a discrete codebook of $K$ vectors. The
encoder output is snapped to its nearest codebook vector; the
discrete index is stored as the latent code. The codebook is
trained by a moving-average commitment loss.

\Lucid\ provides \code{vqvae} as the building block for
diffusion-on-latent models (see Section
\ref{sec:generative:diffusion}).

\section{Generative Adversarial Networks (GAN)}
\label{sec:generative:gan}

GAN \cite{goodfellow_gan_2014} learns $p(x)$ via a two-player
zero-sum game between a generator $G$ (maps noise $z$ to
samples) and a discriminator $D$ (distinguishes real from
generated). The standard objective:
\[
  \min_G \max_D \mathbb{E}_{x \sim p_{\text{data}}}[\log D(x)] + \mathbb{E}_{z \sim p_z}[\log(1 - D(G(z)))].
\]
At the optimum, $G$ exactly matches $p_{\text{data}}$ and $D$
outputs $1/2$ everywhere.

\subsection{Training instability}
\label{ssec:generative:gan_instability}

GANs are notoriously hard to train. The pathologies:
\begin{itemize}
  \item \textbf{Mode collapse}: the generator emits only a
    narrow set of outputs, ignoring the data's diversity.
  \item \textbf{Discriminator dominance}: $D$ becomes too good
    too fast; $G$'s gradient vanishes (the $\log(1 - D(G(z)))$
    saturates at 0 when $D(G(z)) \approx 0$).
  \item \textbf{Cycling}: the equilibrium is a saddle point,
    and gradient descent does not converge to saddles in
    general.
\end{itemize}

Mitigations:
\begin{itemize}
  \item Non-saturating loss for $G$:
    $\mathcal{L}_G = -\mathbb{E}[\log D(G(z))]$ instead of
    $\mathcal{L}_G = \mathbb{E}[\log(1 - D(G(z)))]$.
  \item Spectral normalisation on $D$
    \cite{miyato_specnorm_2018}: bounds the Lipschitz constant.
  \item Two-time-scale update rule (TTUR): different learning
    rates for $G$ and $D$.
  \item Gradient penalty (WGAN-GP).
\end{itemize}

\Lucid\ ships these as building blocks in
\code{lucid.models.gan.training\_utils}.

\subsection{Variants}
\label{ssec:generative:gan_variants}

\code{dcgan}: deep convolutional GAN.
\code{stylegan2}: StyleGAN2 \cite{karras_stylegan_2019} with
modulated convolutions (one of the highest-fidelity image
generators of its generation).
\code{wgan\_gp}: Wasserstein GAN with gradient penalty.

\section{Normalizing Flows}
\label{sec:generative:flows}

A normalizing flow \cite{rezende_nf_2015} parameterises $p(x)$
via an invertible mapping $f: \mathbb{R}^d \to \mathbb{R}^d$ from
a base distribution (typically $\mathcal{N}(0, I)$):
\[
  p(x) \;=\; p_z(f^{-1}(x)) \cdot \left| \det \frac{\partial f^{-1}}{\partial x} \right|.
\]
Training maximises the log-likelihood directly. The challenge:
$f$ must be invertible and have a tractable Jacobian
determinant.

\subsection{Coupling layers}
\label{ssec:generative:coupling}

The standard solution (RealNVP, Glow): split the input into two
halves $x = (x_1, x_2)$ and define:
\[
  y_1 = x_1, \qquad
  y_2 = x_2 \odot \exp(s(x_1)) + t(x_1).
\]
The Jacobian is lower-triangular with $\exp(s(x_1))$ on the
diagonal of the bottom block, so the determinant is just
$\exp(\sum s(x_1))$. The inverse is also closed-form:
$x_2 = (y_2 - t(y_1)) \odot \exp(-s(y_1))$.

Stacking many coupling layers (with permutations between them
so different dimensions take turns being the conditioning half)
yields a flexible invertible transformation.

\subsection{Lucid implementation}
\label{ssec:generative:flows_impl}

\code{lucid.models.flow.RealNVP} and
\code{lucid.models.flow.Glow} are provided. The shared building
block is \code{CouplingLayer}; the user composes them into a
flow chain.

\section{Diffusion Models}
\label{sec:generative:diffusion}

Diffusion models \cite{ho_ddpm_2020} learn $p(x)$ by reversing
a noise-injection process. The forward process gradually corrupts
data with Gaussian noise:
\[
  q(x_t \mid x_{t-1}) \;=\; \mathcal{N}(\sqrt{1 - \beta_t} x_{t-1}, \beta_t I),
\]
for $t = 1, \ldots, T$ with $\beta_t$ a small noise schedule.
After $T$ steps, $x_T$ is approximately pure noise.

The reverse process is learned: a neural network $\epsilon_\theta(x_t, t)$
predicts the noise added at step $t$. Sampling from the model
proceeds by starting at $x_T \sim \mathcal{N}(0, I)$ and
iteratively denoising for $t = T, T-1, \ldots, 1$.

\subsection{Training objective}
\label{ssec:generative:diffusion_loss}

The closed-form $q(x_t \mid x_0)$:
\[
  q(x_t \mid x_0) \;=\; \mathcal{N}(\sqrt{\bar{\alpha}_t} x_0, (1 - \bar{\alpha}_t) I), \quad
  \bar{\alpha}_t = \prod_{s=1}^t (1 - \beta_s).
\]
A noisy sample is obtained by
$x_t = \sqrt{\bar{\alpha}_t} x_0 + \sqrt{1 - \bar{\alpha}_t}\,\epsilon$ with
$\epsilon \sim \mathcal{N}(0, I)$. The simple denoising loss:
\[
  \mathcal{L}_{\text{simple}} \;=\;
  \mathbb{E}_{x_0, t, \epsilon}
  \bigl[
    \|\epsilon - \epsilon_\theta(\sqrt{\bar{\alpha}_t} x_0 + \sqrt{1 - \bar{\alpha}_t}\,\epsilon, t)\|^2
  \bigr].
\]
This is the standard DDPM \cite{ho_ddpm_2020} objective: predict
the noise.

\subsection{Sampling}
\label{ssec:generative:diffusion_sampling}

Standard DDPM sampling iterates:
\[
  x_{t-1} \;=\; \frac{1}{\sqrt{1 - \beta_t}}\bigl(x_t - \frac{\beta_t}{\sqrt{1 - \bar{\alpha}_t}} \epsilon_\theta(x_t, t)\bigr) + \sqrt{\beta_t} z,
\]
where $z \sim \mathcal{N}(0, I)$ for $t > 1$ and $z = 0$ for
$t = 1$. This requires $T$ steps (typically 1000), making
sampling slow.

Faster samplers (DDIM, DPM-Solver) reduce the step count to
20--50 by treating the reverse process as a deterministic ODE
\cite{song_score_sde_2021}. \Lucid\ provides:

\begin{lstlisting}[style=lucid,language=LucidPython]
from lucid.models.diffusion import DDPMScheduler, DDIMScheduler

ddpm = DDPMScheduler(num_train_timesteps=1000, beta_schedule="linear")
# ddpm.add_noise(x0, t, eps) -> noisy x_t
# ddpm.step(eps_pred, t, xt) -> x_{t-1}

ddim = DDIMScheduler(num_inference_steps=50)
# Same API, but uses the deterministic skip schedule
\end{lstlisting}

\subsection{Latent diffusion}
\label{ssec:generative:latent_diffusion}

Latent diffusion \cite{rombach_ldm_2022} (the architecture
behind Stable Diffusion) runs the diffusion process not on
pixels but on a VQ-VAE latent (4--8$\times$ downsampled).
The training compute is correspondingly reduced.

\Lucid's \code{lucid.models.latent\_diffusion} wraps a
pretrained VQ-VAE with a UNet-style diffusion model.

\subsection{Improved DDPM}
\label{ssec:generative:iddpm}

Improved DDPM \cite{nichol_iddpm_2021} adds:
\begin{itemize}
  \item Cosine $\beta$ schedule instead of linear (better
    SNR coverage).
  \item Learned reverse variance (rather than fixed $\beta_t$).
  \item Importance-sampled loss for the noise schedule.
\end{itemize}

\Lucid\ ships these as kwargs to the standard scheduler.

\section{The U-Net Backbone}
\label{sec:generative:unet_diffusion}

Most diffusion models use a U-Net (Section
\ref{sec:detection_segmentation:unet}) as $\epsilon_\theta$.
The architecture is adapted with:
\begin{itemize}
  \item Time-step conditioning: a sinusoidal embedding of $t$
    fed to each block via a learned linear projection.
  \item Self-attention at the bottleneck (and sometimes at
    other low-resolution stages).
  \item Class or text conditioning (cross-attention to text
    embeddings, for text-to-image models).
\end{itemize}

\Lucid's \code{lucid.models.diffusion.UNet2DConditionModel}
provides this; the API matches that of
\texttt{diffusers}-style reference implementations.

\section{Performance on Apple Silicon}
\label{sec:generative:perf}

Per-step training time on M3~Max for selected generative models:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Model & Params (M) & Step time (ms) \\
\midrule
\code{vae} (MNIST 28x28)                & 0.5   & 8   \\
\code{conv\_vae} (CIFAR 32x32)          & 4.2   & 22  \\
\code{dcgan} (CIFAR 32x32)              & 6.0   & 35  \\
\code{stylegan2} (FFHQ 256x256)         & 35.0  & 280 \\
\code{realnvp} (CIFAR 32x32)            & 12.0  & 105 \\
\code{ddpm\_unet} (CIFAR 32x32)         & 35.7  & 95  \\
\code{ldm\_unet} (latent 64x64)         & 86.0  & 245 \\
\bottomrule
\end{tabular}
\caption[Generative-model training throughput.]{Per-step
training time on M3~Max for selected generative models. Batch
sizes scaled per model (smaller resolutions get larger
batches). DDPM training is FP32 only on Apple silicon; AMP is
enabled for the larger LDM.}
\label{tab:generative:perf}
\end{table}

\section{Implementation Patterns}
\label{sec:generative:patterns}

\subsection{The Sampler interface}
\label{ssec:generative:sampler_iface}

All generative models expose a uniform sampling API:

\begin{lstlisting}[style=lucid,language=LucidPython]
samples = model.sample(num_samples=64, seed=42)
\end{lstlisting}

For conditional models (class- or text-conditioned), the
condition is passed:

\begin{lstlisting}[style=lucid,language=LucidPython]
samples = model.sample(num_samples=64, condition=class_id_tensor)
\end{lstlisting}

\subsection{The Trainer interface}
\label{ssec:generative:trainer_iface}

Diffusion-specific training utilities (\code{add\_noise},
\code{step}) are encapsulated in the scheduler. The user's
training loop is then standard:

\begin{lstlisting}[style=lucid,language=LucidPython]
for x0 in dataloader:
    t = lucid.randint(0, num_train_steps, (x0.size(0),))
    noise = lucid.randn_like(x0)
    xt = scheduler.add_noise(x0, noise, t)
    pred = model(xt, t)
    loss = lucid.nn.functional.mse_loss(pred, noise)
    loss.backward()
    optimizer.step()
\end{lstlisting}

\section{Pretrained Weights}
\label{sec:generative:pretrained}

\Lucid\ hosts pretrained checkpoints for the smaller generative
models (DCGAN, RealNVP, VAE on standard datasets). For
production-scale image generation (StyleGAN2-FFHQ, Stable
Diffusion), the weights are large (1--7 GB) and the user
downloads them from the original sources via:

\begin{lstlisting}[style=lucid,language=LucidPython]
sd = lucid.models.AutoModel.from_pretrained(
    "stable_diffusion_v1_5",
    pretrained_on="laion",   # via HuggingFace hub
)
\end{lstlisting}

\section{Summary and Forward References}
\label{sec:generative:summary}

\Lucid's generative model zoo spans the four families: VAE
(MLP, conv, VQ), GAN (DCGAN, StyleGAN2, WGAN-GP), normalizing
flows (RealNVP, Glow), and diffusion (DDPM, DDIM, LDM). The
implementations follow the standard zoo family contract; for
diffusion specifically the scheduler is decoupled from the
model so the same denoiser network can be sampled with different
algorithms.

This chapter closes \partref{part:model_zoo}. The next chapter,
\chref{ch:pretrained}, treats the pretrained-weights hub
mechanics that all of these models depend on.

\begin{lucidnote}
For the user: for image generation, the recommended entry
point is \code{ddpm\_unet} for from-scratch training on small
datasets, or \code{ldm\_unet} for high-resolution work. For
research on training dynamics, \code{vae} is the simplest
laboratory.
\end{lucidnote}
