% !TEX root = ../main.tex
\chapter{Stabilization Chain (3.0.0--3.2.2)}
\label{ch:stabilization}

The 3.0--3.2 release chain was the multi-month engineering work
that took \Lucid\ from ``works correctly'' to ``works correctly
\emph{and} fast enough to use''. This chapter is the retrospective
of that work: what was slow at 3.0.0, what changed across the
3.0.x, 3.1.x, 3.2.x patches, what the speed-ups cost in
complexity, and what is still to be done. The chapter is a
case study in incremental performance engineering --- the kind
that does not produce one big-headline-win but rather a chain
of small disciplined improvements.

\section{Starting Point: Lucid 3.0.0}
\label{sec:stabilization:starting_point}

At the 3.0.0 release, \Lucid\ was functionally complete (every
op had an implementation; the model zoo's tests passed) but
slow:

\begin{itemize}
  \item ResNet-50 train iteration: 240\,ms (vs the reference
    framework's 95\,ms on the same hardware). \textbf{2.5$\times$
    slower.}
  \item GPT-2 small train iteration: 380\,ms (vs 165\,ms).
    \textbf{2.3$\times$ slower.}
  \item BERT-base inference: 105\,ms (vs 70\,ms).
    \textbf{1.5$\times$ slower.}
\end{itemize}

The 3.0.0 release notes acknowledged the gap and committed to
closing it. The goal: median within 20\% of the reference
framework across the full model zoo.

\section{Phase 1: Profiling and Triage (3.0.0 $\to$ 3.0.5)}
\label{sec:stabilization:phase_1}

Before any change, we profiled the 3.0.0 implementation on the
target workloads. The headline findings:

\subsection{Allocator thrash}
\label{ssec:stabilization:allocator_thrash}

The \code{lucid.profiler.memory} report on a ResNet-50 iteration
showed 8,400 allocator calls per iteration. Each call took
$\sim$15\,\textmu s, totalling 126\,ms --- over half the
iteration time was spent in the allocator.

Root cause: the 3.0.0 allocator was a thin wrapper around
\code{mx.alloc}; every intermediate tensor went through a real
allocation. There was no cache.

\subsection{Per-op dispatch overhead}
\label{ssec:stabilization:dispatch_overhead}

\code{cProfile} showed 31\% of CPU time in
\code{lucid.\_C.ops.dispatch\_op}. Each op call paid:
\begin{itemize}
  \item Python-side: argument validation, dtype promotion,
    contiguity check.
  \item C++-side: registry lookup, kernel selection, MLX
    handle construction.
\end{itemize}

For a forward pass of GPT-2-small (containing 4,200 ops), this
multiplied: $4200 \times 12\,\mathrm{\mu s} = 50\,\mathrm{ms}$
per pass just in dispatch.

\subsection{Synchronisation barriers}
\label{ssec:stabilization:sync_barriers}

Several ops called \code{synchronize()} unnecessarily,
collapsing all the pending async GPU work and serialising
subsequent dispatches. The biggest culprits were
\code{tensor.item()} (called by \code{torchvision}-style data
loaders for label conversion) and \code{tensor.to(\textnormal{`cpu'})}.

\section{Phase 2: Allocator Rework (3.0.6 $\to$ 3.0.9)}
\label{sec:stabilization:phase_2}

The 3.0.6 release introduced a free-list allocator: tensors
freed are not actually returned to MLX/Metal but kept on a
shape-keyed free list, ready for reuse on the next allocation
of the same shape.

\subsection{Hit rates}
\label{ssec:stabilization:hit_rates}

Post-rework, the allocator hit rate climbed:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Workload & 3.0.5 hit rate & 3.0.9 hit rate \\
\midrule
ResNet-50 train  & 0\%   & 94\% \\
GPT-2 small      & 0\%   & 96\% \\
BERT-base infer  & 0\%   & 91\% \\
\bottomrule
\end{tabular}
\caption[Allocator hit rates after the free-list rework.]{The
allocator hit rate jumped from 0\% (no cache) to 90\%+ after
the free-list addition. The remaining misses are first-iteration
allocations and dynamic-shape allocations (variable batch or
sequence length).}
\label{tab:stabilization:hit_rates}
\end{table}

\subsection{Resulting speed-ups}
\label{ssec:stabilization:phase_2_speedups}

\begin{itemize}
  \item ResNet-50 iter: 240 $\to$ 165\,ms ($-31\%$).
  \item GPT-2 iter: 380 $\to$ 280\,ms ($-26\%$).
  \item BERT-base infer: 105 $\to$ 85\,ms ($-19\%$).
\end{itemize}

The allocator rework alone closed roughly a third of the gap to
the reference framework. The cost: complexity in the allocator's
eviction logic (when to actually release cached memory to
MLX/Metal under memory pressure).

\section{Phase 3: Dispatch Path Inlining (3.1.0 $\to$ 3.1.4)}
\label{sec:stabilization:phase_3}

The next attack was the per-op dispatch overhead. The 3.1
release line refactored the dispatch path:

\subsection{Hot-path inlining}
\label{ssec:stabilization:hot_path_inlining}

The common-case op call (matching shapes, FP32, contiguous,
default backend) bypassed the generic dispatch and went through
a fast specialised path:

\begin{lstlisting}[style=lucid,language=LucidPython]
# Before (3.0.x): every call went through dispatch_op
# After (3.1.x):
def add(a, b):
    if a.dtype == b.dtype and a.shape == b.shape and a.is_contiguous():
        return _C.ops.add_fast(a, b)
    return _dispatch_full(a, b, op="add")
\end{lstlisting}

The fast path skipped:
\begin{itemize}
  \item Dtype promotion (no promotion needed).
  \item Broadcasting analysis (shapes match).
  \item Contiguity check (already contiguous).
  \item Registry lookup (direct call).
\end{itemize}

The slow path is still correct; it just runs less often.

\subsection{C++-side caching}
\label{ssec:stabilization:cpp_caching}

On the C++ side, the MLX-handle construction (creating the
\code{mx::array} from \Lucid's tensor) was cached per
\code{(data\_ptr, dtype, shape, strides)} tuple. Subsequent
calls with the same tensor reuse the cached handle. The cache
is per-thread to avoid locking.

\subsection{Speed-ups}
\label{ssec:stabilization:phase_3_speedups}

\begin{itemize}
  \item ResNet-50 iter: 165 $\to$ 125\,ms.
  \item GPT-2 iter: 280 $\to$ 210\,ms.
\end{itemize}

The dispatch-path work was harder to land than the allocator
work because the fast path had to be exhaustively tested to
ensure correctness in edge cases (subtle promotion rules, the
\code{0-d} tensor case, etc.). The 3.1.x patch sequence
fixed a long tail of these.

\section{Phase 4: Spurious Synchronisation Removal (3.1.5 $\to$ 3.1.9)}
\label{sec:stabilization:phase_4}

The 3.1.5--3.1.9 patches hunted down unnecessary synchronisation
calls. The methodology:

\begin{enumerate}
  \item Instrumented \code{lucid.runtime.synchronize} to log
    its call sites.
  \item Ran the model zoo's benchmark suite with the
    instrumentation on.
  \item Sorted log entries by frequency; investigated the top
    20.
\end{enumerate}

The leading culprits:

\begin{table}[t]
\centering
\small
\begin{tabular}{lr}
\toprule
Site & Per-iter calls \\
\midrule
\code{tensor.item()} from DataLoader   & 32 \\
\code{tensor.to('cpu')} for visualisation hooks & 7 \\
\code{tensor.cpu().numpy()} for logging & 4 \\
\code{torch.cuda.synchronize()} compatibility shim & 6 \\
Implicit sync in \code{tensor.\_\_repr\_\_} during debug & 1--N \\
\bottomrule
\end{tabular}
\caption[Spurious synchronisation call sites.]{Top sources of
spurious synchronisation in the stabilisation-phase profiling.
The DataLoader case alone added 32 syncs per iteration, each
forcing a GPU $\to$ CPU round-trip.}
\label{tab:stabilization:sync_sites}
\end{table}

\subsection{Fixes}
\label{ssec:stabilization:phase_4_fixes}

\begin{itemize}
  \item The DataLoader stopped calling \code{.item()} on
    integer labels; instead it kept them as
    \code{Tensor}.
  \item \code{tensor.to('cpu')} became async by default
    (returns a tensor that auto-syncs on first
    \code{.numpy()} access).
  \item The compatibility shim
    \code{torch.cuda.synchronize} (used to ease porting from
    CUDA code) was changed to be a no-op, with a deprecation
    warning.
  \item Debug-mode \code{\_\_repr\_\_} was made opt-in (skip
    if the tensor has $>1000$ elements; print shape only).
\end{itemize}

\subsection{Speed-ups}
\label{ssec:stabilization:phase_4_speedups}

\begin{itemize}
  \item ResNet-50 iter: 125 $\to$ 105\,ms.
  \item GPT-2 iter: 210 $\to$ 175\,ms.
\end{itemize}

A small but free win: the cost was almost entirely auditing,
not new code.

\section{Phase 5: Op-Specific Optimisations (3.2.0 $\to$ 3.2.2)}
\label{sec:stabilization:phase_5}

The 3.2 release rounded out the gap with a series of focused
op-level improvements:

\subsection{Softmax}
\label{ssec:stabilization:softmax}

The 3.0 softmax was a 5-op composite (max, sub, exp, sum, div).
The 3.2 release wired in MPSGraph's fused
\code{softMaxWithTensor:axis:} for the GPU stream, reducing it
to a single graph node. End-to-end: BERT inference dropped
from 85 to 75\,ms.

\subsection{LayerNorm}
\label{ssec:stabilization:layernorm}

LayerNorm is called twice per Transformer block; for GPT-2
(12 blocks) that's 24 calls per layer. The 3.0 composite
LayerNorm took $\sim$0.6\,ms per call; the 3.2 MPSGraph-fused
\code{layerNormalizationWithTensor} took 0.18\,ms. Saved 10\,ms
per iter for GPT-2.

\subsection{Attention}
\label{ssec:stabilization:attention}

The biggest win: SDPA via MPSGraph's
\code{scaledDotProductAttention:queries:keys:values:scale:mask:}
(macOS 26+ feature). Replaced the 4-op composite with a single
fused kernel that does the matmul $\to$ scale $\to$ mask $\to$
softmax $\to$ matmul in one pass, with no materialisation of
the $QK^\top$ matrix. Saved 40\,ms per iter for GPT-2.

\subsection{Cumulative result}
\label{ssec:stabilization:cumulative}

End of 3.2.2 vs end of 3.0.0:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & 3.0.0 (ms) & 3.2.2 (ms) & vs ref \\
\midrule
ResNet-50 train       & 240 & 85  & 0.89$\times$ \\
ResNet-101 train      & 480 & 165 & 0.86$\times$ \\
VGG-16 train          & 425 & 220 & 0.95$\times$ \\
GPT-2 small train     & 380 & 155 & 1.06$\times$ \\
GPT-2 medium train    & 940 & 430 & 1.09$\times$ \\
BERT-base train       & 320 & 140 & 1.05$\times$ \\
BERT-large train      & 880 & 410 & 1.10$\times$ \\
DETR-R50 train        & 720 & 410 & 1.18$\times$ \\
Mask R-CNN R50 train  & 920 & 470 & 1.22$\times$ \\
\bottomrule
\end{tabular}
\caption[Stabilisation chain results.]{Per-step training time
before (3.0.0) and after (3.2.2) the stabilisation chain on
M3~Max. The ``vs ref'' column is \Lucid's time divided by the
reference framework's time on the same hardware; values
$<1$ mean \Lucid\ is faster, $>1$ slower. By 3.2.2, \Lucid\ was
faster on the CNN side and within 20\% on the harder
detection/transformer workloads.}
\label{tab:stabilization:cumulative}
\end{table}

\section{What Drove the Wins}
\label{sec:stabilization:what_drove}

In retrospect:

\begin{enumerate}
  \item \textbf{Allocator}: 30\% of the total win. The biggest
    single bottleneck.
  \item \textbf{Dispatch inlining}: 25\%. Many small calls add
    up.
  \item \textbf{Fused MPSGraph ops}: 25\%. SDPA + LayerNorm
    + softmax.
  \item \textbf{Sync removal}: 10\%. Free win.
  \item \textbf{Long tail}: 10\%. Many small fixes.
\end{enumerate}

The lesson: profile-then-fix, not guess-then-fix. The 3.0
hypothesis had been that the GPU kernels were the bottleneck;
the profiling revealed it was the allocator, the dispatch path,
and spurious syncs --- all CPU-side.

\section{The Cost: Complexity}
\label{sec:stabilization:complexity_cost}

The optimisations were not free in code-complexity terms:

\begin{itemize}
  \item Allocator gained \(\sim\)\,400 lines (the cache,
    eviction, statistics).
  \item Dispatch fast-path added \(\sim\)\,150 lines (the
    fast-path check + dispatch).
  \item Per-op MPSGraph emitters added \(\sim\)\,300 lines each.
  \item Test coverage grew correspondingly: 250+ new tests
    targeting the optimised paths.
\end{itemize}

In all cases, the slow path is still present and used as the
correctness reference; the fast path is opt-in and falls back
on edge cases. This is the right shape for performance code:
\emph{never delete the correctness-validated slow path until
the fast path has cleared a year of production usage}.

\section{What 3.2.2 Did Not Fix}
\label{sec:stabilization:not_fixed}

The stabilisation chain did not address:

\begin{itemize}
  \item Mixed-precision training: a 3.3 deliverable
    (\chref{ch:amp_integration}).
  \item Graph capture / compile: a 3.5 deliverable
    (\chref{ch:compile_design}).
  \item Distributed training: not yet planned.
  \item FP8 training: requires HW support not yet in M-series.
\end{itemize}

By 3.2.2, the eager path was within 20\% of the reference
framework's eager throughput. The further gains came from
crossing into compiled execution.

\section{Lessons Learned}
\label{sec:stabilization:lessons}

\begin{enumerate}
  \item \textbf{Measure before optimising.} The intuitive
    bottleneck (kernels) was not the real one (CPU overheads).
  \item \textbf{Optimise the allocator early.} Memory churn is
    the silent killer; a stable allocator is the foundation.
  \item \textbf{Cache only behind a measured benefit.} The
    fast-path inlining was justified by profiling; otherwise
    it would be premature optimisation.
  \item \textbf{Keep the slow path.} Performance code is
    correctness code's accomplice, not its replacement.
  \item \textbf{Test what you optimise.} Each fast-path got
    its own test ID + parity check.
\end{enumerate}

\section{Summary and Forward References}
\label{sec:stabilization:summary}

The 3.0.0 $\to$ 3.2.2 stabilisation chain closed the eager-mode
performance gap to the reference framework. The wins came from
the allocator, the dispatch path, MPSGraph-fused ops, and
sync-call hygiene --- in roughly equal proportions. The eager
path is now production-grade.

The next chapter, \chref{ch:amp_integration}, treats AMP's
performance integration. \chref{ch:mpsgraph_dispatch} treats
the MPSGraph dispatcher details. \chref{ch:no_compile_sweep}
sweeps the no-compile baselines across the entire model zoo.
\chref{ch:graph_capture_wins} measures what
\code{lucid.compile} adds on top of all of this.

\begin{lucidnote}
For the user: if you observe a workload that runs slower than
the reference framework on the same Mac, profile first, then
report. The bench harness's output is the most actionable
artifact; raw stopwatch numbers are rarely useful.
\end{lucidnote}
