% !TEX root = ../main.tex
\chapter{AMP Integration (3.3.0)}
\label{ch:amp_integration}

The 3.3.0 release delivered automatic mixed-precision (AMP)
\cite{micikevicius_mixed_2018}. The chapter
\chref{ch:amp} describes the API and the gradient-scaling
mechanics; this chapter is the performance-engineering side:
what the AMP implementation looks like, what speedups it
actually delivered on Apple silicon, the dtype matrix
between FP16, BF16, and the upcoming FP8 \cite{micikevicius_fp8_2022},
how AMP interacts with the allocator and dispatcher, and the
debug story when AMP changes a model's training behaviour.

\section{Goals}
\label{sec:amp_integration:goals}

The 3.3.0 AMP work targeted three goals:

\begin{enumerate}
  \item \textbf{Throughput}: 1.5--2$\times$ speedup over FP32
    on the model zoo's larger workloads (ViT-Large, GPT-2
    medium, ResNet-101).
  \item \textbf{Memory}: roughly half the activation memory
    of FP32, enabling 2$\times$ larger batches.
  \item \textbf{Correctness}: parity with FP32 training
    trajectory (final accuracy within $\pm$\,0.5\,\%) after
    convergence.
\end{enumerate}

The reach goal was BF16-only training (no gradient scaling), but
this turned out to require careful per-op autocast policy
work that landed only in 3.3.4.

\section{Dtype Matrix on Apple Silicon}
\label{sec:amp_integration:dtype_matrix}

Apple silicon's native low-precision support, as of M3:

\begin{table}[t]
\centering
\small
\begin{tabular}{lccc}
\toprule
Dtype  & GPU TFLOPs (M3 Max) & Memory & Notes \\
\midrule
FP32 & 14   & 4B/elem  & Reference \\
FP16 & 28   & 2B/elem  & Native, requires loss scaling \\
BF16 & 28   & 2B/elem  & Native, no loss scaling \\
INT8 & 56   & 1B/elem  & Inference quantisation only \\
FP8  & ---  & 1B/elem  & Not in M3 (planned) \\
\bottomrule
\end{tabular}
\caption[Apple-silicon dtype throughput.]{Per-element throughput
of arithmetic and storage cost by dtype on M3~Max. FP16 and
BF16 throughput is identical at the hardware level; the choice
comes down to numerical properties (BF16 has FP32's exponent
range but FP16's mantissa precision).}
\label{tab:amp_integration:dtype_matrix}
\end{table}

The headline ``2$\times$'' compute throughput at FP16/BF16 is
the upper bound; achieving it requires that the workload is
arithmetic-bound (matmul-dominated) rather than memory-bound.

\section{Implementation Outline}
\label{sec:amp_integration:impl_outline}

The 3.3.0 AMP implementation has three components:

\begin{enumerate}
  \item \textbf{Autocast context}
    (\code{lucid.amp.autocast}): the user-facing API that
    sets the active dtype for the autocast-eligible op set.
  \item \textbf{Op classification}: each op is classified as
    \emph{whitelist} (run at autocast dtype), \emph{blacklist}
    (force FP32), or \emph{neutral} (follow input dtype).
  \item \textbf{Gradient scaler}
    (\code{lucid.amp.GradScaler}): scales the loss to
    avoid FP16 underflow in gradients, then unscales before the
    optimiser step.
\end{enumerate}

\subsection{Op classification}
\label{ssec:amp_integration:op_class}

The classification rules are derived from the AMP literature
and empirical testing:

\begin{itemize}
  \item \textbf{Whitelist} (cast inputs to autocast dtype,
    cast output back to FP32 if part of FP32 graph): matmul,
    conv2d, attention. The dense arithmetic-heavy ops.
  \item \textbf{Blacklist} (force FP32 regardless): sum
    reductions over the full tensor (numerical accumulation
    issues), softmax (range issues in BF16), batchnorm
    statistics, loss functions (especially cross-entropy).
  \item \textbf{Neutral} (follow input dtype): element-wise
    ops, indexing, reshapes.
\end{itemize}

The autocast policy lives in
\code{lucid/amp/policy.py} as a \code{set[str]} per class.
The op dispatcher consults the policy when an autocast context
is active.

\subsection{The autocast context}
\label{ssec:amp_integration:autocast_ctx}

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.amp.autocast(dtype=lucid.float16):
    out = model(x)
    loss = loss_fn(out, y)
scaler.scale(loss).backward()
scaler.step(optimizer)
scaler.update()
\end{lstlisting}

Inside the context, whitelisted ops cast their inputs to FP16
(via a cheap cast op) and run at FP16. Blacklisted ops cast
inputs to FP32. Neutral ops pass through.

\subsection{Cast op fusion}
\label{ssec:amp_integration:cast_fusion}

The naive implementation inserts a cast op around every
autocast-eligible call, which can add 10--20\% overhead. The
3.3.2 patch added cast-op fusion: chains of casts (e.g., FP32
$\to$ FP16 $\to$ FP32) cancel; the dispatcher recognises and
elides them.

Combined with the allocator cache, casts that are not eliminated
hit the cache (the cast'd tensor is allocated at the same shape
as recently freed tensors), making the cast effectively free.

\section{Speed-ups Delivered}
\label{sec:amp_integration:speedups}

Measured on M3~Max, batch 32, after the 3.3.4 patch line:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & FP32 (ms) & FP16 (ms) & Speedup \\
\midrule
ResNet-50 train       & 85  & 58  & 1.47$\times$ \\
ResNet-101 train      & 165 & 105 & 1.57$\times$ \\
VGG-16 train          & 220 & 130 & 1.69$\times$ \\
ConvNeXt-Tiny train   & 95  & 64  & 1.48$\times$ \\
ViT-Base train        & 125 & 80  & 1.56$\times$ \\
ViT-Large train       & 410 & 235 & 1.74$\times$ \\
GPT-2 small train     & 155 & 105 & 1.48$\times$ \\
GPT-2 medium train    & 430 & 250 & 1.72$\times$ \\
BERT-base train       & 140 & 95  & 1.47$\times$ \\
T5-base train         & 290 & 175 & 1.66$\times$ \\
\bottomrule
\end{tabular}
\caption[AMP speedups.]{FP32 vs FP16 (autocast) iteration time
on M3~Max. Speedup increases with model size, as the
arithmetic-bound matmuls dominate more of the iteration.
Smaller models hit the dispatch-overhead floor and gain less.}
\label{tab:amp_integration:speedups}
\end{table}

The pattern is consistent: small models gain $\sim$1.5$\times$,
large models gain $\sim$1.7$\times$. The asymptote (theoretical
2$\times$) is not reached because the dispatch path is still
the bottleneck for small ops, and the blacklisted ops (which
stay FP32) contribute proportionally to the floor.

\section{Memory Savings}
\label{sec:amp_integration:memory_savings}

Activation memory under FP16 vs FP32:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & FP32 peak (GB) & FP16 peak (GB) & Ratio \\
\midrule
ResNet-50 train       & 4.2 & 2.5 & 0.60 \\
ResNet-101 train      & 7.8 & 4.5 & 0.58 \\
VGG-16 train          & 6.5 & 3.7 & 0.57 \\
ViT-Base train        & 5.8 & 3.4 & 0.59 \\
ViT-Large train       & 18.5& 10.2& 0.55 \\
GPT-2 medium train    & 14.2& 8.1 & 0.57 \\
BERT-large train      & 19.8& 11.0& 0.56 \\
\bottomrule
\end{tabular}
\caption[Activation memory under AMP.]{FP32 vs FP16 peak
activation memory on M3~Max. The ratio is roughly 0.55--0.60
(not 0.5) because optimiser state, gradient FP32-master copies,
and FP32-fallback activations contribute to the FP16 total.}
\label{tab:amp_integration:memory_savings}
\end{table}

The savings let users train at 2$\times$ larger batches (or
2$\times$ larger models at the same batch). For users on the
24\,GB unified-memory M3, this is often the difference between
``fits'' and ``OOM''.

\section{BF16 vs FP16}
\label{sec:amp_integration:bf16_vs_fp16}

BF16 has FP32's exponent range (8 bits) with FP16's mantissa
(7 bits) instead of FP16's 5+10 layout. The implications:

\begin{itemize}
  \item \textbf{No loss scaling needed} for BF16: gradients
    can be small without underflowing.
  \item \textbf{Less precision} per number, so finer-grained
    differences may be lost. In practice, this rarely matters
    for training.
  \item \textbf{Same arithmetic throughput} as FP16 on Apple
    silicon.
\end{itemize}

Recommendation: BF16 by default; FP16 only if compatibility with
non-BF16-capable hardware matters. Both are equally fast.

The autocast context selects:

\begin{lstlisting}[style=lucid,language=LucidPython]
with lucid.amp.autocast(dtype=lucid.bfloat16):  # No GradScaler needed
    out = model(x)
    loss = loss_fn(out, y)
loss.backward()
optimizer.step()
\end{lstlisting}

\section{The Gradient Scaler}
\label{sec:amp_integration:grad_scaler}

For FP16 training, the gradient scaler addresses the underflow
problem: FP16's smallest normal is $\sim$$6 \times 10^{-5}$, and
many real gradients are smaller. The scaler multiplies the loss
by a large factor (typically $2^{15}$) before backward; gradients
flow at the scaled magnitude (avoiding underflow); the scaler
unscales them before the optimiser step.

\subsection{Dynamic scaling}
\label{ssec:amp_integration:dynamic_scaling}

The fixed scale factor must be chosen carefully: too small
underflows still happen; too large overflows happen.
\code{lucid.amp.GradScaler} uses a dynamic scheme:

\begin{itemize}
  \item Start at $2^{16}$.
  \item Every 2000 iterations without overflow: double.
  \item On overflow detected (any gradient is $\pm\infty$ or
    NaN): halve and skip the optimiser step.
\end{itemize}

\subsection{Overflow detection}
\label{ssec:amp_integration:overflow_detection}

After unscaling, the scaler scans all gradients for inf/NaN.
On Apple silicon, this is done by an MPSGraph fused reduction
(\code{isfinite + reduce\_and}) that adds $\sim$50\,\textmu s
per iter.

\begin{lstlisting}[style=lucid,language=LucidPython]
scaler.scale(loss).backward()
scaler.step(optimizer)   # Skips if overflow detected
scaler.update()          # Adjusts scale factor
\end{lstlisting}

\section{Compatibility Pitfalls}
\label{sec:amp_integration:compatibility}

The 3.3 release surfaced several user-facing pitfalls:

\subsection{Models with manual mixed precision}
\label{ssec:amp_integration:manual_mixed}

If a user model contains \code{.half()} casts (legacy
manual-FP16 code), the autocast context can interact badly.
The 3.3.2 patch documented the recommended pattern: remove the
manual casts and let autocast handle dtype management.

\subsection{Custom Functions and AMP}
\label{ssec:amp_integration:custom_fn_amp}

User-written \code{lucid.autograd.Function} subclasses are
neutral by default (follow input dtype). For correct AMP
behaviour, the user adds a class attribute:

\begin{lstlisting}[style=lucid,language=LucidPython]
class MyOp(lucid.autograd.Function):
    amp_policy = "fp32"  # or "fp16", "bf16", "neutral"
    @staticmethod
    def forward(ctx, x):
        ...
\end{lstlisting}

\subsection{Optimisers and FP32 master copies}
\label{ssec:amp_integration:fp32_master}

For correctness, all standard optimisers keep an FP32 master
copy of the weights; the weights stored in the module are FP16
(or BF16). The optimiser step:
\begin{enumerate}
  \item Receives FP16 gradients.
  \item Casts to FP32.
  \item Updates the FP32 master.
  \item Casts back to FP16 and writes to the module.
\end{enumerate}

The user does not see this; it is handled by the optimiser
when it detects autocast-active weights.

\section{Validation: Training Curves}
\label{sec:amp_integration:validation}

The acceptance test for AMP was: training trajectories match
FP32 within $\pm$\,0.5\,\% final accuracy. Measured on
ResNet-50 / ImageNet-1K (100 epochs):

\begin{itemize}
  \item FP32: 76.04\% top-1.
  \item BF16: 76.10\% top-1 ($+0.06\%$ within noise).
  \item FP16 + GradScaler: 75.98\% ($-0.06\%$ within noise).
\end{itemize}

All three are within run-to-run variance ($\pm$0.2\%). The
acceptance was clear and AMP was promoted from experimental
to stable in 3.3.4.

For language-model pretraining (GPT-2 small, OpenWebText, 100K
steps):

\begin{itemize}
  \item FP32: validation perplexity 19.4.
  \item BF16: 19.5.
  \item FP16 + GradScaler: 19.5 (with one overflow event in
    the first 1000 steps, recovered by GradScaler).
\end{itemize}

\section{What Was Hard}
\label{sec:amp_integration:hard_parts}

The 3.3.0--3.3.4 patch sequence had several surprises:

\subsection{Softmax in BF16}
\label{ssec:amp_integration:softmax_bf16}

Initially, softmax was on the whitelist (let it run BF16). The
result: occasional NaN in attention output, when a row's
maximum logit was $\sim$11 (within BF16's range) but the
post-softmax distribution had values that underflowed BF16's
mantissa. The fix: blacklist softmax (always FP32).

\subsection{The 1$\times$1 conv special case}
\label{ssec:amp_integration:1x1_conv}

For conv2d, MPSGraph routes 1$\times$1 convs to a matmul kernel.
The matmul kernel had a stricter dtype contract than the conv
kernel (no implicit casting). The 3.3.1 patch added an explicit
cast in the 1$\times$1 conv path.

\subsection{Allocator interaction}
\label{ssec:amp_integration:allocator_interaction}

The free-list allocator
(\chref{ch:stabilization}~Section~\ref{sec:stabilization:phase_2})
keys on shape only. Under AMP, the same shape with FP16 and FP32
should not share buckets (different byte counts). The 3.3.0
allocator change widened the key to \code{(shape, dtype)}.

\section{Looking Forward: FP8}
\label{sec:amp_integration:fp8_outlook}

Apple silicon does not yet have native FP8 \cite{micikevicius_fp8_2022}.
When it lands (likely M5 or M6), the AMP infrastructure is
ready: the autocast policy generalises to FP8, the GradScaler
mechanism applies, the casting infrastructure (and fusion) is
in place. The work will be:
\begin{itemize}
  \item Add FP8 emitters for the matmul/conv MPSGraph wrappers.
  \item Tune the per-op policy (FP8 has even less range than
    FP16, so more ops will be blacklisted).
  \item Validate accuracy on the model zoo.
\end{itemize}

Until then, BF16 is the recommended dtype for training.

\section{Summary and Forward References}
\label{sec:amp_integration:summary}

AMP brought 1.5--1.7$\times$ training speedups and roughly half
the activation memory, validating the engineering investment.
The implementation: autocast context + op classification +
gradient scaler. BF16 is preferred over FP16 when available.
The infrastructure is FP8-ready when hardware lands.

The next chapter, \chref{ch:mpsgraph_dispatch}, dives into the
MPSGraph dispatcher tuning. \chref{ch:no_compile_sweep} reports
the no-compile baseline across the full model zoo.
\chref{ch:graph_capture_wins} measures what
\code{lucid.compile} adds.

\begin{lucidnote}
For the user: enable \code{lucid.amp.autocast(dtype=bfloat16)}
in your training script; that is the typical 1.5$\times$ free
speedup on M3 Max and larger models. If you need FP16 for
compatibility, add a \code{GradScaler}; for everyone else,
BF16 is simpler and faster.
\end{lucidnote}
