% !TEX root = ../main.tex
\chapter{MPSGraph Per-Op Dispatch (3.4.0)}
\label{ch:mpsgraph_dispatch}

By 3.3.4 the eager-mode path was performance-competitive with
the reference framework on small models, slightly better on
the CNNs, and within 20\% on transformers. The remaining gap on
larger workloads came from two sources: per-op dispatch
overhead (still 8--15\% of iteration time for transformer
models) and missed fusion opportunities at the MPSGraph
boundary. The 3.4.0 release closed both gaps via a deeper
integration with MPSGraph: per-op graph caching, executable
reuse, and a small library of fused emitters for hot patterns.
This chapter is the engineering retrospective.

\section{Why MPSGraph and Not Raw Metal}
\label{sec:mpsgraph_dispatch:why_mpsgraph}

A digression first. \Lucid\ runs GPU ops via MPSGraph
\cite{apple_mpsgraph}, Apple's graph-execution library, rather
than writing Metal Shading Language kernels directly. The
trade-off:

\begin{itemize}
  \item \textbf{MPSGraph pros}: tuned per-op kernels (Apple
    has done the autotuning work), automatic kernel fusion
    when chained, hardware-aware dispatch
    (\code{MTLDevice} placement), supports newer ops out of
    the box.
  \item \textbf{MPSGraph cons}: per-op construction overhead
    (a graph node is a heavy object), opaque kernel-selection
    decisions, dependency on macOS-version-specific features.
  \item \textbf{Raw Metal pros}: direct control over kernel
    code, no graph-construction overhead.
  \item \textbf{Raw Metal cons}: \Lucid\ would have to write
    and tune kernels for every op, every dtype, every shape
    pattern --- prohibitive work.
\end{itemize}

The decision in 3.0 was MPSGraph for breadth; the 3.4 work
focused on minimising the per-op construction overhead while
keeping the breadth.

\section{The Per-Op Graph Cache}
\label{sec:mpsgraph_dispatch:graph_cache}

Before 3.4, every op invocation constructed a fresh
\code{MPSGraph} object, added one node, compiled, executed,
and discarded the graph. The compile step was the bottleneck:
$\sim$\,40\,\textmu s of overhead per op for a graph with one
node.

The 3.4 solution: cache the constructed graphs keyed by op +
shape + dtype. Subsequent invocations with the same key reuse
the cached graph, skipping construction and compile.

\subsection{Cache key}
\label{ssec:mpsgraph_dispatch:cache_key}

\begin{lstlisting}[style=lucid,language=LucidPython]
key = (
    op_name,                  # "add", "matmul", etc.
    tuple(input_shapes),
    tuple(input_dtypes),
    output_dtype,
    op_attrs_tuple,           # axis, keepdim, etc.
)
\end{lstlisting}

The cache is a \code{LRUCache} (size 4096) protected by a
read-write lock (most accesses are reads). On a cache hit, the
saved \code{MPSGraphExecutable} is invoked directly with the
new input buffers; no graph construction.

\subsection{Hit rates}
\label{ssec:mpsgraph_dispatch:hit_rates_mpsgraph}

For typical workloads, the cache hits 98\%+ after the first
iteration:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Workload & Unique keys (post-warmup) & Hit rate \\
\midrule
ResNet-50 train     & 412 & 99.2\% \\
GPT-2 small train   & 287 & 99.5\% \\
BERT-base train     & 244 & 99.4\% \\
DETR train          & 1041 (variable shapes) & 88\% \\
\bottomrule
\end{tabular}
\caption[Per-op graph cache hit rates.]{The cache is highly
effective on fixed-shape workloads (training with a fixed batch
size). The detection workloads are lower because the proposal
count varies per image, producing many shape variants.}
\label{tab:mpsgraph_dispatch:hit_rates}
\end{table}

\subsection{Cache size and eviction}
\label{ssec:mpsgraph_dispatch:cache_eviction}

The 4096 entry limit is conservative; for any model with
$<$\,1000 unique op-shape combinations, the cache never evicts.
For the dynamic-shape detection workloads, LRU eviction does
happen; the eviction cost (drop a graph executable) is fast,
and re-construction on the next call is the slow path that was
the pre-3.4 normal.

\subsection{Memory footprint}
\label{ssec:mpsgraph_dispatch:cache_memory}

A cached graph entry is $\sim$\,2--10\,KB (the
\code{MPSGraphExecutable} plus metadata). At 4096 entries, the
cache caps at $\sim$\,40\,MB --- negligible relative to a
training workload.

\section{Speed-ups from Caching Alone}
\label{sec:mpsgraph_dispatch:cache_speedups}

The graph cache landed in 3.4.0-rc1. The measured deltas:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & 3.3.4 (ms) & 3.4.0 (ms) & Speedup \\
\midrule
ResNet-50 train       & 85  & 75  & 1.13$\times$ \\
ResNet-101 train      & 165 & 142 & 1.16$\times$ \\
ViT-Base train        & 125 & 105 & 1.19$\times$ \\
ViT-Large train       & 410 & 335 & 1.22$\times$ \\
GPT-2 small train     & 155 & 130 & 1.19$\times$ \\
GPT-2 medium train    & 430 & 350 & 1.23$\times$ \\
BERT-base train       & 140 & 115 & 1.22$\times$ \\
\bottomrule
\end{tabular}
\caption[Graph-cache speedups.]{3.3.4 (graph constructed per
call) vs 3.4.0 (graph cached) iteration time. The gain
correlates with op count: GPT-2 medium has the most ops per
iteration and gains the most.}
\label{tab:mpsgraph_dispatch:cache_speedups}
\end{table}

A clean 1.2$\times$ win on transformer workloads from a
purely-infrastructure change.

\section{Fused Emitters}
\label{sec:mpsgraph_dispatch:fused_emitters}

The second 3.4 deliverable was a small library of
\emph{fused emitters}: high-level ops that the \Lucid\
dispatcher recognises and emits as a single fused MPSGraph
operation rather than a chain. The fusion catalogue:

\subsection{Bias-add + activation}
\label{ssec:mpsgraph_dispatch:bias_act}

The common pattern \texttt{F.relu(x + b)} (or with GELU, SiLU,
etc.) was emitted as a chain of two MPSGraph ops. The fused
emitter:

\begin{lstlisting}[style=lucid,language=LucidPython]
y = lucid.nn.functional.linear_bias_act(x, w, b, act="gelu")
# Single MPSGraph node: matmul + bias-add + GELU
\end{lstlisting}

The fusion saves the intermediate-tensor allocation and one
graph-execution round-trip; on a GPT-2 forward (12 blocks,
each with linear-bias-act in the FFN), this is 6\,ms saved.

\subsection{LayerNorm + linear (NormLinear)}
\label{ssec:mpsgraph_dispatch:norm_linear}

In a pre-norm Transformer, each block begins with LayerNorm
followed by a linear. The fused emitter computes both in one
graph node, saving the materialisation of the
post-LayerNorm tensor.

\subsection{SDPA}
\label{ssec:mpsgraph_dispatch:sdpa}

Already discussed at length in
\chref{ch:attention}~Section~\ref{sec:attention:sdpa}: the
MPSGraph fused SDPA replaces a chain of 4 ops (Q@K, scale,
softmax, @V) with one. On Apple silicon, the fused kernel does
not materialise the $T \times T$ attention matrix; for long
sequences (512+), this is the difference between fitting and
not.

\subsection{The fusion registry}
\label{ssec:mpsgraph_dispatch:fusion_registry}

The dispatcher consults a small registry of fused emitters; a
top-level op call (\code{lucid.nn.functional.scaled\_dot\_product\_attention})
maps directly to the fused emitter when available, falling back
to the composite path when not:

\begin{lstlisting}[style=lucid,language=LucidPython]
# In lucid/_C/ops/fused.cpp:
REGISTER_FUSED_EMITTER("sdpa", emit_sdpa);
REGISTER_FUSED_EMITTER("linear_bias_act", emit_linear_bias_act);
REGISTER_FUSED_EMITTER("layer_norm_linear", emit_norm_linear);
\end{lstlisting}

The fallback path is the composite (chain of base ops) and
remains the correctness reference.

\section{Combined Effect}
\label{sec:mpsgraph_dispatch:combined_effect}

By 3.4.4 (final 3.4 patch), with graph cache + fused emitters:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & 3.3.4 (ms) & 3.4.4 (ms) & Speedup \\
\midrule
ResNet-50 train       & 85  & 72  & 1.18$\times$ \\
ViT-Base train        & 125 & 95  & 1.32$\times$ \\
ViT-Large train       & 410 & 295 & 1.39$\times$ \\
GPT-2 small train     & 155 & 110 & 1.41$\times$ \\
GPT-2 medium train    & 430 & 290 & 1.48$\times$ \\
BERT-base train       & 140 & 92  & 1.52$\times$ \\
T5-base train         & 290 & 200 & 1.45$\times$ \\
DETR train            & 410 & 320 & 1.28$\times$ \\
\bottomrule
\end{tabular}
\caption[3.4.4 cumulative speedups.]{3.3.4 (graph constructed
per call, no fused emitters) vs 3.4.4 (cache + fused) iteration
time on M3~Max. Transformer models gain the most because they
benefit from both improvements; CNN gains are smaller but
non-trivial.}
\label{tab:mpsgraph_dispatch:combined_effect}
\end{table}

The combined effect on BERT-base is 1.52$\times$ faster than
3.3.4 --- nearly half the iteration time eliminated by
infrastructure-only changes (no model-side code touched).

\section{Detailed Profile Diff}
\label{sec:mpsgraph_dispatch:profile_diff}

A representative GPT-2 small forward pass before/after, by
phase:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Phase & 3.3.4 (ms) & 3.4.4 (ms) \\
\midrule
Python dispatch overhead   & 18  & 16 \\
C++ op dispatch + reg lookup & 12  & 4 \\
MPSGraph construction      & 31  & 0.5 \\
MPSGraph execution         & 75  & 70 \\
Memory allocator           & 8   & 7 \\
Cast ops (autocast)        & 4   & 2 \\
Other                      & 7   & 10 \\
\midrule
Total                      & 155 & 110 \\
\bottomrule
\end{tabular}
\caption[Phase-level profile diff.]{GPT-2 small forward,
batch 32, sequence 512. The biggest 3.4 wins: MPSGraph
construction (cached, near-zero) and C++ dispatch + registry
lookup (cache-side fast path). The kernel-execution phase
remains the dominant cost --- the MPSGraph compute itself ---
and is largely unchanged.}
\label{tab:mpsgraph_dispatch:profile_diff}
\end{table}

The kernel-execution time stays essentially the same; the
optimisations targeted the overhead, not the math. This
matches the intuition: at FP16, GPT-2 small's matmuls run at
$\sim$\,70\,\% of M3 Max's FP16 peak --- already near the
ceiling, and further gains require either better kernels
(Apple's job) or fewer kernel calls (the fused emitters).

\section{What Did Not Work}
\label{sec:mpsgraph_dispatch:what_did_not_work}

Two attempted optimisations that did not deliver and were
reverted:

\subsection{Multi-graph batched dispatch}
\label{ssec:mpsgraph_dispatch:multi_graph}

The 3.4.0-beta tried to batch multiple ops into a single
MPSGraph execution, reasoning that one large graph would
amortise the kernel-launch overhead. The result was actually
slower (kernel-launch is cheap; graph-construction and
intermediate-allocation overhead dominated). Reverted.

\subsection{Per-op kernel pre-warming}
\label{ssec:mpsgraph_dispatch:pre_warming}

Initial idea: at module-load time, construct graphs for the
common ops at common shapes to warm the cache. Discarded
because (a) it added 800\,ms to import time and (b) the
warmed entries were rarely the ones the user's specific
model needed.

\section{Apple-Silicon-Specific Lessons}
\label{sec:mpsgraph_dispatch:apple_lessons}

The 3.4 work surfaced several MPSGraph-specific gotchas worth
documenting for future maintainers:

\begin{itemize}
  \item \textbf{Graph construction is the bottleneck},
    not graph execution. The kernels themselves are fast.
  \item \textbf{Per-op cache pays off massively}; per-graph
    (multi-op) cache pays nothing on top.
  \item \textbf{Fused MPSGraph ops are the only path to
    sub-kernel-count execution}; you cannot batch dispatches
    into a single Metal command buffer from outside MPSGraph.
  \item \textbf{Dtype-change in cached graphs is fatal}: the
    cache key must include dtype, not just shape.
  \item \textbf{Stride changes invalidate the cache}: graphs
    are stride-specific. \Lucid\ contiguifies tensors before
    dispatching to maximise cache hits.
\end{itemize}

\section{The MPSGraph Dispatcher Interface}
\label{sec:mpsgraph_dispatch:dispatcher_iface}

The cleaned-up 3.4 interface in
\code{lucid/backend/metal/mpsgraph\_dispatcher.h}:

\begin{lstlisting}[style=lucid,language=LucidPython]
class MpsGraphDispatcher:
  cache: LRUCache[GraphKey, MPSGraphExecutable]

  def dispatch(op_name, inputs, attrs) -> Tensor:
    key = make_key(op_name, inputs, attrs)
    if exe := cache.get(key):
      return exe.run(inputs)
    graph = build_graph(op_name, inputs, attrs)
    exe = graph.compile()
    cache.put(key, exe)
    return exe.run(inputs)
\end{lstlisting}

The hot path is a single hash lookup + a single
\code{exe.run}; the cold path adds the build + compile.

\section{Summary and Forward References}
\label{sec:mpsgraph_dispatch:summary}

The 3.4 release closed the remaining eager-mode performance
gap to the reference framework via graph caching and fused
emitters, both purely-infrastructure changes that delivered
1.2--1.5$\times$ speedups on transformer workloads. The
fundamental kernel performance is now near the hardware ceiling;
further gains require crossing into compiled execution (treated
in the next part).

The next chapter, \chref{ch:no_compile_sweep}, reports the full
no-compile baseline across the model zoo at 3.4.4.
\chref{ch:graph_capture_wins} measures what \code{lucid.compile}
adds on top.

\begin{lucidnote}
For the user: the graph cache and fused emitters are
transparent --- nothing to enable; the dispatcher uses them
automatically. The one observable effect is the first iteration
of a new model being $\sim$\,2--3$\times$ slower than steady
state (cache cold); discount the first-iteration timing in
benchmarks.
\end{lucidnote}
