% !TEX root = ../main.tex
\chapter{No-Compile Algorithmic Sweep (3.4.1)}
\label{ch:no_compile_sweep}

After the infrastructure-side optimisations of 3.3 (AMP) and
3.4 (graph cache + fused emitters), \Lucid's eager path is
within parity of the reference framework on most of the
zoo. The 3.4.1 release shifted attention to a different class
of optimisation: \emph{algorithmic} improvements that do not
require the compile path, applicable equally to eager and
compiled execution. This chapter is the sweep's catalogue ---
what was tried, what worked, what landed, what was rejected ---
and gives the final state of the no-compile baseline against
which the compile path (\chref{ch:compile_design}) is measured.

\section{Methodology}
\label{sec:no_compile_sweep:methodology_outline}

The sweep was structured as a fixed list of algorithmic
candidates, each evaluated for:

\begin{enumerate}
  \item \textbf{Correctness}: does the new algorithm produce
    bitwise-equivalent or within-tolerance results?
  \item \textbf{Speedup}: does it actually run faster on
    M-series hardware?
  \item \textbf{Robustness}: does it work across the model
    zoo's full shape/dtype matrix?
  \item \textbf{Maintenance cost}: how much code does it add,
    and does the slow path remain as a reference?
\end{enumerate}

Candidates were ranked by speedup-per-LoC; the top entries
landed in 3.4.1.

\section{The Candidate List}
\label{sec:no_compile_sweep:candidates}

The full sweep evaluated 19 candidates. The shortlist that
landed:

\begin{enumerate}
  \item Winograd convolution for 3$\times$3 stride-1 convs
    (CNN backbones).
  \item Flash-style streaming softmax for SDPA fallback path.
  \item BNNS direct path for small CPU-stream matmuls.
  \item GEMM tiling tuned for M-series GPU clusters.
  \item RoPE precomputed cos/sin cache.
  \item DropPath as a single-shot mask (vs per-block).
\end{enumerate}

We treat each in turn.

\section{Winograd Convolution}
\label{sec:no_compile_sweep:winograd}

For 3$\times$3 stride-1 convolutions (the dominant kernel size
in ResNet/VGG/ConvNeXt), the Winograd algorithm
\cite{dumoulin_guide_2016} reduces multiplies from $9$ to $4$
per output element by transforming both input and filter into
a frequency-like domain.

\subsection{The transform}
\label{ssec:no_compile_sweep:winograd_transform}

The $F(2, 3)$ Winograd transform: pre-multiply input tiles by
$B^\top$ and filters by $G$, perform element-wise multiply,
post-multiply by $A^\top$:
\[
  Y = A^\top \bigl( (G g G^\top) \odot (B^\top d B) \bigr) A.
\]
The $4 \to 9$ multiply reduction is offset by additional
add/subtract operations and the transforms themselves. On
appropriately-sized tiles ($4 \times 4$), the net is a real
1.3$\times$ speedup.

\subsection{Implementation}
\label{ssec:no_compile_sweep:winograd_impl}

MPSGraph provides Winograd internally for some shapes; explicit
opt-in via the \code{enableWinograd:} property. \Lucid\
enables it conditionally for $3 \times 3$ stride-1 conv layers
with input channel count $\geq 64$ (below which the transform
overhead dominates).

\subsection{Result}
\label{ssec:no_compile_sweep:winograd_result}

\begin{itemize}
  \item ResNet-50 train iter: 72 $\to$ 65\,ms (1.11$\times$).
  \item ResNet-101 train iter: 142 $\to$ 122\,ms (1.16$\times$).
  \item ConvNeXt-Base train: 215 $\to$ 195\,ms (1.10$\times$).
\end{itemize}

Free win; the slow path (non-Winograd conv) remains for
correctness fallback.

\section{Flash-style Streaming Softmax}
\label{sec:no_compile_sweep:flash_softmax}

When the MPSGraph fused SDPA falls back (e.g., for unsupported
mask shapes or older macOS), the composite path materialises
the $T \times T$ attention matrix --- $O(T^2)$ memory. For long
sequences this is expensive.

The Flash-style algorithm \cite{dao_flashattention_2022}
streams the softmax in tiles: process $K_t$ at a time, maintain
running max and running sum across tiles, never materialise the
full matrix. The end result is identical (the math is
algebraic) but with $O(T \cdot K_t)$ memory.

\subsection{Implementation}
\label{ssec:no_compile_sweep:flash_impl}

\Lucid's composite-path SDPA was rewritten to stream:

\begin{lstlisting}[style=lucid,language=LucidPython]
def streaming_sdpa(q, k, v, mask=None):
    # Streams keys/values in blocks of TILE; never materializes T x T
    TILE = 64
    out, running_max, running_sum = init_states(q.shape)
    for k_block, v_block in tiles(k, v, TILE):
        scores = q @ k_block.T  # (B, T, TILE)
        if mask is not None:
            scores = scores + mask_block
        # online softmax update
        new_max = max(running_max, scores.max(-1))
        scale = exp(running_max - new_max)
        running_sum = running_sum * scale + exp(scores - new_max).sum(-1)
        out = out * scale.unsqueeze(-1) + exp(scores - new_max) @ v_block
        running_max = new_max
    return out / running_sum.unsqueeze(-1)
\end{lstlisting}

\subsection{Result}
\label{ssec:no_compile_sweep:flash_result}

For shapes where MPSGraph fused SDPA is unavailable (rare on
macOS 26+, common on macOS 25), the streaming path was 1.4--3$\times$
faster than the materialising composite, and used 4--8$\times$
less peak memory.

On macOS 26+, the fused MPSGraph SDPA is preferred and the
streaming path is the fallback. Either way, the user-visible
behaviour is correct.

\section{BNNS Direct Path for Small CPU Matmuls}
\label{sec:no_compile_sweep:bnns_matmul}

For matmuls of dimension $\leq 128$ on the CPU stream, the MLX
fallback path was slower than direct \code{vDSP\_mmul} via
Accelerate's BNNS \cite{apple_accelerate}. The 3.4.1 patch
added a fast path: small CPU matmuls bypass MLX and go to
\texttt{vDSP\_mmul} directly.

The threshold (128) was chosen by sweep on M3~Max. Above 128,
MLX is competitive; below, the function-call overhead of MLX's
async dispatch dominates.

Affected workloads: classifier heads at the end of a backbone
(which run on CPU stream for the final FC layer when batch is
small).

\section{GEMM Tiling for M-Series GPU Clusters}
\label{sec:no_compile_sweep:gemm_tiling}

The M-series GPU has 8--40 cores organised in 1--5 clusters.
For matmuls of intermediate size (say $1024 \times 1024 \times 1024$),
the per-cluster tiling decision affects throughput by
$\sim$\,15\%. MPSGraph's default tiling was conservative; the
3.4.1 sweep tuned the tile sizes for the specific cluster
counts on M3 Pro, M3 Max, M4 Pro, M4 Max.

The tuning is purely a metadata change (passed via
\code{MPSGraphCompilationDescriptor}), not a kernel change.

Result: 1.05--1.12$\times$ matmul speedups across the affected
shape range. Most visible on Transformer blocks where each
layer has 4 matmuls of similar shape.

\section{RoPE Precomputed Cache}
\label{sec:no_compile_sweep:rope_cache}

The rotary position embedding \cite{su_roformer_2021} rotates
$Q$ and $K$ by precomputed angles. The naive implementation
recomputes the cos/sin values per call. The 3.4.1 patch caches
them per (sequence length, dtype, theta) tuple, reusing across
calls.

For language-model inference with sequence growing
incrementally, the cache hits on every step after the first.
Saved $\sim$\,0.4\,ms per token on GPT-2 small generation
(0.5\% of per-step time; small but free).

\section{DropPath Vectorisation}
\label{sec:no_compile_sweep:droppath_vec}

DropPath (stochastic depth) randomly drops entire residual
blocks during training. The naive implementation: per-block,
sample a Bernoulli mask, multiply.

The 3.4.1 vectorisation: at the start of the forward pass,
sample all per-block masks at once, store them, multiply at
each block. Saves 11 RNG calls per iteration (for ViT-Large)
and improves CPU--GPU pipelining (the RNG runs on CPU while
the GPU is busy).

Result: 1.02--1.05$\times$ on ViT-Large.

\section{Candidates Rejected}
\label{sec:no_compile_sweep:rejected}

Candidates that did not land, with the reason:

\subsection{Strassen matrix multiplication}
\label{ssec:no_compile_sweep:strassen}

Strassen's $O(n^{2.807})$ matmul beats naive $O(n^{3})$ at
large dimensions. Tested for $N \times N$ matmul; the
crossover was $N \geq 4096$. The model zoo's matmuls are
typically smaller (transformer hidden dims 768--1024); the
gain at $N = 768$ was negative due to the recursion overhead.
Rejected.

\subsection{Mixed FP16/INT8 quantised matmul}
\label{ssec:no_compile_sweep:mixed_fp16_int8}

Tested INT8 weights + FP16 activations for inference. Accuracy
loss was 0.4--1.2\,\% top-1 (acceptable for some uses) but the
matmul speedup on Apple silicon was only 1.05$\times$ ---
INT8 matmul throughput on M3 GPU is not yet 2$\times$ FP16.
Saved for when INT4/INT8 hardware paths land (M5+).

\subsection{Per-layer activation checkpointing toggling}
\label{ssec:no_compile_sweep:per_layer_ckpt}

The idea was to enable gradient checkpointing per layer based
on a memory/time trade-off score. Implementation was complex
and the benefits were modest; rejected in favour of the
simpler all-or-nothing checkpointing already supported.

\subsection{Custom Metal kernel for LayerNorm}
\label{ssec:no_compile_sweep:custom_metal_ln}

Tested a hand-written Metal kernel that exploited shared
memory. It was 1.15$\times$ faster than MPSGraph's
\code{layerNormalization}, but: (a) MPSGraph improves with
each macOS release; (b) maintenance cost of a custom kernel
is high; (c) the same logic should be added for RMSNorm,
GroupNorm, InstanceNorm, doubling the maintenance load.
Postponed in favour of relying on MPSGraph.

\section{The 3.4.1 Baseline}
\label{sec:no_compile_sweep:baseline}

After 3.4.1 landed, the no-compile model-zoo baseline:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & 3.4.0 (ms) & 3.4.1 (ms) & vs ref \\
\midrule
ResNet-50 train       & 72 & 64 & 0.79$\times$ \\
ResNet-101 train      & 142 & 122 & 0.82$\times$ \\
VGG-16 train          & 220 & 215 & 0.95$\times$ \\
ConvNeXt-Tiny train   & 95 & 90 & 1.02$\times$ \\
ViT-Base train        & 95 & 92 & 0.98$\times$ \\
ViT-Large train       & 295 & 280 & 1.10$\times$ \\
GPT-2 small train     & 110 & 105 & 1.06$\times$ \\
GPT-2 medium train    & 290 & 280 & 1.07$\times$ \\
BERT-base train       & 92 & 88 & 0.95$\times$ \\
T5-base train         & 200 & 190 & 1.04$\times$ \\
Mask R-CNN R50 train  & 480 & 470 & 1.22$\times$ \\
DETR train            & 320 & 310 & 1.14$\times$ \\
DDPM-UNet train       & 105 & 95 & 0.99$\times$ \\
\bottomrule
\end{tabular}
\caption[The 3.4.1 no-compile baseline.]{Iteration time for the
model zoo at 3.4.1, no compile. The ``vs ref'' column is
\Lucid\ time divided by the reference-framework time on the
same hardware. \Lucid\ is now faster than the reference on the
CNN side, within 10\% on transformers, and within 25\% on the
detection-style workloads.}
\label{tab:no_compile_sweep:baseline}
\end{table}

This is the table the project shipped as ``\Lucid\ 3.4.1
performance''. The reference-framework parity is comfortably
within the 20\% acceptance criterion for most models.

\section{Where the Eager-Mode Ceiling Is}
\label{sec:no_compile_sweep:ceiling}

After 3.4.1, the remaining gap between \Lucid\ eager and the
theoretically achievable performance is in two places:

\begin{itemize}
  \item \textbf{Per-op dispatch}: still 5--10\,\% of iteration
    time for transformer models. Cannot be reduced further
    without compile (which fuses dispatches).
  \item \textbf{Memory-bandwidth-bound ops}: LayerNorm and
    softmax saturate the memory subsystem; reducing call count
    via compile lets the next op start sooner (more
    parallelism).
\end{itemize}

These set the stage for the compile work. The next chapter
(\chref{ch:graph_capture_wins}) measures what compile actually
delivers on top of this baseline.

\section{Per-Op Sweep Sample}
\label{sec:no_compile_sweep:per_op_sample}

For interest: a microbenchmark sweep over individual ops at
the start of 3.4.1 vs end (after all the algorithmic landings):

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Op & Pre-3.4.1 (\textmu s) & Post-3.4.1 (\textmu s) \\
\midrule
\code{matmul} (768x768x768)   & 480 & 425 \\
\code{conv2d} (3x3, 64ch, 32x32) & 220 & 180 \\
\code{softmax} (256, 256)      & 35  & 33  \\
\code{layernorm} (1024)        & 22  & 21  \\
\code{rope} (sequence 1024)    & 95  & 12  \\
\code{droppath} (12 blocks)    & 140 & 65  \\
\bottomrule
\end{tabular}
\caption[Per-op microbench, 3.4.1 sweep.]{Per-op time before
and after the 3.4.1 algorithmic sweep. The biggest wins (RoPE
cache, DropPath vectorisation) are on ops with high call
counts but relatively cheap math, where the constant overhead
per call dominated.}
\label{tab:no_compile_sweep:per_op_sample}
\end{table}

\section{Summary and Forward References}
\label{sec:no_compile_sweep:summary}

The 3.4.1 algorithmic sweep landed six algorithmic improvements
(Winograd, streaming softmax, BNNS direct path, GEMM tiling,
RoPE cache, DropPath vec) that closed the remaining eager-mode
gap to within 10\% of the reference framework on most of the
zoo. The candidates that did not land were either marginal
(Strassen) or premature (INT8 path waiting on hardware).

This sets the no-compile baseline. The next chapter,
\chref{ch:graph_capture_wins}, measures what
\code{lucid.compile} (\chref{ch:compile_design}) adds on top.
The combined eager+compile path delivers another 1.3--1.6$\times$
on top of these numbers.

\begin{lucidnote}
For the user: all the 3.4.1 improvements are transparent
(automatic on the appropriate ops). There is nothing to enable.
If you observe a regression vs 3.4.0, please file with the
bench-harness output --- the most common cause is hitting an
edge case in the Winograd path that the unit tests missed.
\end{lucidnote}
