% !TEX root = ../main.tex
\chapter{Graph-Capture Wins (3.5.0)}
\label{ch:graph_capture_wins}

The 3.5.0 release shipped \code{lucid.compile}, the
graph-capture and compilation infrastructure described in
detail in \partref{part:graph_compile}. This chapter is the
performance retrospective: how much additional speedup compile
delivers on top of the eager-mode baseline, how the speedup
distributes across model families, how the compile cost (one-
time) amortises, and where the compile path still has
headroom.

\section{Headline Numbers}
\label{sec:graph_capture_wins:headline}

The 3.5.0 release ships the following compile speedups vs
3.4.1 eager (which is itself within 10\% of the reference
framework):

\begin{table}[t]
\centering
\small
\begin{tabular}{lrrr}
\toprule
Workload & 3.4.1 eager (ms) & 3.5.0 compile (ms) & Speedup \\
\midrule
ResNet-50 train       & 64  & 49  & 1.31$\times$ \\
ResNet-101 train      & 122 & 88  & 1.39$\times$ \\
ConvNeXt-Tiny train   & 90  & 72  & 1.25$\times$ \\
ConvNeXt-Base train   & 195 & 145 & 1.34$\times$ \\
ViT-Base train        & 92  & 60  & 1.53$\times$ \\
ViT-Large train       & 280 & 180 & 1.56$\times$ \\
Swin-Tiny train       & 85  & 56  & 1.52$\times$ \\
GPT-2 small train     & 105 & 65  & 1.62$\times$ \\
GPT-2 medium train    & 280 & 170 & 1.65$\times$ \\
BERT-base train       & 88  & 56  & 1.57$\times$ \\
BERT-large train      & 285 & 175 & 1.63$\times$ \\
T5-base train         & 190 & 115 & 1.65$\times$ \\
DDPM-UNet train       & 95  & 70  & 1.36$\times$ \\
\bottomrule
\end{tabular}
\caption[3.5.0 compile speedups.]{Iteration time at 3.4.1
eager (the prior baseline) vs 3.5.0 compiled. Transformer
models gain the most (1.5--1.65$\times$); CNNs gain 1.25--1.4$\times$.
Detection models (Mask R-CNN, DETR) are not in this table
because their dynamic-shape requirement prevents compile in 3.5;
they will benefit from 3.5.1's dynamic-shape support.}
\label{tab:graph_capture_wins:headline}
\end{table}

A consistent 1.3--1.6$\times$ on top of an already-tuned eager
baseline. Combined with the 3.0 $\to$ 3.4.1 stabilisation
work, the total improvement from 3.0.0 to 3.5.0 on GPT-2 small
is $380 \to 65 = 5.8\times$ faster.

\section{Where the Speedup Comes From}
\label{sec:graph_capture_wins:where_speedup}

A phase-level breakdown for GPT-2 small forward, comparing 3.4.1
eager to 3.5.0 compiled:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Phase & 3.4.1 eager (ms) & 3.5.0 compile (ms) \\
\midrule
Python dispatch overhead   & 16  & 0  \\
C++ op dispatch + reg lookup & 4   & 0  \\
MPSGraph construction      & 0.5 & 0  \\
MPSGraph execution         & 70  & 60 \\
Memory allocator           & 7   & 2  \\
Cast ops (autocast)        & 2   & 0  \\
Compile sync / launch      & 0   & 3  \\
Other                      & 5.5 & 0  \\
\midrule
Total                      & 105 & 65 \\
\bottomrule
\end{tabular}
\caption[Phase-level: eager vs compile.]{All the per-op-call
overhead (Python dispatch, C++ dispatch, MPSGraph construction,
cast ops, allocator) collapses to zero under compile because
the entire graph is one MPSGraph execution. The MPSGraph
execution itself is also slightly faster (60 vs 70\,ms) due to
intra-graph fusion that the per-op dispatcher cannot perform.}
\label{tab:graph_capture_wins:phase_diff}
\end{table}

The speedup is mostly \emph{not} from faster math --- it's
from eliminating the per-call overhead. This matches the
chapter-58 lesson: at the kernel-compute level, the eager
path was already near-optimal; the wins came from cutting
the kernel-call count.

\section{Variance Across Model Families}
\label{sec:graph_capture_wins:variance}

The 1.3--1.6$\times$ range is informative:

\begin{itemize}
  \item \textbf{Transformers (1.5--1.65$\times$)}: many small
    ops per layer (LayerNorm, residual add, GELU activation,
    multiple matmuls), so per-op overhead was a large fraction
    of eager time.
  \item \textbf{CNNs (1.25--1.4$\times$)}: fewer small ops
    (conv dominates by compute), so per-op overhead was a
    smaller fraction.
  \item \textbf{UNets (1.35$\times$)}: balanced --- some
    transformer-like attention blocks plus convolution
    backbone.
  \item \textbf{Detection (compile not enabled)}: dynamic
    shapes; will benefit from 3.5.1.
\end{itemize}

The pattern matches the prediction: compile helps most where
there are many per-op calls.

\section{Compile Time}
\label{sec:graph_capture_wins:compile_time}

The one-time cost of compiling a model:

\begin{table}[t]
\centering
\small
\begin{tabular}{lrr}
\toprule
Model & Compile time (s) & Iters to break even \\
\midrule
ResNet-50         & 0.7  & 47  \\
ResNet-101        & 1.2  & 35  \\
ViT-Base          & 0.9  & 28  \\
ViT-Large         & 2.4  & 24  \\
GPT-2 small       & 1.8  & 45  \\
GPT-2 medium      & 4.2  & 38  \\
BERT-base         & 1.5  & 47  \\
T5-base           & 2.7  & 36  \\
\bottomrule
\end{tabular}
\caption[Compile time and break-even.]{Compile time for the
3.5.0 path, plus the number of training iterations after which
the compile cost is amortised (compile time / per-iter
speedup). For all models, break-even is within the first 50
iterations --- a fraction of a typical training run.}
\label{tab:graph_capture_wins:compile_time}
\end{table}

For any training run of more than 100 iterations, compile is
purely beneficial. For one-shot inference (a single forward
pass), compile is a net loss; for batched inference (the same
model called repeatedly), it pays off after the second or third
call.

\section{Memory Effects}
\label{sec:graph_capture_wins:memory}

Compile also affects peak activation memory, in two competing
directions:

\subsection{Saved by fusion}
\label{ssec:graph_capture_wins:memory_saved}

Fused ops do not materialise intermediates. Examples:
\begin{itemize}
  \item Conv-BN-ReLU as a single fused kernel: no intermediate
    BN-output and ReLU-input tensors.
  \item LayerNorm-Linear: no post-LN tensor.
  \item SDPA: no $T \times T$ attention matrix.
\end{itemize}

Cumulatively, this saves 15--30\% peak activation memory.

\subsection{Cost: graph executable}
\label{ssec:graph_capture_wins:memory_cost}

The compiled MPSGraph executable itself takes memory --- a few
MB for ResNet-50, up to 40\,MB for GPT-2 medium. Negligible
in the context of a multi-GB training workload.

\subsection{Net effect}
\label{ssec:graph_capture_wins:memory_net}

For GPT-2 small training, batch 32:

\begin{itemize}
  \item Eager: 5.4\,GB peak.
  \item Compile: 4.1\,GB peak ($-24\%$).
\end{itemize}

The user can train at $\sim$30\% larger batches under compile
than under eager.

\section{Latency Variance}
\label{sec:graph_capture_wins:variance_latency}

Compile also reduces iteration-time variance. Eager iterations
include Python-side variance (GC pauses, allocator misses,
dispatch overhead jitter); compile reduces these to zero.

For GPT-2 small training over 100 iterations:

\begin{itemize}
  \item Eager: median 105\,ms, p05 98, p95 122 (variance
    $\pm$\,8\%).
  \item Compile: median 65\,ms, p05 64, p95 68 (variance
    $\pm$\,3\%).
\end{itemize}

For latency-sensitive serving, the tighter distribution is as
important as the lower median.

\section{Workloads That Do Not Benefit}
\label{sec:graph_capture_wins:no_benefit}

Some workloads gain little or nothing from compile:

\begin{itemize}
  \item \textbf{Dynamic-shape models}: compile recompiles per
    shape, which can outweigh the per-iter savings. Detection
    models (variable proposal counts) are in this category.
  \item \textbf{Tiny models}: per-iteration time below 10\,ms,
    so the per-op overhead is already a small absolute amount.
    The relative speedup may be 1.5$\times$ but the absolute is
    a few hundred microseconds --- often not worth the compile
    cost.
  \item \textbf{Custom Functions in the hot path}: the
    Function triggers eager fallback; the surrounding compile
    region must round-trip data, partially negating the win.
  \item \textbf{First-iteration inference}: the compile cost is
    not amortised.
\end{itemize}

The 3.5.0 release documents these explicitly to avoid user
disappointment.

\section{Compile vs Eager: Operational Differences}
\label{sec:graph_capture_wins:op_diff}

Beyond pure throughput, the user experience differs:

\begin{itemize}
  \item \textbf{Stack traces under compile}: collapsed (the
    entire graph executes as one MPSGraph call). For
    debugging, the user must drop \code{compile} and re-run
    eager.
  \item \textbf{Anomaly detection}: \code{check\_nan=True}
    flag at compile time enables boundary checks; per-op
    anomaly detection requires eager.
  \item \textbf{Profile attribution}: the compile path attributes
    all time to ``compiled\_graph''; per-op attribution
    requires eager mode for the profiler.
\end{itemize}

The recommendation: develop in eager (for visibility) and
deploy in compile (for speed). The
\code{lucid.compile(model)} call is the only line difference.

\section{The Long Tail of Compile Wins}
\label{sec:graph_capture_wins:long_tail}

Beyond the headline-table workloads, some unexpected wins:

\subsection{Optimizer step}
\label{ssec:graph_capture_wins:optimizer_compile}

\Lucid's optimisers (Adam, AdamW \cite{loshchilov_adamw_2019},
SGD with momentum) execute many small per-parameter ops:
exponential moving-average updates, parameter writes, etc. For
a 100M-parameter model, this is 1000+ ops, each a microsecond.

The 3.5 \code{lucid.compile} now wraps optimiser steps too
(\code{compile(optimizer)} alongside \code{compile(model)}).
For GPT-2 small, optimiser-step time dropped from 8 to 3\,ms.

\subsection{Loss-fn fusion}
\label{ssec:graph_capture_wins:loss_compile}

The cross-entropy loss (used by classification, language
modelling) is a softmax + NLL + reduce. Compiled, it is one
fused MPSGraph call; eager is three. Saves $\sim$\,1\,ms per
iter.

\section{Comparison to Other Compile Stacks}
\label{sec:graph_capture_wins:comparison}

A note on how the \Lucid\ compile path compares to alternatives:

\begin{itemize}
  \item \textbf{Reference-framework's compile path}: similar
    speedup ratio ($\sim$1.4--1.6$\times$) on the same hardware.
    \Lucid\ has parity with reference compile.
  \item \textbf{MLX's lazy mode}: MLX itself uses lazy
    evaluation; some of the per-op cache benefit is already
    available without explicit compile. \Lucid\ adds the
    backward-graph compile, which MLX does not.
  \item \textbf{Apple's CoreML compile}: deployment-focused,
    not training. Comparable speedup but inflexible for
    research workflows.
\end{itemize}

\section{What Compile Does NOT Speed Up}
\label{sec:graph_capture_wins:no_speedup}

To set expectations correctly:

\begin{itemize}
  \item Data loading (CPU side, separate process pool).
  \item Disk I/O for checkpoint save/load.
  \item Python-side training-loop overhead (logging,
    metrics).
  \item Network communication (distributed training, not
    yet supported).
\end{itemize}

If the training loop is bottlenecked on data loading, compile
saves nothing. The user must address each bottleneck separately;
\Lucid\ provides tools (\code{lucid.profiler.system}) to
identify which.

\section{Future Compile Wins (3.5.1+)}
\label{sec:graph_capture_wins:future}

The compile path has several gaps slated for closing:

\begin{itemize}
  \item \textbf{Dynamic shapes (3.5.1)}: unblocks detection
    models.
  \item \textbf{Cross-iteration fusion (3.6)}: fuse the
    optimiser step with the backward pass.
  \item \textbf{Memory planning (3.6)}: pre-allocate the entire
    compile-graph's memory footprint; eliminate per-step
    allocator calls.
  \item \textbf{Whole-program compile (3.7)}: include the
    data-augmentation pipeline in the compile region (for
    GPU-side augmentation).
\end{itemize}

Each item targets a different remaining bottleneck.

\section{Summary and Forward References}
\label{sec:graph_capture_wins:summary}

The 3.5.0 compile path delivers a consistent 1.3--1.6$\times$
speedup on top of the 3.4.1 eager baseline, predominantly by
eliminating per-op dispatch overhead. The wins are largest on
transformer workloads (1.5--1.65$\times$), smallest on detection
(no compile yet) and tiny models (eager is already cheap).
Compile costs amortise within $<$\,50 iterations for all
zoo models.

This chapter closes \partref{part:performance}. The next part
treats production-engineering topics: testing
(\chref{ch:testing}), serialization (\chref{ch:serialization}),
data pipeline (\chref{ch:data_pipeline}), error handling
(\chref{ch:error_handling}), and reproducibility
(\chref{ch:reproducibility}). These chapters cover the
non-throughput aspects of production-grade work that have been
in the background of every previous chapter.

\begin{lucidnote}
For the user: add \code{lucid.compile(model)} to your
training script. For inference, additionally
\code{lucid.compile(optimizer)} if you keep optimiser state.
The speedup is automatic on the next iteration. If you observe
a slowdown rather than a speedup, profile to confirm; sometimes
small models or dynamic-shape workloads are not yet served by
the compile path.
\end{lucidnote}
