% !TEX root = ../main.tex
\chapter{Testing Strategy}
\label{ch:testing}

A framework's tests are not a courtesy; they are part of the
framework. \Lucid\ ships with $\sim$\,3,500 unit, integration,
property, and parity tests covering every public API, every
op, every layer, every model variant, and every backward
formula. This chapter explains the layered testing strategy,
the parity-test infrastructure that compares \Lucid\ against
a reference framework on every PR, the property-based testing
that catches edge cases the developer did not anticipate,
and the regression-tracking that keeps performance on track.
The chapter also discusses the test infrastructure as a design
choice: most of it is opt-in (the reference framework is a test
dependency, not a runtime dependency).

\section{The Test Layers}
\label{sec:testing:layers}

\Lucid's tests are organised into seven layers, each with a
distinct purpose and execution strategy.

\begin{enumerate}
  \item \textbf{Unit tests} (\code{lucid/test/unit/}): single-
    function, single-class tests. Pure-Python, fast.
  \item \textbf{Integration tests}
    (\code{lucid/test/integration/}): multi-component flows
    (e.g., model forward + backward + optimiser step).
  \item \textbf{Parity tests} (\code{lucid/test/parity/}):
    compare \Lucid\ outputs to a reference framework on the
    same input. Opt-in; require \code{ref\_framework} install.
  \item \textbf{Property tests} (\code{lucid/test/property/}):
    Hypothesis-driven \cite{hypothesis_maciver} tests over
    random shapes/dtypes/strides.
  \item \textbf{Model contract tests}
    (\code{lucid/test/unit/models/test\_family\_contract.py}):
    verify the H10/H11/H12 model-zoo conventions.
  \item \textbf{Benchmark tests} (\code{lucid/bench/}):
    measure throughput, not correctness. Gated by performance
    regression rules.
  \item \textbf{End-to-end training tests}
    (\code{lucid/test/e2e/}): one-epoch training runs that
    verify convergence (loss decreases).
\end{enumerate}

\section{The pytest Configuration}
\label{sec:testing:pytest_config}

\Lucid\ uses pytest \cite{pytest_krekel}. The \code{pyproject.toml}
configuration:

\begin{lstlisting}[style=lucid,language=LucidPython]
[tool.pytest.ini_options]
testpaths = ["lucid/test"]
markers = [
    "slow: marks tests as slow",
    "parity: requires reference framework",
    "needs_metal: requires M-series GPU",
    "needs_e2e: end-to-end training (>1 min)",
]
addopts = "-q --strict-markers"
\end{lstlisting}

The marker system separates the fast/CI-friendly tests from the
slow/optional tests. The standard CI run is:

\begin{lstlisting}[style=lucid,language=LucidPython]
pytest lucid/test/ -m "not slow and not parity and not needs_e2e"
\end{lstlisting}

This completes in $\sim$\,90\,s on M3~Max --- fast enough to
run on every commit.

\section{Parity Tests: The Reference Framework}
\label{sec:testing:parity_framework}

The parity tests are \Lucid's correctness gold standard. Rather
than write parameterised expected values by hand (error-prone,
unmaintainable), the parity tests compute the same operation
in a reference ML framework and compare.

The reference framework is loaded through a small adapter at
\code{lucid/test/\_fixtures/ref\_framework.py}. The adapter is
the \emph{only} test-tree file that may import the reference
framework by name (per Rule H5); other tests receive a
\code{ref} fixture.

\subsection{The ref fixture}
\label{ssec:testing:ref_fixture}

\begin{lstlisting}[style=lucid,language=LucidPython]
def test_softmax_parity(ref):
    x = lucid.randn(8, 16)
    ref_x = ref.tensor(x.numpy())  # Convert via bridge

    lucid_out = lucid.nn.functional.softmax(x, dim=-1)
    ref_out   = ref.nn.functional.softmax(ref_x, dim=-1)

    assert lucid.allclose(lucid_out, ref.numpy(ref_out), atol=1e-5)
\end{lstlisting}

The fixture handles:
\begin{itemize}
  \item Skip if the reference framework is not installed (the
    test is marked \code{@pytest.mark.parity}).
  \item Provide a uniform conversion API
    (\code{ref.tensor}, \code{ref.numpy}).
  \item Provide module proxies (\code{ref.nn},
    \code{ref.nn.functional}, etc.) so test code can write
    framework-agnostic calls.
\end{itemize}

\subsection{Parity scope}
\label{ssec:testing:parity_scope}

Parity tests cover:
\begin{itemize}
  \item Every op in \code{lucid.\_C.ops.*}.
  \item Every layer in \code{lucid.nn.*}.
  \item Every optimiser in \code{lucid.optim.*}.
  \item Every model-zoo factory (forward + backward + one
    optimiser step).
  \item Edge cases: 0-dim tensors, broadcasting, mixed dtypes.
\end{itemize}

At PR time, the parity tests run in the slow CI track (because
they require the reference framework to be installed); they
block merge if any fail.

\section{Property Tests}
\label{sec:testing:property_tests}

Property tests \cite{hypothesis_maciver} generate random inputs
and verify properties that should hold for all valid inputs.
The \Lucid\ property tests are particularly useful for
catching corner cases that the developer would not write
explicitly.

\subsection{Example}
\label{ssec:testing:property_example}

\begin{lstlisting}[style=lucid,language=LucidPython]
from hypothesis import given, strategies as st

@given(
    shape=st.lists(st.integers(1, 16), min_size=1, max_size=4),
    dtype=st.sampled_from([lucid.float32, lucid.float16]),
)
def test_add_is_commutative(shape, dtype):
    a = lucid.randn(*shape, dtype=dtype)
    b = lucid.randn(*shape, dtype=dtype)
    assert lucid.allclose(a + b, b + a)
\end{lstlisting}

Hypothesis tries many random shapes (and shrinks on failure to
the simplest reproducing case). For \Lucid, property tests have
historically found:

\begin{itemize}
  \item Broadcasting bugs when one operand was 0-dim.
  \item Stride bugs when a tensor was non-contiguous.
  \item Dtype-promotion bugs in mixed-dtype binary ops.
  \item Index-overflow bugs in \code{gather} with negative
    indices.
\end{itemize}

\subsection{Strategies catalogue}
\label{ssec:testing:strategies_catalogue}

\Lucid\ has a shared strategy catalogue at
\code{lucid/test/\_strategies.py}:
\begin{lstlisting}[style=lucid,language=LucidPython]
random_tensor()       # Any shape, any dtype, any contiguity
random_2d_matrix()    # Shape (N, M) with N, M in [1, 128]
random_4d_image()     # NCHW with valid image dimensions
random_attention()    # Q, K, V with matching dimensions
\end{lstlisting}

Test authors use these strategies rather than rolling their
own; the result is broader and more uniform coverage.

\section{Model Contract Tests}
\label{sec:testing:model_contract}

The model-zoo conventions (H10/H11/H12) are checked by a
contract test that runs on every PR:

\begin{lstlisting}[style=lucid,language=LucidPython]
# lucid/test/unit/models/test_family_contract.py

def test_full_name_size_suffix(model_id):
    """H10: size suffix must be full word, not abbreviation."""
    parts = model_id.split("_")
    for part in parts:
        assert part not in {"t", "s", "b", "m", "l", "xl"}, \
            f"Abbreviated suffix in {model_id}"

def test_paper_cited_variant(model_id):
    """H11: variant must exist in the cited paper."""
    info = registry[model_id]
    assert info.citation is not None
    assert info.params == get_paper_param_count(info.citation, model_id)
\end{lstlisting}

The contract tests run on the full set of registered models.
H10 violations (e.g., a new \code{vit\_b} factory) fail
immediately; H11 violations (an invented variant) require the
author to either cite the paper or delete the factory.

\section{Gradient Check}
\label{sec:testing:gradcheck}

For correctness of backward formulas, \Lucid\ uses numerical
gradient checking (\chref{ch:gradcheck}):

\begin{lstlisting}[style=lucid,language=LucidPython]
from lucid.autograd import gradcheck

def test_softmax_grad():
    x = lucid.randn(4, 8, dtype=lucid.float64, requires_grad=True)
    assert gradcheck(
        lambda x: lucid.nn.functional.softmax(x, dim=-1),
        (x,),
        eps=1e-6, atol=1e-4,
    )
\end{lstlisting}

gradcheck perturbs each input element by $\pm\epsilon$ and
compares the analytical gradient to the finite-difference
gradient. The test runs in FP64 (more precision needed for
finite-difference comparison than for actual training).

Every \code{Function} in \code{lucid.autograd} has a
corresponding gradcheck test. New Functions without one fail
the model-zoo contract test.

\section{Benchmark Regression Tests}
\label{sec:testing:bench_regression}

Performance is also tested, not just correctness. The
\code{lucid/bench/regression/} directory holds a small fast
benchmark suite (under 5\,s total) that runs on every PR:

\begin{lstlisting}[style=lucid,language=LucidPython]
@pytest.mark.benchmark
def test_resnet50_iter_time():
    model = lucid.models.resnet_50()
    x = lucid.randn(32, 3, 224, 224)
    result = bench(lambda: model(x).sum().backward())
    assert result.median_ms < REGRESSION_THRESHOLD["resnet_50"]
\end{lstlisting}

The threshold is set at $1.15 \times$ the current baseline
(15\% headroom for noise). A regression that crosses the
threshold blocks the PR. The author either justifies the
regression (and updates the threshold) or fixes it.

\section{End-to-End Training Tests}
\label{sec:testing:e2e_training}

The \code{lucid/test/e2e/} directory holds short training runs
that verify the loss decreases:

\begin{lstlisting}[style=lucid,language=LucidPython]
@pytest.mark.needs_e2e
def test_resnet50_one_epoch():
    model = lucid.models.resnet_50()
    opt = lucid.optim.SGD(model.parameters(), lr=0.1)
    losses = []
    for x, y in cifar10_subset(num_batches=50):
        out = model(x)
        loss = lucid.nn.functional.cross_entropy(out, y)
        loss.backward()
        opt.step()
        opt.zero_grad()
        losses.append(loss.item())
    assert losses[-1] < losses[0] * 0.9  # 10% decrease minimum
\end{lstlisting}

These tests are slow (1--5\,min each) and run only on the
release branch, not per-commit.

\section{Coverage}
\label{sec:testing:coverage}

\Lucid\ tracks line + branch coverage via \code{coverage.py}.
The current state:

\begin{table}[t]
\centering
\small
\begin{tabular}{lr}
\toprule
Module & Coverage \\
\midrule
\code{lucid/\_tensor/}    & 96\% \\
\code{lucid/nn/}          & 94\% \\
\code{lucid/optim/}       & 98\% \\
\code{lucid/autograd/}    & 92\% \\
\code{lucid/linalg/}      & 95\% \\
\code{lucid/fft/}         & 93\% \\
\code{lucid/models/}      & 88\% \\
\code{lucid/compile/}     & 86\% \\
\midrule
Overall                   & 93\% \\
\bottomrule
\end{tabular}
\caption[Test coverage by module.]{Line+branch coverage,
\Lucid\ 3.5.0. Models and compile are lower because they
include some path-rich code (per-variant initialisation, fallback
paths) that is harder to fully exercise. The 93\% overall is
above the project's 85\% target.}
\label{tab:testing:coverage}
\end{table}

Untested lines are mostly error-path branches; we accept
coverage gaps in clearly defensive code (e.g., type-check
errors that ``can't happen'' under the public-API contracts).

\section{The CI Pipeline}
\label{sec:testing:ci_pipeline}

The CI pipeline runs on GitHub Actions, on three test tracks:

\subsection{Fast track (per commit)}
\label{ssec:testing:fast_track}

\begin{itemize}
  \item Compile C++ engine.
  \item Run unit + integration + property tests
    (\code{not slow and not parity and not needs\_e2e}).
  \item Run mypy + ruff + the model-zoo contract.
  \item Run benchmark-regression tests.
\end{itemize}

Wall-clock: $\sim$\,3.5 minutes. Blocking for merge.

\subsection{Slow track (per PR)}
\label{ssec:testing:slow_track}

\begin{itemize}
  \item All fast-track tests.
  \item Parity tests (with reference framework installed).
  \item Slow integration tests.
\end{itemize}

Wall-clock: $\sim$\,18 minutes. Blocking for merge.

\subsection{Nightly track}
\label{ssec:testing:nightly_track}

\begin{itemize}
  \item All slow-track tests.
  \item End-to-end training tests.
  \item Full model-zoo benchmark sweep.
  \item Hypothesis with extended example count.
\end{itemize}

Wall-clock: $\sim$\,3 hours. Failures alert via Slack but do
not block merges (the failure is investigated next morning).

\section{Test Data Management}
\label{sec:testing:test_data}

The tests need data. \Lucid's policy:

\begin{itemize}
  \item \textbf{Tiny data inline}: hard-coded tensors for
    correctness tests (a $4 \times 4$ matrix with known
    determinant, etc.).
  \item \textbf{Generated data}: random tensors via
    \code{lucid.randn} with fixed seeds.
  \item \textbf{No external downloads in unit tests}: the
    fast track must run offline. Larger data (CIFAR-10 subset
    for e2e) is downloaded on the nightly track.
\end{itemize}

The seed-fixing convention: every test that uses randomness
calls \code{lucid.seed(42)} at the top. This makes failures
reproducible.

\section{When a Test Fails}
\label{sec:testing:test_failure}

The expected debugging flow:

\begin{enumerate}
  \item Identify the test: \code{pytest} output names the file
    and test.
  \item Run in isolation:
    \code{pytest lucid/test/<file>.py::<test\_name> -v}.
  \item Read the failure: \Lucid's assertion library produces
    informative diffs (shape, dtype, max-abs-diff for
    tensor mismatches).
  \item For parity failures: bisect across \Lucid\ commits to
    find the regression-introducing commit (\code{git bisect}
    + the failing test command).
  \item For correctness failures: drop into the failing test
    with \code{pytest --pdb}.
\end{enumerate}

The \code{git bisect} flow is so common that we have a small
helper script: \code{tools/test-bisect.sh}.

\section{What We Do Not Test}
\label{sec:testing:not_tested}

Some intentional gaps:

\begin{itemize}
  \item \textbf{Performance under thermal stress}: the bench
    harness assumes steady-state; we don't test the
    throttled-state behaviour (impractical in CI).
  \item \textbf{Memory pressure}: triggering OOM
    deterministically across CI machines is hard; we test the
    error-handling path with synthetic injection instead.
  \item \textbf{Multi-day training runs}: only single-epoch
    e2e tests; full multi-day convergence to specific accuracy
    is left to the user.
\end{itemize}

These gaps are documented; user reports often fill them.

\section{Summary and Forward References}
\label{sec:testing:summary}

The testing strategy: layered (unit, integration, parity,
property, contract, benchmark, e2e), automated (CI on three
tracks), and gold-standard-verified (parity tests against a
reference framework). The 93\% line+branch coverage and 3,500
test count are the framework's correctness foundation.

The next chapter, \chref{ch:serialization}, treats checkpoint
serialization. \chref{ch:data_pipeline} treats the data
pipeline. \chref{ch:error_handling} treats error messaging.
\chref{ch:reproducibility} treats reproducibility.

\begin{lucidnote}
For the user: when reporting a bug, the most actionable
artifact is a failing test. Write the smallest reproducing
test you can, attach it to the issue, and \Lucid\ maintainers
can add it to the regression suite once fixed.
\end{lucidnote}
