% !TEX root = ../main.tex
\chapter{Error Handling and Anomaly Detection}
\label{ch:error_handling}

A framework's error messages are part of its user interface,
and most of the time the only part that the user notices.
\Lucid\ treats errors as a first-class design surface: every
error message names the user-side cause when possible, suggests
a concrete fix, and never punts the user to a stack trace inside
\Lucid\ internals. This chapter documents the error
infrastructure: the exception hierarchy, the message-quality
standard, the anomaly-detection mode for catching NaN
gradients, the shape-mismatch diagnostic, and the per-op
debug context that surfaces when something goes wrong.

\section{The Exception Hierarchy}
\label{sec:error_handling:hierarchy}

All \Lucid-thrown exceptions descend from \code{LucidError}.
The major subclasses:

\begin{itemize}
  \item \code{LucidError(Exception)}: base.
  \item \code{ShapeError(LucidError)}: tensor shape
    mismatch.
  \item \code{DtypeError(LucidError)}: dtype mismatch or
    unsupported dtype.
  \item \code{DeviceError(LucidError)}: tensor on wrong
    device (e.g., GPU op called on CPU tensor).
  \item \code{NotImplementedError(LucidError)}: op not
    implemented in the requested configuration.
  \item \code{NaNError(LucidError)}: NaN/Inf detected (anomaly
    mode).
  \item \code{HubError(LucidError)}: pretrained-weights hub
    failure (network, integrity, etc.).
  \item \code{CompileError(LucidError)}: lucid.compile failure.
\end{itemize}

The hierarchy is shallow on purpose: catching is meant to be
``\code{except ShapeError}'' or
``\code{except LucidError}'', not a deep tree.

\section{Message-Quality Standard}
\label{sec:error_handling:message_standard}

Every error message must satisfy:

\begin{enumerate}
  \item \textbf{Name the API surface}: the function the user
    called, with file:line of their code.
  \item \textbf{State what was expected vs received}: shapes,
    dtypes, devices, valid values.
  \item \textbf{Suggest the most likely fix}: not just a
    description of the problem, but a hint at what to do.
\end{enumerate}

An example of a good message:

\begin{lstlisting}[style=lucid,language=LucidPython]
ShapeError: matmul expected input shapes compatible for
  broadcasting, got:
    a: (16, 32, 64)
    b: (32, 128)
  The last dim of `a` (64) must equal the second-to-last dim of `b` (32).
  Likely fix: did you intend `b` to be (64, 128)?
  At lucid/_C/ops/matmul.cpp:142
  Called from user code at /Users/.../train.py:87
\end{lstlisting}

The user knows what went wrong, why, and what to try.
A bad message:

\begin{lstlisting}[style=lucid,language=LucidPython]
RuntimeError: dim mismatch
\end{lstlisting}

\Lucid's CI does not gate messages on this standard
automatically, but reviewers reject PRs with low-quality
error strings; the message is part of the public API.

\section{Shape-Mismatch Diagnostic}
\label{sec:error_handling:shape_diagnostic}

Shape errors are the most common; the diagnostic is correspondingly
detailed. The \code{ShapeError} machinery in
\code{lucid/\_C/error.cpp} formats:

\begin{itemize}
  \item Both shapes with axis labels (if available).
  \item A red highlight on the mismatched axis.
  \item The broadcasting rule that was violated.
  \item A suggested fix if the shapes are off by one obvious
    transposition or contiguity issue.
\end{itemize}

\begin{lstlisting}[style=lucid,language=LucidPython]
ShapeError: add expected broadcast-compatible shapes, got:
    a:    (4,  8, 16)        # 3 dims
    b:    (   16,  8)        # 2 dims (right-aligned)
                ^^
  Mismatched axis: dim -2 of a (8) != dim -2 of b (16)
  Likely fix: did you mean to transpose `b`?
\end{lstlisting}

The diagnostic uses C++ source location info combined with the
Python user-code site (via the binding's interpreter frame).

\section{Anomaly Detection}
\label{sec:error_handling:anomaly}

The hardest debugging case in deep learning: a NaN appears mid-
training and you don't know which op produced it. \Lucid's
anomaly-detection mode addresses this:

\begin{lstlisting}[style=lucid,language=LucidPython]
lucid.autograd.set_detect_anomaly(True)
\end{lstlisting}

When enabled, every op records its call stack into the
\code{GradFn}; if any output is NaN/Inf, the framework
identifies the offending op and raises:

\begin{lstlisting}[style=lucid,language=LucidPython]
NaNError: NaN detected in output of `lucid.log`.
  Input was a tensor with values in [-0.5, 3.0]; the negative
  value caused NaN in the log.
  Stack:
    File "/Users/.../train.py", line 142, in compute_loss
      log_p = lucid.log(p)
    File "/Users/.../model.py", line 81, in forward
      p = sigmoid(self.linear(x))
\end{lstlisting}

The cost of anomaly mode is $\sim$20\% slower training; it is
opt-in for debugging, not always on.

\subsection{Forward-only anomaly}
\label{ssec:error_handling:fwd_anomaly}

For a forward-pass NaN, \code{set\_detect\_anomaly(True)} works
the same way but checks each op's output, not just gradient.

\subsection{Backward-only anomaly}
\label{ssec:error_handling:bwd_anomaly}

For a gradient NaN with finite forward, the anomaly mode
records each \code{GradFn}'s execution and reports the offending
backward op. The first NaN-producing backward is identified.

\subsection{Under compile}
\label{ssec:error_handling:anomaly_compile}

The compile path collapses per-op stacks; anomaly detection is
less informative under compile. The 3.5 release added the
\code{check\_nan=True} flag at compile time
(\chref{ch:compile_hardening}~Section~\ref{sec:compile_hardening:nan_handling})
that checks outputs at the compile boundary; for per-op
diagnosis, drop compile and re-run.

\section{Memory-Pressure Errors}
\label{sec:error_handling:memory_errors}

When allocation fails (typically due to OOM):

\begin{lstlisting}[style=lucid,language=LucidPython]
DeviceError: GPU memory allocation failed.
  Requested: 2.3 GB (tensor shape (32, 768, 768), fp32)
  Available: 1.1 GB free of 64 GB unified
  Allocator cache holds 18 GB.

  Suggestions:
  - Call lucid.runtime.empty_cache() to release the allocator cache.
  - Reduce batch size from 32 to 16.
  - Enable gradient checkpointing for memory-vs-compute tradeoff.
  - Use bfloat16 autocast to halve activation memory.
\end{lstlisting}

The error includes the precise free vs requested numbers, the
allocator's cache size (since releasing it may resolve the
issue), and a list of orthogonal fixes.

\section{Device-Mismatch Errors}
\label{sec:error_handling:device_mismatch}

When an op receives tensors on different devices (e.g., one on
CPU stream, one on GPU stream):

\begin{lstlisting}[style=lucid,language=LucidPython]
DeviceError: matmul received tensors on different devices.
  a.device: cpu
  b.device: metal:0
  All op inputs must be on the same device.
  Likely fix: `b = b.cpu()` or `a = a.metal()`.
\end{lstlisting}

The fix suggestion is mechanical (move one tensor to match
the other) but useful for the user who lost track of where
tensors are.

\section{Unsupported-Op Errors}
\label{sec:error_handling:unsupported_op}

When an op is called on an unsupported dtype/shape/configuration:

\begin{lstlisting}[style=lucid,language=LucidPython]
NotImplementedError: cholesky requires square positive-definite
  matrix, got shape (3, 4).
  cholesky is only defined for square matrices.

NotImplementedError: scaled_dot_product_attention(...,
  is_causal=True) requires macOS 26+. Detected macOS 25.4.
  Falls back to the composite path automatically; please
  upgrade for the fast path.
\end{lstlisting}

\section{Hub Errors}
\label{sec:error_handling:hub_errors}

Hub-related failures get their own subclass with detailed
context:

\begin{lstlisting}[style=lucid,language=LucidPython]
HubDownloadError: Failed to download `resnet_50` (imagenet1k-v2).
  URL: https://hub.lucid-ml.org/vision/cnn/resnet_50/imagenet1k-v2/weights.safetensors
  Reason: Connection timeout after 30s.
  Retry: lucid.hub.download("resnet_50") will retry with 3 attempts.
  Offline alternative: download via browser to ~/.lucid/hub/...

HubIntegrityError: Downloaded file failed SHA-256 check.
  Expected: a3b4c5d6...
  Got:      b8e7c2f1...
  The file is corrupted or has been tampered with.
  The local copy has been deleted. Retry the download.
\end{lstlisting}

The user knows the cause (network vs corruption), what
\Lucid\ will do, and what they can do.

\section{Compile Errors}
\label{sec:error_handling:compile_errors}

The compile path can fail in several ways:

\begin{lstlisting}[style=lucid,language=LucidPython]
CompileError: Tracer encountered unsupported op `fft.rfft`.
  Falling back to eager for the region containing this op.
  To disable fallback (and surface this error immediately),
  call lucid.compile(model, strict=True).

CompileError: MpsBuilder failed to emit `gather` for shape
  (1024, dynamic_dim, 64). Dynamic shape support is not yet
  available in 3.5; use a fixed shape or pad to the maximum.
\end{lstlisting}

The first variant (unsupported op + fallback) is a warning-
level event by default; the user opted into strict mode to
make it an error. The second is a fatal error in 3.5 (will be
fixed in 3.5.1).

\section{Logging Integration}
\label{sec:error_handling:logging}

\Lucid\ also produces non-fatal log messages via the standard
\code{logging} module under the \code{lucid} root logger:

\begin{lstlisting}[style=lucid,language=LucidPython]
import logging
logging.basicConfig(level=logging.INFO)
logger = logging.getLogger("lucid")

# Example messages:
# INFO  lucid.amp: autocast enabled with dtype=float16
# WARN  lucid.compile: fallback to eager for region (5 ops): fft.rfft
# DEBUG lucid.alloc: cache hit for shape (32, 3, 224, 224)
\end{lstlisting}

The default level is WARN; users can set DEBUG for verbose
output during debugging.

\section{The Debug Print of a Tensor}
\label{sec:error_handling:tensor_debug_print}

\code{print(tensor)} produces a human-readable summary:

\begin{lstlisting}[style=lucid,language=LucidPython]
print(x)
# Tensor(shape=(4, 8, 16), dtype=fp32, device=metal:0)
# [[[ 0.123  -0.456 ... 0.789  -0.012]
#   ...
#   [ ...                          ...]]]
\end{lstlisting}

For large tensors, only corner elements are shown (the standard
NumPy-style ellipsis abbreviation). The shape, dtype, and device
are always included.

For diagnostic purposes, \code{tensor.summary()} gives
statistics:

\begin{lstlisting}[style=lucid,language=LucidPython]
print(x.summary())
# Tensor summary:
#   shape: (4, 8, 16)
#   dtype: fp32
#   device: metal:0
#   min: -2.45, max: 3.12, mean: 0.04, std: 0.98
#   nan_count: 0, inf_count: 0, zero_count: 17
#   requires_grad: True (grad_fn: ReLUBackward)
\end{lstlisting}

\section{Custom Error Wrapping}
\label{sec:error_handling:custom_wrapping}

User-written code can wrap \Lucid\ errors with additional
context using \code{LucidError.with\_context}:

\begin{lstlisting}[style=lucid,language=LucidPython]
def train_epoch(model, loader):
    for batch_idx, (x, y) in enumerate(loader):
        try:
            out = model(x)
        except lucid.ShapeError as e:
            raise e.with_context(
                f"At batch {batch_idx}, input shape {x.shape}"
            )
\end{lstlisting}

The wrapped error preserves the original chain (Python
\code{raise from} semantics) so the user sees both the
high-level user-code site and the framework-internal cause.

\section{Test Coverage of Error Paths}
\label{sec:error_handling:error_test_coverage}

The \code{lucid/test/error/} directory contains tests that
exercise each error-path explicitly:

\begin{lstlisting}[style=lucid,language=LucidPython]
def test_matmul_shape_error_message():
    a = lucid.randn(16, 32, 64)
    b = lucid.randn(32, 128)
    with pytest.raises(lucid.ShapeError) as exc_info:
        a @ b
    assert "last dim of `a` (64)" in str(exc_info.value)
    assert "second-to-last dim of `b` (32)" in str(exc_info.value)
\end{lstlisting}

These tests verify both that the error is raised \emph{and}
that its message contains the expected diagnostic detail. The
message is part of the API contract.

\section{Error Telemetry}
\label{sec:error_handling:error_telemetry}

\Lucid\ has \emph{no telemetry} (per the project's
privacy stance: no anonymous error reporting, no usage
statistics, no remote phone-home). Users diagnose locally
or open a public issue.

\section{Summary and Forward References}
\label{sec:error_handling:summary}

\Lucid's error handling: shallow exception hierarchy, strict
message-quality standard, shape-mismatch diagnostic with
suggested fixes, anomaly detection for NaN/Inf catching,
device-mismatch / OOM / hub / compile errors all
context-rich, no telemetry. The error infrastructure is
treated as a public API: tests cover the messages, not just
the exception types.

The final chapter, \chref{ch:reproducibility}, treats
reproducibility --- closing the loop on the data pipeline +
serialization + error handling that this part of the report
covers.

\begin{lucidnote}
For the user: if you see a vague error or one without a
suggested fix, that is a bug; please report it. The error
infrastructure is iteratively improved, and a user-side
``confusing error'' report is one of the most actionable kinds
of issue.
\end{lucidnote}
